### 2024-02-09 products/manuscript/Manuscript.qmd
Project Part 1

---
title: "Example Manuscript Template for a Data Analysis Project"
subtitle: ""
author: Andreas Handel
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../assets/dataanalysis-references.bib
csl: ../assets/apa.csl
---

The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

# Summary/Abstract
_Write a summary of your project._

{{< pagebreak >}}

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

Finite mixture models are used to analyze data presumably arising from several separate, yet unobserved, groups. Among these models, Gaussian mixture models and mixture of t-distributions are particularly prevalent. However, such models which are based on symmetric distribution often do not accurately represent data with asymmetric distribution. Skewed data can be more effectively managed using asymmetric distributions, thereby extending their applicability across a broader spectrum of model-based clustering scenarios [1]. If the observed data is positive, distribution is right-skewed and assumed to come from a mixture distribution, employing mixture of gamma density is a rational choice [2].

This study intends to identify distinct customer groups within sales data of luxury goods, with each group comprising customers who exhibit similar characteristics in terms of purchase frequency and amounts. Modeling customer spending as a mixture of gamma distributions helps to identify different spending behaviors. The objective is to develop an Expectation Maximization (EM) algorithm which segregates groups based on their spending patterns, considering the individual purchases transactions made by each customer.

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._

This study employs simulated data to assess the effectiveness of the EM algorithm. Following successful testing, the refined algorithm will be applied to actual sales data pertaining to luxury goods.

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._

This study aims to answer the question, "How effective is the Expectation Maximization (EM) algorithm in identifying distinct customer groups within luxury goods sales data when accounting for individual spending by each customer, as demonstrated by initial validation using simulated data?"

[1] Lee, S. X., & McLachlan, G. J. (2013). Model-based clustering and classification with non-normal mixture distributions. Statistical Methods & Applications, 22(4), 427-454.

[2] Young, D. S., Chen, X., Hewage, D. C., & Nilo-Poyanco, R. (2019). Finite mixture-of-gamma distributions: estimation, inference, and model-based clustering. Advances in Data Analysis and Classification, 13, 1053-1082.

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
#knitr::include_graphics(here("products","assets","antigen-recognition.png"))
```

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

## Statistical analysis
_Explain anything related to your statistical analyses._

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","height-weight-stratified.png"))
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables-files","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

{{< pagebreak >}}

# References

### 2024-02-24 R/eda-code/eda_superstore.qmd
Project Part 2 test

---
title: "An example exploratory analysis script"
author: "Andreas Handel"
date: "10/28/2021"
output: html_document
---

# Setup
Required packages are loaded.
```{r}
#load needed packages. 
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
```

Data is loaded.

```{r}
#Path to data using here() package
data_location <- here::here("data","processed-data","processed_superstore.rds")
#load data
mydata <- readRDS(data_location)
```

# Data exploration through tables

Creating summary table and saving it to the results folder.

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results","tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

# Data exploration through figures

Histograms were plotted for the continuous outcomes. The data contains individual observations for sales of different stocks in a single invoice. As this project intends to study customer spending, Invoice totals are calculated for purchases by each customer at a time to visualize distribution of customer spending each time.The histogram is heavily right-skewed with higher frequencies at less than 500.

```{r}
# Histogram of Invoice values
## First calculate total invoice value per invoice
invoice_values <- mydata %>%
  group_by(InvoiceNo) %>%
  summarise(TotalInvoiceValue = sum(Sales, na.rm = TRUE))

##Create histogram
p1 <- ggplot(invoice_values, aes(x = TotalInvoiceValue)) +
  geom_histogram(binwidth = 50, fill = "lightblue", color = "darkblue") + # Adjust binwidth as needed
  labs(title = "Histogram of Total Invoice Values",
       x = "Total Invoice Value", 
       y = "Frequency") +
  theme_minimal()
plot(p1)

figure_file = here("results","figures","Sales_distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

Similarly, histogram plot was created for total Quantity of purchase per invoice which is also heavily right-skewed.

```{r}
## First calculate total invoice quantity per invoice
invoice_quantity <- mydata %>%
  group_by(InvoiceNo) %>%
  summarise(TotalInvoiceQuantity = sum(Quantity, na.rm = TRUE))

p2 <- ggplot(invoice_quantity, aes(x = TotalInvoiceQuantity)) +
  geom_histogram(binwidth = 50, fill = "lightblue", color = "darkblue") + # Adjust binwidth as needed
  labs(title = "Histogram of Total Invoice Quantity",
       x = "Total Invoice Quantity", 
       y = "Frequency") +
  theme_minimal()
plot(p2)

figure_file = here("results","figures","Quantity_distribution.png")
ggsave(filename = figure_file, plot=p2)
```

Histogram is also plotted for price range of the products.Most of the products fall below the price of 5 sterling pounds.

```{r}
# Create histogram of Unit Price

p3 <- ggplot(mydata, aes(x = UnitPrice)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "darkblue") + # Adjust binwidth as needed
  labs(title = "Histogram of Unit Price",
       x = "Unit Price", 
       y = "Frequency") +
  theme_minimal()
plot(p3)

figure_file = here("results","figures","UnitPrice_distribution.png")
ggsave(filename = figure_file, plot=p3)
```

Density plot is created for total purchase by each Customer during the period of the data. The average total purchase amount appears to be less than 5000 sterling pounds during that period.

```{r}
# Aggregate sales by CustomerID
sales_per_customer <- mydata %>%
  group_by(CustomerID) %>%
  summarise(TotalSales = sum(Sales, na.rm = TRUE)) %>%
  arrange(desc(TotalSales))

#Density Plot of Sales per Customer
# Assuming sales_per_customer is already calculated
p4 <- ggplot(sales_per_customer, aes(x = TotalSales)) +
  geom_density(fill = "blue", alpha = 0.5) +
  labs(title = "Density of Sales Across Customers", x = "Total Sales", y = "Density")

plot(p4)

figure_file = here("results","figures","Customer_AggregateSales.png")
ggsave(filename = figure_file, plot=p4)

```

Next, histogram of purchase frequency of each customer is created by grouping each customer by invoice no.The frequency appears to be less than 5 times for most of the customers.

```{r}
# Calculate purchase frequency per customer
purchase_frequency <- mydata %>%
  distinct(CustomerID, InvoiceNo) %>% # Ensure unique InvoiceNo per CustomerID
  count(CustomerID, name = "PurchaseFrequency") # Count the number of invoices per customer
library(ggplot2)

# Histogram of purchase frequencies
p5 <- ggplot(purchase_frequency, aes(x = PurchaseFrequency)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "darkblue") + # Adjust binwidth as needed
  labs(title = "Histogram of Customer Purchase Frequencies",
       x = "Purchase Frequency", 
       y = "Count of Customers") +
  theme_minimal()
plot(p5)
figure_file = here("results","figures","Customer_purchasefreq.png")
ggsave(filename = figure_file, plot=p5)

```

Next, total purchase by a customer was plotted against purchase frequency.A few outliers were noted which needs to be addressed while model building.

```{r}
# Calculate purchase frequency and total invoice value per customer
customer_stats <- mydata %>%
  group_by(CustomerID) %>%
  summarise(PurchaseFrequency = n_distinct(InvoiceNo),
            TotalInvoiceValue = sum(Sales, na.rm = TRUE)) %>%
  ungroup()  # Ensure the data is ungrouped for further analysis

# Scatter plot of Total Invoice Value by Purchase Frequency
p6 <- ggplot(customer_stats, aes(x = PurchaseFrequency, y = TotalInvoiceValue)) +
  geom_point(alpha = 0.5) +  # Adjust alpha for point transparency, if needed
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add a linear regression line without standard error
  labs(title = "Total Invoice Value by Purchase Frequency",
       x = "Purchase Frequency",
       y = "Total Invoice Value") +
  theme_minimal()
plot(p6)

figure_file = here("results","figures","Sales_trend.png")
ggsave(filename = figure_file, plot=p6)

```

Bar chart of Sales Distribution by Country

```{r}
p8 <- mydata %>% 
  group_by(Country) %>% 
  summarise(TotalSales = sum(Sales)) %>%
  ggplot(aes(x = reorder(Country, TotalSales), y = TotalSales, fill = Country)) + 
  geom_bar(stat = "identity") + 
  coord_flip() + 
  theme(legend.position = "none")

plot(p8)

figure_file = here("results","figures","Countrywise_Sales.png")
ggsave(filename = figure_file, plot=p8)
```

# Notes

For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.

### 2024-02-24 R/processing-code/data_without_uk-v1.qmd
Project Part 2 test

---
title: "An example cleaning script"
author: "Andreas Handel"
date: "2023-01-03"
output: html_document
---

# Processing script

This contains the same code and comments/information as `processingcode.R`.

This just shows it as Quarto file, to give you an idea how to do it in a setup that combines code and text in a single file.

See the other Quarto file for my currently preferred approach of pulling code from the R script into the Quarto file.

# Setup

Load needed packages. make sure they are installed.

```{r}
library(readxl) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
```

# Data loading

Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so:
package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives",
but it adds typing. You can do it either way.

```{r}
# path to data
# note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","superstoredata.csv")
raw_data <- read.csv(data_location)
```

# Check data

Several ways of looking at the data

```{r}
dplyr::glimpse(raw_data)
summary(raw_data)
head(raw_data)
skimr::skim(raw_data)
```

# Cleaning

The data inspection showed missing values in CustomerID, issues with signs and distribution of Quantity and Sales and format of date. It also pointed out that InvoiceNo, StockCode, Description and Country are coded as character variables in the data which need to be changed to factors for better data manipulation.

First, there are 135080 missing values in CustomerID. Customer behavior is a crucial part of this analysis which cannot be tracked without the unique customerID The observations without CustomerID do not help in doing meaningful analysis and hence dropped. Further, CustomerID is coded as a numerical variable in the data and it needs to be converted to factor variable. Country, Description and StockCode are also converted to factors from character variables.

```{r}
# Subset the data by excluding rows where CustomerID is NA
  clean_data_step1 <- raw_data %>%
  filter(!is.na(CustomerID)) %>%
  mutate(InvoiceNo = as.factor(InvoiceNo),
         CustomerID = as.factor(CustomerID),
         Country = as.factor(Country),
         Description = as.factor(Description),
         StockCode = as.factor(StockCode))

# Check the dimensions of the cleaned dataset
dim(clean_data_step1)
```

There are negative values in quantity and sales due to order cancellation. An inspection is required to check how many of such observations are in the data set. It is found that there are 8905 observations with negative Quantity and Sales.

```{r}
#Filter observations where Quantity and Sales are negative
negative_sales <- clean_data_step1 %>%
  filter (Quantity < 0 & Sales < 0)
# Count the negative sales
count(negative_sales)
```

Next step is to find matching original sales entries corresponding to the negative entries based on CustomerID, StockCode, Quantity and Sales figures. 8905 matches were found, which is equal to the no. of observations with negative Quantity and Sales figures.

```{r}
positive_sales <- clean_data_step1 %>%
  filter(Quantity > 0 & Sales > 0)
count(positive_sales)

#Checking for matches
matched_sales <- negative_sales %>%
  rowwise() %>%
  mutate(match_found = any(StockCode %in% positive_sales$StockCode &
                             CustomerID %in% positive_sales$CustomerID))

count(matched_sales)
```

The next attempt is to match returns to their original sales and get rid of both matched transactions. However, this needs further work as the below codes are resulting 3072 observations in original sales instead of 8905.

```{r}
# Add an identifier to each row in both data sets to track them (Marked as comment because this chunk of code needs further work and also takes a longer time to run)

"negative_sales <- negative_sales %>% mutate(neg_id = row_number())
positive_sales <- positive_sales %>% mutate(pos_id = row_number())

# Initialize a vector to keep track of matched negative_sales' IDs
matched_neg_ids <- integer(0)

# Initialize an empty data frame to store matches
matches <- data.frame(neg_id = integer(), pos_id = integer())"

# Iterate over positive_sales to find matches in  (Marked as comment because it takes a longer time to run this code)
"for (pos_row in 1:nrow(positive_sales)) {
  pos_sale <- positive_sales[pos_row, ]

  # Attempt to match with negative_sales, excluding already matched rows
  potential_matches <- negative_sales %>%
    filter(
      !neg_id %in% matched_neg_ids, # Exclude already matched
      CustomerID == pos_sale$CustomerID,
      StockCode == pos_sale$StockCode,
      Quantity == -pos_sale$Quantity, # Matching negative of Quantity
      Sales == -pos_sale$Sales # Matching negative of Sales
    )

  if (nrow(potential_matches) > 0) {
    # If a match is found, take the first one
    first_match <- potential_matches[1, ]
    matched_neg_ids <- c(matched_neg_ids, first_match$neg_id)

    # Record the match
    matches <- rbind(matches, data.frame(neg_id = first_match$neg_id, pos_id = pos_sale$pos_id))
  }
}

# Extract matched rows based on IDs 
#matched_negative_sales <- negative_sales %>% filter(neg_id %in% matches$neg_id)
#matched_positive_sales <- positive_sales %>% filter(pos_id %in% matches$pos_id)"
```

Observations with negative Quantity and Sales are dropped though unsuccessful in finding exact no. of matching original transactions.

```{r}
clean_data_step2 <- clean_data_step1 %>%
  filter(Quantity > 0, Sales > 0)
```

Inspection is required in the distribution of Quantity, unit price and Sales, the higher values of these variable are concentrated at the 100 percentile. I focused on distribution of UnitPrice at small intervals from 75% to 100%.

```{r}
# Calculate higher percentiles within the 3rd quartile to max range
percentiles <- quantile(clean_data_step2$UnitPrice, probs = c(0.75, 0.90, 0.95, 0.99, 0.995, 0.996, 0.997, 0.998, 0.999, 1))
print(percentiles)
```

The results indicate that there are only limited items that are above the unit price of 41.89. Observations wherein UnitPrice is above 50 are filtered. It resulted into 241 observations with 14 levels of Description.

```{r}
# Filter rows where UnitPrice is greater than 50

UnitPrice_above50 <- clean_data_step2[clean_data_step2$UnitPrice > 50, ]

#check the no. of categoriesis in description
n_distinct(UnitPrice_above50$Description)
```

The components of description was checked using the unique function which resulted into 3896 levels for 241 observations. With the help of ChatGPT, it was found that in R, when a subset of a data frame is created, the factor levels in the subset are not automatically dropped even if they are not present in the subset. To resolve this, the unused levels required to be dropped using the droplevels() function.

```{r}
#Attempt without dropping the unused factor levels
unique(UnitPrice_above50$Description)

# Dropping unused factor levels in the Description column
UnitPrice_above50$Description <- droplevels(UnitPrice_above50$Description)

# Finding the unique categories again
unique_categories <- unique(UnitPrice_above50$Description)

# Print the unique categories
print(unique_categories)

```

The above code resulted into 14 categories in product description. Most of the high UnitPrice are related to the descriptions such as Manual, POSTAGE, DOTCOM POSTAGE, appearing to be related to logistics. I got rid of observations with all those levels in description since those are less likely to be directly related to customer purchase.

```{r}
# Remove observations with specified descriptions
clean_data_step3 <- clean_data_step2[!grepl("Manual", clean_data_step2$Description) &
                                !grepl("POSTAGE", clean_data_step2$Description) &
                                !grepl("DOTCOM POSTAGE", clean_data_step2$Description), ]

# Inspect the cleaned data
skimr::skim(clean_data_step3)

```

Next the InovoiceDate is in character format, and needs to be converted to date format.

```{r}
clean_data_step3$InvoiceDate <- as.POSIXct(clean_data_step3$InvoiceDate, format = "%m/%d/%Y %H:%M")

skimr::skim(clean_data_step3)

```

The data 'clean_data_step3 is clean now. The purpose of this project is to study purchase pattern of customers residing outside of UK.The final data set data_without_uk excludes all sales in the UK. It has 42455 observations.

```{r}
# Exclude "United Kingdom" from the data set
data_without_uk <- subset(clean_data_step3, Country != "United Kingdom")

skimr::skim(data_without_uk)
```

A spike in Quantity and Sales observed between 75 and 100 percentile. The store also sells merchandises in wholesale quantities. As this study focuses on end user purchases, observations appearing to sales to wholesales need to be dropped. This calls for a close view of distribution of Sales by Invoice value.

```{r}

# Sum sales by InvoiceNo to get total invoice value
invoice_totals <- data_without_uk %>%
  group_by(InvoiceNo) %>%
  summarise(TotalInvoiceValue = sum(Sales, na.rm = TRUE))

# Calculate percentiles of total invoice value
percentiles_InvoiceValue <- quantile(invoice_totals$TotalInvoiceValue, 
                                     probs = c(0.25, 0.50, 0.75, 0.80, 0.85, 0.86, 0.87, 0.88, 0.89, 0.90, 0.91, 0.92, 0.93, 0.94, 0.95, 0.99, 1), 
                                     na.rm = TRUE)

print(percentiles_InvoiceValue)
```
Checking the distribution of quantity.
```{r}
library(dplyr)

# Sum quantities by InvoiceNo to get total invoice quantity
invoice_quantity_totals <- data_without_uk %>%
  group_by(InvoiceNo) %>%
  summarise(TotalInvoiceQuantity = sum(Quantity, na.rm = TRUE))

# Calculate percentiles of total invoice quantity
percentiles_InvoiceQuantity <- quantile(invoice_quantity_totals$TotalInvoiceQuantity, 
                                        probs = c(0.25, 0.50, 0.75, 0.80, 0.85, 0.86, 0.87, 0.88, 0.89, 0.90, 0.91, 0.92, 0.93, 0.94, 0.95, 0.99, 1), 
                                        na.rm = TRUE)

print(percentiles_InvoiceQuantity)
```
Invoice Value and Quantity have different distribution pattern. Based on the products the store sales, it is difficult to make a decision based on the Invoice Quantity. So, the sample is filtered to 90 percentile of Invoice Value subject to further investigation.

```{r}
# Calculate the 90th percentile of Invoice value
quantity_90th_percentile <- quantile(data_without_uk$Quantity, 0.9)

# Create a new data frame excluding observations with Quantity above the 90th percentile
data_without_uk1 <- data_without_uk %>%
  filter(Quantity <= quantity_90th_percentile)

```

Assigning a final name to the subset to make it easier to add further cleaning steps.

```{r}
processed_superstore <- data_without_uk1
```

# Save data

Finally, we save the clean data as RDS file. I suggest you save your processed and cleaned data as RDS or RDA/Rdata files. 
This preserves coding like factors, characters, numeric, etc.  If you save as CSV, that information would get lost.
However, CSV is better for sharing with others since it's plain text. If you do CSV, you might want to write down somewhere what each variable is.

See here for some suggestions on how to store your processed data:
http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

```{r}
save_data_location <- here::here("data","processed-data","processed_superstore.rds")
saveRDS(processed_superstore, file = save_data_location)

```

# Notes

Removing anyone who had "faulty" or missing data is one approach. It's often not the best. based on your question and your analysis approach, you might want to do cleaning differently (e.g. keep individuals with some missing information).

### 2024-02-24 R/processing-code/data_without_uk-v2.qmd
Project Part 2 test

---
title: "An example cleaning script with code pulled in"
author: "Andreas Handel"
date: "2023-01-03"
output: html_document
---

# Processing script with code pulled in

This is essentially the same as the other Quarto file, but now the code is not inside this file. Instead, it is pulled in from the R script using the code chunk labels.

# Setup

This needs to run to load the R script so we can include the code chunks below.

Note that you can pull in more than one R script if you want.

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('processingcode.R')
```

Load needed packages. Instead of having R commands here, this just pulls in the code from the R script we loaded above and labelled.

```{r, packages, message = FALSE, warning = FALSE}
```

# Data loading

Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so:
package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives",
but it adds typing. You can do it either way.

```{r, loaddata}
```

# Explore data

Several ways of looking at the data

```{r, exploredata}
```

# Cleaning

By inspecting the data as done above, we find some problems that need addressing:

First, there is an entry for height which says "sixty" instead of a number. 
Does that mean it should be a numeric 60? It somehow doesn't make sense since the weight is 60kg, which can't happen for a 60cm person (a baby).
Since we don't know how to fix this, we might decide to remove the person. This "sixty" entry also turned all Height entries into characters instead of numeric. That conversion to character also means that our summary function isn't very meaningful. So let's fix that first.

```{r, cleandata1}
```

Now we see that there is one person with a height of 6. That could be a typo, or someone mistakenly entered their height in feet. Since we unfortunately don't know, we might need to remove this person, which we'll do here.

```{r, cleandata2}
```

Height values seem ok now.

Now let's look at the `Weight` variable. There is a person with weight of 7000, which is impossible, and one person with missing weight.
To be able to analyze the data, we'll remove those individuals as well.

Note that removing anyone who had "faulty" or missing data is one approach. It's often not the best. Based on your question and your analysis approach, you might want to do cleaning differently (e.g. keep individuals with some missing information).

```{r, cleandata3}
```

Now checking the `Gender` variable.

```{r, cleandata4}
```

```{r, cleandata5}
```

# Save data

All done, data is clean now. Time to save.
Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.

Finally, we save the clean data as RDS file. I suggest you save your processed and cleaned data as RDS or RDA/Rdata files. 
This preserves coding like factors, characters, numeric, etc.  If you save as CSV, that information would get lost.
However, CSV is better for sharing with others since it's plain text. If you do CSV, you might want to write down somewhere what each variable is.

See here for some suggestions on how to store your processed data:
http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

```{r, savedata}
```

### 2024-02-24 products/manuscript/Manuscript.qmd
Project Part 2 test

@@
 
-This study intends to identify distinct customer groups within sales data of luxury goods, with each group comprising customers who exhibit similar characteristics in terms of purchase frequency and amounts.
+This study intends to identify distinct customer groups within sales data, with each group comprising customers who exhibit similar characteristics in terms of purchase frequency and amounts.
 Modeling customer spending as a mixture of gamma distributions helps to identify different spending behaviors.
 The objective is to develop an Expectation Maximization (EM) algorithm which segregates groups based on their spending patterns, considering the individual purchases transactions made by each customer.
+
+The exploration begins with applying commonly used alternative models to a publicly available data and ultimately applying the gamma mixture model to this data.
 
@@
 
-This study employs simulated data to assess the effectiveness of the EM algorithm. Following successful testing, the refined algorithm will be applied to actual sales data pertaining to luxury goods.
+This study employs sales data, 'superstoredata' and focuses on international sales of the online retail store.
+
+Initially this data was downloaded from Kaggle.com to it's usability for the project.
+Upon confirming the usability and revisiting the website seeking for further details, it could not be found in Kaggle.com.
+After few Google searches, the data and details were found on UC Irvine Machine Learning Repository [https://archive.ics.uci.edu/dataset/352/online+retail].
+As per the website, this is a transnational data set containing all the transactions occurring between 12/1/2010 and 12/09/2011 for a UK-based and registered non-store online retail.
+The company primarily sells unique all-occasion gifts and many customers of the company are wholesalers.
+
+This data has nine variables. The details are explained below:
+
+(1)	InvoiceNo: It is a categorical variable with a 6-digit integral number uniquely assigned to each transaction. If this code starts with the letter ‘c’, it indicates a cancellation. 
+(2)	StockCode: It is a categorical variable with a 5-digit integral number uniquely assigned to each distinct product.
+(3)	Description: This represents Product name and is a categorical variable.
+(4)	Quantity: It represents quantities of each product (item) per transaction and is an integer.
+(5)	InvoiceDate: It represents the day and time each transaction was generated.
+(6)	UnitPrice: It represents product price per unit in sterling.
+(7)	CustomerID: It is a categorical variable with a 5-digit integral number uniquely assigned to each customer.
+(8)	Country: It represents the name of the country where each customer resides.
+
+The study will also use simulated data to formulate the EM algorithm. Following successful testing, the refined algorithm will be applied to the sales data.
 
@@
 
-This study aims to answer the question, "How effective is the Expectation Maximization (EM) algorithm in identifying distinct customer groups within luxury goods sales data when accounting for individual spending by each customer, as demonstrated by initial validation using simulated data?"
+This study aims to answer the question, "How effective is the Expectation Maximization (EM) algorithm in identifying distinct customer groups when accounting for individual spending by each customer, as demonstrated by initial validation using simulated data?"
 
@@
 
+This study employs sales data, 'superstoredata' and focuses on international sales of the online retail store.
+
+Initially this data was downloaded from Kaggle.com to it's usability for the project.
+Upon confirming the usability and revisiting the website seeking for further details, it could not be found in Kaggle.com.
+After few Google searches, the data and details were found on UC Irvine Machine Learning Repository [https://archive.ics.uci.edu/dataset/352/online+retail].
+As per the website, this is a transnational data set containing all the transactions occurring between 12/1/2010 and 12/09/2011 for a UK-based and registered non-store online retail.
+The company primarily sells unique all-occasion gifts and many customers of the company are wholesalers.
+
+This data has nine variables. The details are explained below:
+
+(1)	InvoiceNo: It is a categorical variable with a 6-digit integral number uniquely assigned to each transaction. If this code starts with the letter ‘c’, it indicates a cancellation. 
+(2)	StockCode: It is a categorical variable with a 5-digit integral number uniquely assigned to each distinct product.
+(3)	Description: This represents Product name and is a categorical variable.
+(4)	Quantity: It represents quantities of each product (item) per transaction and is an integer.
+(5)	InvoiceDate: It represents the day and time each transaction was generated.
+(6)	UnitPrice: It represents product price per unit in sterling.
+(7)	CustomerID: It is a categorical variable with a 5-digit integral number uniquely assigned to each customer.
+(8)	Country: It represents the name of the country where each customer resides.
+
+The study will also use simulated data to formulate the EM algorithm. Following successful testing, the refined algorithm will be applied to the sales data.
+
 ## Statistical analysis
@@
 #| echo: FALSE
-knitr::include_graphics(here("results","height-weight-stratified.png"))
+knitr::include_graphics(here("results", "figures", "Sales_distribution.png"))
 ```
@@
 #| echo: FALSE
-resulttable2 = readRDS(here("results","tables-files","resulttable2.rds"))
+resulttable2 = readRDS(here("results","tables","summarytable.rds"))
 knitr::kable(resulttable2)

### 2024-02-24 R/analysis-code/superstore.R

@@ -1,0 +1,181 @@
+library(readxl) #for loading Excel files
+library(dplyr) #for data processing/cleaning
+library(tidyr) #for data processing/cleaning
+library(skimr) #for nice visualization of data
+library(here)  #to set paths
+library(ggplot2)
+
+#path to data
+data_location <- here::here("data","raw-data","superstoredata.csv")
+raw_data <- read.csv(data_location)
+#Check data
+dplyr::glimpse(raw_data)
+head(raw_data)
+tail(raw_data)
+summary(raw_data)
+str(raw_data)
+skimr::skim(raw_data) #gives missing nos., completion rate, min, max, empty, n_unique, whitespace,mean, standard deviaiont and percentile 
+
+#By inspecting the data, we find some problems that need to be addressed:
+
+#First, there are 135080 missing values in CustomerID. The analysis looks into customer behaviors for which the unique customer ID is an essential part of the study.
+#Hence, the observations without CustomerID do not provide meaningful analysis. Such observations are dropped for analyses. Further 
+#CustomerID is loaded as a numerical variable in the data and it needs to be converted to categorical/factor variable
+#Further Country is also converted to a factor.
+
+
+# Subset the data by excluding rows where CustomerID is NA
+  clean_data_step1 <- raw_data %>%
+  filter(!is.na(CustomerID)) %>%
+  mutate(CustomerID = as.factor(CustomerID),
+         Country = as.factor(Country),
+         Description = as.factor(Description),
+         StockCode = as.factor(StockCode))
+
+# Check the dimensions of the cleaned dataset
+dim(clean_data_step1)
+
+#There are negative values in quantity, unit price and sales due to order cancellation. Let's inspect how many of such data are there
+# Filter observations where Quantity, UnitPrice, and Sales are negative
+negative_sales <- clean_data_step1 %>%
+  filter (Quantity < 0 & Sales < 0)
+
+# View the filtered observations
+count(negative_sales)
+#There are 8905 observations where either Quantity and Sales are negative.
+
+#I check to see if there is a corresponding sales (+ve entry) for these returned items. 
+
+positive_sales <- clean_data_step1 %>%
+  filter(Quantity > 0 & Sales > 0)
+count(positive_sales)
+
+#Attempt to match each negative Sale/Quantity with a corresponding positive sale by CustomerID and StockCode
+matched_sales <- negative_sales %>%
+  rowwise() %>%
[…]

### 2024-02-24 R/eda-code/edacode_superstore.R

@@ -1,0 +1,162 @@
+## ---- packages --------
+#load needed packages. make sure they are installed.
+library(here) #for data loading/saving
+library(dplyr)
+library(skimr)
+library(ggplot2)
+
+## ---- loaddata --------
+#Path to data. Note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed-data","processed_superstore.rds")
+#load data
+mydata <- readRDS(data_location)
+
+## ---- table1 --------
+summary_df = skimr::skim(mydata)
+print(summary_df)
+# save to file
+summarytable_file = here("results", "summarytable.rds")
+saveRDS(summary_df, file = summarytable_file)
+
+## ---- Invoice Sales --------
+
+# Histogram of Invoice values
+## First calculate total invoice value per invoice
+invoice_values <- mydata %>%
+  group_by(InvoiceNo) %>%
+  summarise(TotalInvoiceValue = sum(Sales, na.rm = TRUE))
+
+##Create histogram
+p1 <- ggplot(invoice_values, aes(x = TotalInvoiceValue)) +
+  geom_histogram(binwidth = 50, fill = "lightblue", color = "darkblue") + # Adjust binwidth as needed
+  labs(title = "Histogram of Total Invoice Values",
+       x = "Total Invoice Value", 
+       y = "Frequency") +
+  theme_minimal()
+plot(p1)
+
+figure_file = here("results","figures","Sales_distribution.png")
+ggsave(filename = figure_file, plot=p1) 
+
+# Histogram of Invoice Quantity
+## First calculate total invoice quantity per invoice
+invoice_quantity <- mydata %>%
+  group_by(InvoiceNo) %>%
+  summarise(TotalInvoiceQuantity = sum(Quantity, na.rm = TRUE))
+
+p2 <- ggplot(invoice_quantity, aes(x = TotalInvoiceQuantity)) +
+  geom_histogram(binwidth = 50, fill = "lightblue", color = "darkblue") + # Adjust binwidth as needed
+  labs(title = "Histogram of Total Invoice Quantity",
+       x = "Total Invoice Quantity", 
+       y = "Frequency") +
+  theme_minimal()
+plot(p2)
+
+figure_file = here("results","figures","Quantity_distribution.png")
+ggsave(filename = figure_file, plot=p2)
+
+## ---- UnitPrice-Price range of products --------
+# Create histogram of Unit Price
+
[…]

### 2024-03-07 R/eda-code/eda_superstore.qmd
Project Part 2 - Rev

@@
 ---
-title: "An example exploratory analysis script"
-author: "Andreas Handel"
-date: "10/28/2021"
+title: "Exploratory Analysis of Superstore"
+author: "Malika Dhakhwa"
+date: "2024-02-23"
 output: html_document
@@
 ```{r}
+
+#| message: false
+#| warning: false
 #load needed packages. 
 library(here) #for data loading/saving
-library(dplyr)
-library(skimr)
-library(ggplot2)
+library(dplyr) #for data processing/cleaning
+library(skimr) #for nice visualization of data 
+library(ggplot2) # for visualization
+library(tibble) # For rownames_to_column
+library(gt) # for table
+library(gridExtra)
+library(webshot2) #Capture as an Image the gttable before converting it to pdf
 ```
 
-Data is loaded.
+Data is loaded using here function.
 
@@
 
+# Descriptive Statistics tables
+
+It is important to note that this data accounts sales of a particular item at a time as an observation leading different items sold in the same Invoice to appear as separate observations.
+As such most of the Invoice numbers appear several times in different observations of the data.
+A meaningful analysis of customer spending pattern require the Invoice Sales Amount, total purchase by a Customer during a period, frequency of purchase by a customer during a period or a combination of those.
+A table is created which provides the descriptive statistics of sales by Invoice value, Customer purchase and frequency of customer purchase.
+This table reveals that there were 1479 transactions to 414 different customers during 2010-12-01 and	2011-12-09.
+
+```{r}
+
+# Aggregate total sales by InvoiceNo for calculating mean, standard deviation, minimum and maximum
+Sales_by_Invoice <- mydata %>%
+  group_by(InvoiceNo) %>%
+  summarise(TotalSales = sum(Sales, na.rm = TRUE)) %>%
+  ungroup() %>% # Ensure we ungroup before further summarisation
+  summarise(
+    Count = n(),
+    Particulars = "Sales by Invoice",
+    Mean = round(mean(TotalSales, na.rm = TRUE), 2),
+    SD = round(sd(TotalSales, na.rm = TRUE), 2),
+    Min = round(min(TotalSales, na.rm = TRUE), 2),
+    Max = round(max(TotalSales, na.rm = TRUE), 2)
+  )
+
+# Aggregate total sales by CustomerID to calculate mean, standard deviation, minimum and maximum
+TotalSales_customer <- mydata %>%
+  group_by(CustomerID) %>%
+  summarise(TotalSales = sum(Sales, na.rm = TRUE)) %>%
+  ungroup() %>%
+  summarise(
+    Count = n(),
+    Particulars = "Cumulative Sales to Individual Customers",
+    Mean = round(mean(TotalSales, na.rm = TRUE), 2),
+    SD = round(sd(TotalSales, na.rm = TRUE), 2),
+    Min = round(min(TotalSales, na.rm = TRUE), 2),
+    Max = round(max(TotalSales, na.rm = TRUE), 2)
+  )
+
+# Frequency of sales per CustomerID (i.e., the count of unique InvoiceNo per CustomerID)for calculating mean, standard deviation, minimum and maximum
+
+sales_frequency_per_customer <- mydata %>%
+  group_by(CustomerID) %>%
+  summarise(FrequencyOfSales = n_distinct(InvoiceNo)) %>%
+  ungroup() %>%
+  summarise(
+    Count = n(),
+    Mean = round(mean(FrequencyOfSales, na.rm = TRUE), 2),
+    SD = round(sd(FrequencyOfSales, na.rm = TRUE), 2),
+    Min = round(min(FrequencyOfSales, na.rm = TRUE), 2),
+    Max = round(max(FrequencyOfSales, na.rm = TRUE), 2)
+  ) %>%
+  mutate(Particulars = "Sales frequency to Individual Customers")
+
+# Combine the statistics into a single data frame
+combined_stats <- bind_rows(Sales_by_Invoice, TotalSales_customer, sales_frequency_per_customer) %>%
+  select(Particulars, everything())  # Ensure Particulars column is at the front
+
+# Create and format the gt table
+gt_table <- gt(combined_stats) %>%
+  tab_header(
+    title = "Descriptive Statistics of Sales") %>%
+  cols_label(
+    Particulars = "Particulars",
+    Count = "Count",
+    Mean = "Mean",
+    SD = "St.Dev.",
+    Min = "Min",
+    Max = "Max"
+  ) %>%
+  cols_align(
+    align = "center",
+    columns = c("Count", "Mean", "SD", "Min", "Max")
+  )
+
+# Display the gt table
+print(gt_table)
+
+gt_table %>%
+  gtsave(here("results","tables", "gt_table.png"))
+
+```
 # Data exploration through figures
 
-Histograms were plotted for the continuous outcomes.
-The data contains individual observations for sales of different stocks in a single invoice.
-As this project intends to study customer spending, Invoice totals are calculated for purchases by each customer at a time to visualize distribution of customer spending each time.The histogram is heavily right-skewed with higher frequencies at less than 500.
+Next, density plots were created for Sales across customers and Invoice Values.
+Both figures are heavily right-skewed.The averages of total sales to a customer and total invoice values appear to be less than 2500 and 1250 sterling pounds respectively supporting the mean values of 2025 and 563 in the descriptive table.
 
-```{r}
-# Histogram of Invoice values
-## First calculate total invoice value per invoice
-invoice_values <- mydata %>%
-  group_by(InvoiceNo) %>%
-  summarise(TotalInvoiceValue = sum(Sales, na.rm = TRUE))
-
-##Create histogram
-p1 <- ggplot(invoice_values, aes(x = TotalInvoiceValue)) +
-  geom_histogram(binwidth = 50, fill = "lightblue", color = "darkblue") + # Adjust binwidth as needed
-  labs(title = "Histogram of Total Invoice Values",
-       x = "Total Invoice Value", 
-       y = "Frequency") +
-  theme_minimal()
-plot(p1)
-
-figure_file = here("results","figures","Sales_distribution.png")
-ggsave(filename = figure_file, plot=p1) 
-```
-
-Similarly, histogram plot was created for total Quantity of purchase per invoice which is also heavily right-skewed.
-
-```{r}
-## First calculate total invoice quantity per invoice
-invoice_quantity <- mydata %>%
-  group_by(InvoiceNo) %>%
-  summarise(TotalInvoiceQuantity = sum(Quantity, na.rm = TRUE))
-
-p2 <- ggplot(invoice_quantity, aes(x = TotalInvoiceQuantity)) +
-  geom_histogram(binwidth = 50, fill = "lightblue", color = "darkblue") + # Adjust binwidth as needed
-  labs(title = "Histogram of Total Invoice Quantity",
-       x = "Total Invoice Quantity", 
-       y = "Frequency") +
-  theme_minimal()
-plot(p2)
-
-figure_file = here("results","figures","Quantity_distribution.png")
-ggsave(filename = figure_file, plot=p2)
-```
-
-Histogram is also plotted for price range of the products.Most of the products fall below the price of 5 sterling pounds.
-
-```{r}
-# Create histogram of Unit Price
-
-p3 <- ggplot(mydata, aes(x = UnitPrice)) +
-  geom_histogram(binwidth = 1, fill = "lightblue", color = "darkblue") + # Adjust binwidth as needed
-  labs(title = "Histogram of Unit Price",
-       x = "Unit Price", 
-       y = "Frequency") +
-  theme_minimal()
-plot(p3)
-
-figure_file = here("results","figures","UnitPrice_distribution.png")
-ggsave(filename = figure_file, plot=p3)
-```
-
-Density plot is created for total purchase by each Customer during the period of the data. The average total purchase amount appears to be less than 5000 sterling pounds during that period.
+A histogram of purchase frequency of each customer is created by grouping each customer by invoice no.The average frequency appears to be less than 5 times for most of the customers which is also observed in the descriptive statistics table.
 
@@
 # Aggregate sales by CustomerID
-sales_per_customer <- mydata %>%
+TotalSales_customer <- mydata %>%
   group_by(CustomerID) %>%
@@
 # Assuming sales_per_customer is already calculated
-p4 <- ggplot(sales_per_customer, aes(x = TotalSales)) +
-  geom_density(fill = "blue", alpha = 0.5) +
-  labs(title = "Density of Sales Across Customers", x = "Total Sales", y = "Density")
+p1 <- ggplot(TotalSales_customer, aes(x = TotalSales)) +
+  geom_density(fill = "skyblue", alpha = 0.5) +
+  labs(title = "Density Plot of Sales across Customers", x = "Total Sales", y = "Density")
 
-plot(p4)
+# Histogram of Sales by Invoice values
+## First calculate total invoice value per invoice
+Sales_by_Invoice <- mydata %>%
+  group_by(InvoiceNo) %>%
+  summarise(TotalInvoiceValue = sum(Sales, na.rm = TRUE))
 
-figure_file = here("results","figures","Customer_AggregateSales.png")
-ggsave(filename = figure_file, plot=p4)
+##Create histogram
+p2 <- ggplot(Sales_by_Invoice, aes(x = TotalInvoiceValue)) +
+  geom_density(fill = "lightgreen", alpha = 0.5) +
+  labs(title = "Density Plot of  Invoice Sales",
+       x = "Invoice Value", 
+       y = "Density") +
+  theme_minimal()
 
-```
-
-Next, histogram of purchase frequency of each customer is created by grouping each customer by invoice no.The frequency appears to be less than 5 times for most of the customers.
-
-```{r}
 # Calculate purchase frequency per customer
@@
 # Histogram of purchase frequencies
-p5 <- ggplot(purchase_frequency, aes(x = PurchaseFrequency)) +
-  geom_histogram(binwidth = 1, fill = "lightblue", color = "darkblue") + # Adjust binwidth as needed
+p3 <- ggplot(purchase_frequency, aes(x = PurchaseFrequency)) +
+  geom_histogram(binwidth = 1, fill = "salmon", color = "darkblue") + # Adjust binwidth as needed
   labs(title = "Histogram of Customer Purchase Frequencies",
-       x = "Purchase Frequency", 
+       x = "No. of Purchases", 
        y = "Count of Customers") +
   theme_minimal()
-plot(p5)
-figure_file = here("results","figures","Customer_purchasefreq.png")
-ggsave(filename = figure_file, plot=p5)
+
+# Modify p1 to adjust title font size
+p1 <- p1 + theme(plot.title = element_text(size = 11)) # Adjusting font size of the title of p1
+# Modify p2 to adjust title font size
+p2 <- p2 + theme(plot.title = element_text(size = 11)) # Adjusting font size of the title of p2
+# Modify p3 to adjust title font size
+p3 <- p3 + theme(plot.title = element_text(size = 11)) # Adjusting font size of the title of p3
+
+plot1<-grid.arrange(p1, p2, p3, ncol = 2, nrow=2)
+
+figure_file = here("results","figures","Sales_distribution.png")
+ggsave(filename = figure_file, plot=plot1)
 
@@
 
-Next, total purchase by a customer was plotted against purchase frequency.A few outliers were noted which needs to be addressed while model building.
+A histogram is also plotted for price range of the products.Most of the products fall below the price of 12.50 sterling pounds.
 
 ```{r}
+# Create histogram of Unit Price
+
+p4 <- ggplot(mydata, aes(x = UnitPrice)) +
+  geom_histogram(binwidth = 1, fill = "lightblue", color = "darkblue") + # Adjust binwidth as needed
+  labs(title = "Histogram of Unit Price",
+       x = "Unit Price", 
+       y = "Frequency") +
+  theme_minimal()
+
+p4
+
+figure_file = here("results","figures","Price_range.png")
+ggsave(filename = figure_file, plot=p4)
+
+```
+Next, total purchase by a customer was plotted against purchase frequency.A few outliers were noted.
+
+```{r}
+#| message: false
+#| warning: false
 # Calculate purchase frequency and total invoice value per customer
@@
 # Scatter plot of Total Invoice Value by Purchase Frequency
-p6 <- ggplot(customer_stats, aes(x = PurchaseFrequency, y = TotalInvoiceValue)) +
+p5 <- ggplot(customer_stats, aes(x = PurchaseFrequency, y = TotalInvoiceValue)) +
   geom_point(alpha = 0.5) +  # Adjust alpha for point transparency, if needed
@@
   theme_minimal()
-plot(p6)
+plot(p5)
 
 figure_file = here("results","figures","Sales_trend.png")
-ggsave(filename = figure_file, plot=p6)
+ggsave(filename = figure_file, plot=p5)
 
@@
 
-Bar chart of Sales Distribution by Country
+Next, Bar chart is created for Sales Distribution by Country. The top five sales are in Germany, France, Spain, Switzerland and Japan.
 

### 2024-03-07 R/processing-code/processing_superstore-v1.qmd
Project Part 2 - Rev

---
title: "Cleaning data 'Superstore'"
author: "Malika Dhakhwa"
date: "2024-02-23"
output: html_document
---
# Setup

Install and load needed packages.

```{r}
#| message: false
#| warning: false
library(readxl) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(ggplot2) # for visualization
library(scales) # for labels
```

# Data loading
Loading data using here function.

```{r}
# path to data
# note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","superstoredata.csv")
raw_data <- read.csv(data_location)
```

# Check data

Checking data

```{r}
dplyr::glimpse(raw_data)
skimr::skim(raw_data)

```
# Cleaning

The purpose of this project is to study purchase pattern of customers residing outside of UK.The data set data_without_uk excludes all sales in the UK. It has 46431 observations with 9 variables. It is important to note that this data set accounts sales of a particular item as an observation. This means different items sold in the same Invoice appears as separate observations and it is most likely that an Invoice number could appear in separate observations.

```{r}
# Exclude "United Kingdom" from the data set
data_without_uk <- subset(raw_data, Country != "United Kingdom")
skimr::skim(data_without_uk)
```

The data inspection showed missing values in CustomerID, issues with signs and distribution of Quantity and Sales and format of date. It also pointed out that InvoiceNo, StockCode, Description and Country are coded as character variables in the data which need to be changed to factors for better data manipulation.

First, there are 1480 missing values in CustomerID. Customer behavior is a crucial part of this analysis which cannot be tracked without the unique customerID. The observations without CustomerID do not help in doing meaningful analysis and are dropped. Further, CustomerID is coded as a numerical variable in the original data and it needs to be converted to factor variable. Country, Description and StockCode are also converted to factors from character variables.

```{r}
# Subset the data by excluding rows where CustomerID is NA
  clean_data_step1 <- data_without_uk %>%
  filter(!is.na(CustomerID)) %>%   #excludes observation without CustomerID
  mutate(InvoiceNo = as.factor(InvoiceNo),  #converts InvoiceNo to a factor
         CustomerID = as.factor(CustomerID), #converts CustomerID to a factor
         Country = as.factor(Country), #converts Country to a factor
         Description = as.factor(Description), #converts Description to a factor
         StockCode = as.factor(StockCode)) #converts StockCode to a factor

# Checking the cleaned data
skimr::skim(clean_data_step1)

```

There are negative values in quantity and sales due to order cancellation. An inspection is carried out to check the number of such observations are in the data set. It is found that there are 1372 observations with negative Quantity and Sales.

```{r}
#Filter observations where Quantity and Sales are negative
negative_sales <- clean_data_step1 %>%
  filter (Quantity < 0 & Sales < 0)
# Count the negative sales
count(negative_sales)
```

Next step is to find matching original sales entries corresponding to the negative entries based on CustomerID, StockCode, Quantity and Sales figures. 1372 matches were found, which is equal to the no. of observations with negative Quantity and Sales figures.

```{r}
positive_sales <- clean_data_step1 %>%
  filter(Quantity > 0 & Sales > 0)
count(positive_sales)

#Checking for matches
matched_sales <- negative_sales %>%
  rowwise() %>%
  mutate(match_found = any(StockCode %in% positive_sales$StockCode &
                             CustomerID %in% positive_sales$CustomerID))

count(matched_sales)
```

#The next attempt is to match returns to their original sales and get rid of both matched transactions. However, this needs further work as the below codes are resulting 582 observations in original sales instead of 1372.

```{r}
# Add an identifier to each row in both data sets to track them (Marked as comment because this chunk of code needs further work and also takes a longer time to run)

"negative_sales <- negative_sales %>% mutate(neg_id = row_number())
positive_sales <- positive_sales %>% mutate(pos_id = row_number())

# Initialize a vector to keep track of matched negative_sales' IDs
matched_neg_ids <- integer(0)

# Initialize an empty data frame to store matches
matches <- data.frame(neg_id = integer(), pos_id = integer())

# Iterate over positive_sales to find matches in  (Marked as comment because it takes a longer time to run this code)
for (pos_row in 1:nrow(positive_sales)) {
  pos_sale <- positive_sales[pos_row, ]

  # Attempt to match with negative_sales, excluding already matched rows
  potential_matches <- negative_sales %>%
    filter(
      !neg_id %in% matched_neg_ids, # Exclude already matched ones
      CustomerID == pos_sale$CustomerID,
      StockCode == pos_sale$StockCode,
      Quantity == -pos_sale$Quantity, # Matching negative of Quantity
      Sales == -pos_sale$Sales # Matching negative of Sales
    )

  if (nrow(potential_matches) > 0) {
    # If a match is found, take the first one
    first_match <- potential_matches[1, ]
    matched_neg_ids <- c(matched_neg_ids, first_match$neg_id)

    # Record the match
    matches <- rbind(matches, data.frame(neg_id = first_match$neg_id, pos_id = pos_sale$pos_id))
  }
}

# Extract matched rows based on IDs 
matched_negative_sales <- negative_sales %>% filter(neg_id %in% matches$neg_id)
matched_positive_sales <- positive_sales %>% filter(pos_id %in% matches$pos_id)
print(matched_negative_sales)
print(matched_positive_sales)"
```

Observations with negative Quantity and Sales are dropped though unsuccessful in finding exact no. of matching original transactions. This brought down the total number of observations to 43563.

```{r}
clean_data_step2 <- clean_data_step1 %>%
  filter(Quantity > 0, Sales > 0)
skimr::skim(clean_data_step2)
```

Inspection is required in the distribution of Quantity, unit price and Sales, the higher values of these variable are concentrated at the higher percentiles. I focused on distribution of UnitPrice at small intervals from 75% to 100%.

```{r}
# Calculate higher percentiles within the 3rd quartile to max range
percentiles <- quantile(clean_data_step2$UnitPrice, probs = c(0.75, 0.90, 0.95, 0.99, 0.995, 0.996, 0.997, 0.998, 0.999, 1))
print(percentiles)
```

The results indicate that there are only limited items that are above the unit price of 25. Observations wherein UnitPrice is above 25 are filtered for further examination. It resulted into 433 observations.

```{r}
# Filter rows where UnitPrice is greater than 25

UnitPrice_above25 <- clean_data_step2[clean_data_step2$UnitPrice > 25, ]

str(UnitPrice_above25)
```

The components of description was checked using the unique function which resulted into 2903 levels for 433 observations. With the help of ChatGPT, it was found that in R, when a subset of a data frame is created, the factor levels in the subset are not automatically dropped even if they are not present in the subset. To resolve this, the unused levels required to be dropped using the droplevels() function.

```{r}
#Attempt without dropping the unused factor levels
unique_categories_initial <- unique(UnitPrice_above25$Description)
print(unique_categories_initial)

# Dropping unused factor levels in the Description column
UnitPrice_above25$Description <- droplevels(UnitPrice_above25$Description)

# Finding the unique categories again
unique_categories_final <- unique(UnitPrice_above25$Description)
print(unique_categories_final)
```

The above code resulted into 19 categories in product description which have the Unitprice above 25. Some of the high UnitPrice are related to the descriptions such as CARRIAGE, POSTAGE, Manual, appearing to be related to logistics. 1216 observations with such descriptions were dropped from the data since those are less likely to be directly related to customer purchase. The no. of observations came down to 42347.

```{r}
# Remove observations with specified descriptions
clean_data_step3 <- clean_data_step2[!grepl("Manual", clean_data_step2$Description) &
                                !grepl("POSTAGE", clean_data_step2$Description) &
                                !grepl("CARRIAGE", clean_data_step2$Description), ]

# Inspect the cleaned data
skimr::skim(clean_data_step3)

```
Next the InovoiceDate is in character format, and needs to be converted to date format.

```{r}
clean_data_step3$InvoiceDate <- as.POSIXct(clean_data_step3$InvoiceDate, format = "%m/%d/%Y %H:%M")

skimr::skim(clean_data_step3)

```

The store also sells merchandises in wholesale quantities. As this study focuses on end user purchases, observations of sales to wholesales need to be disregarded. A wholesaler potentially purchases in big volumes and in higher frequencies. In absence of clear information to identify wholesalers, a scatter plot of Total Invoice Value by Purchase Frequency could provide helpful information to identify potential wholesalers.

```{r}
# Calculate purchase frequency and total invoice value per customer
customer_stats <- clean_data_step3 %>%
  group_by(CustomerID) %>%  
  summarise(PurchaseFrequency = n_distinct(InvoiceNo),
            TotalInvoiceValue = sum(Sales, na.rm = TRUE)) %>%
  ungroup()  # Data ungrouped for further analysis

# Scatter plot of Total Invoice Value by Purchase Frequency
p1 <- ggplot(customer_stats, aes(x = PurchaseFrequency, y = TotalInvoiceValue)) +
  geom_point(alpha = 0.5) +  # Alpha can be adjusted as needed for point transparency
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Adding a linear regression line without standard error
  labs(title = "Total Invoice Value by Purchase Frequency",
       x = "Purchase Frequency",
       y = "Total Invoice Value") +
  scale_y_continuous(labels = label_comma()) +  # This line changes the y-axis labels from exp. number to actual value

    theme_minimal()
plot(p1)

figure_file = here("results","figures","InvoiceValue_by_frequency.png")
ggsave(filename = figure_file, plot=p1)

skimr::skim(customer_stats)

```

The scatter plot reveals that most of the customers made purchases less than 50 times during that period and most of the cumulative purchase values were less than 50,000. Assuming that the outliers represent purchases by wholesalers, observations with purchase frequencies above 50 and cumulative Invoice value above 50,000 are dropped from the data.

This involves first sub-setting customer_stats such that it includes only those observations where PurchaseFrequency <= 50 and TotalInvoiceValue <=50000 and finally, merging this sub-set with the original dataset (clean_data_step3).

```{r}

# Filtering the customer_stats to meet the conditions
filtered_customer_stats <- customer_stats %>%
  filter(PurchaseFrequency <= 50, TotalInvoiceValue <= 50000)

# Excluding purchase frequencies above 50 and cumulative invoice values above 50000 from the data
clean_data_step4 <- clean_data_step3 %>%
  semi_join(filtered_customer_stats, by='CustomerID')

skimr::skim(clean_data_step4)
```

This step excluded 9745 observations from the data for a final total observation of 32602. A similar scatter plot is plotted for the new data.

```{r}
# Calculate purchase frequency and total invoice value per customer
customer_stats4 <- clean_data_step4 %>%
  group_by(CustomerID) %>%
  summarise(PurchaseFrequency = n_distinct(InvoiceNo),
            TotalInvoiceValue = sum(Sales, na.rm = TRUE)) %>%
  ungroup()  # Ensure the data is ungrouped for further analysis

# Scatter plot of Total Invoice Value by Purchase Frequency
p2 <- ggplot(customer_stats4, aes(x = PurchaseFrequency, y = TotalInvoiceValue)) +
  geom_point(alpha = 0.5) +  # Adjust alpha for point transparency, if needed
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add a linear regression line without standard error
  labs(title = "Total Invoice Value by Purchase Frequency",
       x = "Purchase Frequency",
       y = "Total Invoice Value") +
  scale_y_continuous(labels = label_comma()) +  # This line changes the y-axis labels

    theme_minimal()
plot(p2)

```

The distribution of the data appears relatively homogeneous.

Assigning a final name to the subset to make it more relevant to the actual data.

```{r}
processed_superstore <- clean_data_step4

str(processed_superstore)
```

# Save data

Finally, the cleaned data is saved as RDS file. As learnt in the class, saving the cleaned data as RDS or RDA/Rdata files, preserves coding like factors, characters, numeric, etc.  Saving as CSV, that information would get lost. However, CSV is better for sharing with others since it's plain text.

See here for some suggestions on how to store your processed data:
http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

```{r}
save_data_location <- here::here("data","processed-data","processed_superstore.rds")
saveRDS(processed_superstore, file = save_data_location)

```

### 2024-03-07 products/manuscript/Manuscript.qmd
Project Part 2 - Rev

@@
     highlight-style: github
-bibliography: ../assets/dataanalysis-references.bib
-csl: ../assets/apa.csl
+bibliography: ../../assets/dataanalysis-references.bib
+csl: ../../assets/apa.csl
 ---
-
-The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
-
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
 
@@
 ## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
 
-Finite mixture models are used to analyze data presumably arising from several separate, yet unobserved, groups.
-Among these models, Gaussian mixture models and mixture of t-distributions are particularly prevalent.
-However, such models which are based on symmetric distribution often do not accurately represent data with asymmetric distribution.
-Skewed data can be more effectively managed using asymmetric distributions, thereby extending their applicability across a broader spectrum of model-based clustering scenarios [1].
-If the observed data is positive, distribution is right-skewed and assumed to come from a mixture distribution, employing mixture of gamma density is a rational choice [2].
+This project aims to gain knowledge of customer segmentation from the supervised and un-supervised learning perspectives.
+In the initial analysis phase, the study applies basic statistical techniques such as logistic regression to analyze a retail store’s data, to grasp general patterns of customer behavior.
+As the project progresses, it integrates more advanced tool, Expectation Maximization algorithms.
 
-This study intends to identify distinct customer groups within sales data, with each group comprising customers who exhibit similar characteristics in terms of purchase frequency and amounts.
-Modeling customer spending as a mixture of gamma distributions helps to identify different spending behaviors.
-The objective is to develop an Expectation Maximization (EM) algorithm which segregates groups based on their spending patterns, considering the individual purchases transactions made by each customer.
+It is crucial that organizations understand their customer’s behaviors and their purchasing patterns.
+By studying the purchasing patterns of its customers, an organization can design and execute effective customized customer engagement strategies, which can enhance customer service, and consequently increase sales.
+Thus, in order to understand their customer’s purchasing behavior, organizations can engage in customer segmentation.
+This analytical approach involves classifying customers into groups exhibiting similar characteristics or purchasing behaviors.
+When organizations analyze factors, such as purchase frequency and spending habits, they can identify distinct segments within their customer base and subsequently tailor their marketing and sales efforts to meet the unique needs of each group.
+This tailored personalized approach can enhance customer satisfaction and also improve multiple end results like customer loyalty, operational efficiency, and profitability.
 
-The exploration begins with applying commonly used alternative models to a publicly available data and ultimately applying the gamma mixture model to this data.
+The project will employ a simple model for classification, logistic regression.
+Since the data doesn't have a labeled categorical variable for classification, a target variable will be created based on the purchase amount and frequency during a fixed period.
+
+The project will also explore the use of mixture models as a potential analytical tool based on the assumption that customer data is inherently complex which can exhibit multiple underlying patterns that are not immediately evident through conventional analysis.
+Thus, the use of finite mixture models is useful in this project.
+These models are adept at analyzing data that may contain multiple, distinct groups hidden within the data.
+Gaussian mixture models and mixtures of t-distributions are examples of the most commonly used finite mixture models in analyzing such data.
+However, models like these that are primarily grounded in symmetric distributions may not always provide an accurate representation of data characterized by asymmetric distributions.
+Therefore, in cases where skewed data need to be analyzed, asymmetric models become useful since they offer a broader applicability across various model-based clustering scenarios [@lee2013model].
+More specifically, the application of a mixture of gamma densities is a rational choice when dealing with observed data that is positive and exhibits right-skewness, suggesting a mixture distribution origin [@young2019finite].
+This project will also explore the application and compare the results of Gaussian mixture model and Gamma mixture model on the retail store’s data.
 
@@
 
-Initially this data was downloaded from Kaggle.com to it's usability for the project.
+Initially this data was downloaded from Kaggle.com to assess it's usability for the project.
 Upon confirming the usability and revisiting the website seeking for further details, it could not be found in Kaggle.com.
@@
 (8)	Country: It represents the name of the country where each customer resides.
-
-The study will also use simulated data to formulate the EM algorithm. Following successful testing, the refined algorithm will be applied to the sales data.
+(9) Sales: It represents product price multiplied by UnitPrice.
 
@@
 
-This study aims to answer the question, "How effective is the Expectation Maximization (EM) algorithm in identifying distinct customer groups when accounting for individual spending by each customer, as demonstrated by initial validation using simulated data?"
-
-[1] Lee, S. X., & McLachlan, G. J. (2013). Model-based clustering and classification with non-normal mixture distributions. Statistical Methods & Applications, 22(4), 427-454.
-
-[2] Young, D.
-S., Chen, X., Hewage, D.
-C., & Nilo-Poyanco, R.
-(2019).
-Finite mixture-of-gamma distributions: estimation, inference, and model-based clustering.
-Advances in Data Analysis and Classification, 13, 1053-1082.
-
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
+This project aims to gain a comprehensive understanding of customer segmentation from different perspectives; predictive (logistic regression), probabilistic clustering (GMM), and skewness-aware clustering (Gamma Mixture Models).
 
@@
 _Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
-
-## Schematic of workflow
-
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
-
-```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-#| echo: FALSE
-#knitr::include_graphics(here("products","assets","antigen-recognition.png"))
-```
-
-## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
 
@@
 
-This study employs sales data, 'superstoredata' and focuses on international sales of the online retail store.
+This project employs sales data , 'superstoredata', of an online retail store based in United Kingdom, to explore purchase patterns of customers outside of United Kingdom.
 
@@
 
-This data has nine variables. The details are explained below:
+The data has 541909 observations and nine variables detailed as below:
 
@@
 (8)	Country: It represents the name of the country where each customer resides.
+(9) Sales: It represents product price multiplied by UnitPrice.
 
-The study will also use simulated data to formulate the EM algorithm. Following successful testing, the refined algorithm will be applied to the sales data.
+One key factor of this data is it accounts sales of a particular item from an Invoice as an observation leading different items sold in the same Invoice to appear as separate observations.
+As such most of the Invoice numbers appear several times in different observations of the data.
+
+As the purpose of the project is to study purchase pattern of customers residing outside of UK, first all of the UK based observations were dropped from the data which brought down the data size to 46431 observations with 9 variables.
+
+The data inspection pointed out to several issues requiring cleaning such as missing values in CustomerID, negative signs in Quantity and Sales and incorrect format of date.
+It also pointed out that InvoiceNo, StockCode, Description and Country were coded as character variables in the data which needed to be changed to factors for better data manipulation.
+
+First, there were 1480 missing values in CustomerID.
+Customer behavior is a crucial part of this analysis which cannot be tracked without the unique customerID.
+The observations without CustomerID do not help in doing meaningful analysis and are dropped.
+Further, CustomerID was coded as a numerical variable in the original data and was converted to factor variable during the cleaning process.
+Country, Description and StockCode were also converted to factors from character variables.
+Transaction date was also converted to the date format.
+
+There were 1372 observations with negative values in quantity and sales due to order cancellation which were dropped from the data.
+
+A detailed scrutiny of the distribution of UnitPrice revealed that all the high value goods with a Unitprice above 25 were distributed between 75 and 100 percentile.
+A closer look at this inter-quartile range disclosed that there were 19 categories of product description which had the Unitprice above 25.
+Some of those categories had descriptions such as CARRIAGE, POSTAGE, Manual, which appeared to be related to logistics.
+1216 observations with such descriptions were dropped from the data since those were less likely to be directly related to customer purchase.
+
+The store also sells merchandises in wholesale quantities.
+A wholesaler potentially purchased in big volumes and in higher frequencies.
+In absence of clear information on wholesalers, a scatter plot of Total Invoice Value by Purchase Frequency was plotted to identify potential wholesalers.
+The scatter plot revealed that most of the customers made purchases less than a frequency of 50 during that period and most of the cumulative purchase values were less than 50,000.
+Assuming that the outliers represented purchases by wholesalers, observations with purchase frequencies above 50 and cumulative Invoice value above 50,000 are dropped from the data.
+This step excluded 9745 observations from the data.
+The cleaned data has 32602 observations with 9 variables.
 
@@
 
-_Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project._
+A meaningful analysis of customer spending pattern require the Invoice Sales Amount, total purchase by Customers, frequency of customer purchases during a period or a combination of those.
+@tbl-resulttable1 is created which provides the descriptive statistics of sales by Invoice value, Customer purchase and frequency of customer purchase.
+It revealed that there were 1479 transactions with 414 different customers during the period of 2010-12-01 to	2011-12-09.
+The store recorded a mean sales of GBP 567 per transaction, an average cumulative sales of GBP 2025 to individual customers and sales frequency of above 3 times per customer.
 
-@tbl-summarytable shows a summary of the data.
+```{r out.width="100%", fig.show='hold', warning=FALSE}
+#| label: tbl-resulttable1
+#| tbl-cap: "Descriptive Statistics"
+#| echo: FALSE
+knitr::include_graphics(here("results","tables", "gt_table.png"))
 
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
+```
+
+@fig-result depicts two density plots for the Cumulative Sales across customers and Invoice Sales and a histogram for Customer purchase frequencies.
+Both density plots are heavily right-skewed.The averages of total sales to a customer and total invoice values appear to be less than 2500 and 1250 sterling pounds respectively supporting the mean values of 2025 and 563 in the descriptive table.
+The average frequency appears to be less than 5 times which is also observed in the descriptive statistics table.
 
 ```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
+#| label: fig-result
+#| fig-cap: "Distribution of Sales and Purchase frequency"
 #| echo: FALSE
-resulttable=readRDS("../../results/summarytable.rds")
-knitr::kable(resulttable)
+knitr::include_graphics(here("results", "figures", "Sales_distribution.png"))
 ```
@@
 
-@fig-result shows a scatterplot figure produced by one of the R scripts.
+#@fig-result1 shows a scatterplot figure produced by one of the R scripts.
 
 ```{r}
-#| label: fig-result
+#| label: fig-result1
 #| fig-cap: "Height and weight stratified by gender."
 #| echo: FALSE
-knitr::include_graphics(here("results", "figures", "Sales_distribution.png"))
+#knitr::include_graphics(here("results", "figures", "Sales_distribution.png"))
 ```
@@
 
-Example @tbl-resulttable2 shows a summary of a linear model fit.
+#Example @tbl-resulttable2 shows a summary of a linear model fit.
 
@@
 #| echo: FALSE
-resulttable2 = readRDS(here("results","tables","summarytable.rds"))
-knitr::kable(resulttable2)
+#resulttable2 = readRDS(here("results","tables","summarytable.rds"))
+#knitr::kable(resulttable2)
 ```

### 2024-03-17 R/eda-code/eda_superstore.qmd
Project Part 3

@@
 library(webshot2) #Capture as an Image the gttable before converting it to pdf
+library(zoo) #To plot seasonality of sales
 ```
@@
 #Path to data using here() package
-data_location <- here::here("data","processed-data","processed_superstore.rds")
+#data_location <- here::here("data","processed-data","processed_superstore.rds")
+data_location <- here::here("data","processed-data","processed_superstore_RFM.rds")
 #load data
@@
 
-It is important to note that this data accounts sales of a particular item at a time as an observation leading different items sold in the same Invoice to appear as separate observations.
-As such most of the Invoice numbers appear several times in different observations of the data.
-A meaningful analysis of customer spending pattern require the Invoice Sales Amount, total purchase by a Customer during a period, frequency of purchase by a customer during a period or a combination of those.
+A meaningful analysis of customer spending behavior require the Invoice Sales Amount, total purchase by a Customer during a period, frequency of purchase by a customer during a period or a combination of those.
 A table is created which provides the descriptive statistics of sales by Invoice value, Customer purchase and frequency of customer purchase.
-This table reveals that there were 1479 transactions to 414 different customers during 2010-12-01 and	2011-12-09.
+This table reveals that there were 1402 transactions to 399 different customers during 2010-12-01 and	2011-12-09.
 
@@
 Sales_by_Invoice <- mydata %>%
-  group_by(InvoiceNo) %>%
-  summarise(TotalSales = sum(Sales, na.rm = TRUE)) %>%
-  ungroup() %>% # Ensure we ungroup before further summarisation
     summarise(
     Count = n(),
-    Particulars = "Sales by Invoice",
-    Mean = round(mean(TotalSales, na.rm = TRUE), 2),
-    SD = round(sd(TotalSales, na.rm = TRUE), 2),
-    Min = round(min(TotalSales, na.rm = TRUE), 2),
-    Max = round(max(TotalSales, na.rm = TRUE), 2)
+    Particulars = "Invoice Value",
+    Mean = round(mean(InvoiceValue, na.rm = TRUE), 2),
+    SD = round(sd(InvoiceValue, na.rm = TRUE), 2),
+    Min = round(min(InvoiceValue, na.rm = TRUE), 2),
+    Max = round(max(InvoiceValue, na.rm = TRUE), 2)
   )
@@
   group_by(CustomerID) %>%
-  summarise(TotalSales = sum(Sales, na.rm = TRUE)) %>%
+  summarise(TotalSales = sum(InvoiceValue, na.rm = TRUE)) %>%
   ungroup() %>%
@@
   group_by(CustomerID) %>%
-  summarise(TotalSales = sum(Sales, na.rm = TRUE)) %>%
+  summarise(TotalSales = sum(InvoiceValue, na.rm = TRUE)) %>%
   arrange(desc(TotalSales))
@@
 # Histogram of Sales by Invoice values
-## First calculate total invoice value per invoice
-Sales_by_Invoice <- mydata %>%
-  group_by(InvoiceNo) %>%
-  summarise(TotalInvoiceValue = sum(Sales, na.rm = TRUE))
 
-##Create histogram
-p2 <- ggplot(Sales_by_Invoice, aes(x = TotalInvoiceValue)) +
+p2 <- ggplot(mydata, aes(x = InvoiceValue)) +
   geom_density(fill = "lightgreen", alpha = 0.5) +
@@
 
-A histogram is also plotted for price range of the products.Most of the products fall below the price of 12.50 sterling pounds.
-
-```{r}
-# Create histogram of Unit Price
-
-p4 <- ggplot(mydata, aes(x = UnitPrice)) +
-  geom_histogram(binwidth = 1, fill = "lightblue", color = "darkblue") + # Adjust binwidth as needed
-  labs(title = "Histogram of Unit Price",
-       x = "Unit Price", 
-       y = "Frequency") +
-  theme_minimal()
-
-p4
-
-figure_file = here("results","figures","Price_range.png")
-ggsave(filename = figure_file, plot=p4)
-
-```
 Next, total purchase by a customer was plotted against purchase frequency.A few outliers were noted.
@@
   summarise(PurchaseFrequency = n_distinct(InvoiceNo),
-            TotalInvoiceValue = sum(Sales, na.rm = TRUE)) %>%
+            TotalInvoiceValue = sum(InvoiceValue, na.rm = TRUE)) %>%
   ungroup()  # Ensure the data is ungrouped for further analysis
@@
 # Scatter plot of Total Invoice Value by Purchase Frequency
-p5 <- ggplot(customer_stats, aes(x = PurchaseFrequency, y = TotalInvoiceValue)) +
+p4 <- ggplot(customer_stats, aes(x = PurchaseFrequency, y = TotalInvoiceValue)) +
   geom_point(alpha = 0.5) +  # Adjust alpha for point transparency, if needed
@@
   theme_minimal()
-plot(p5)
+plot(p4)
 
 figure_file = here("results","figures","Sales_trend.png")
-ggsave(filename = figure_file, plot=p5)
+ggsave(filename = figure_file, plot=p4)
 
@@
 ```{r}
-p8 <- mydata %>% 
+p5 <- mydata %>% 
   group_by(Country) %>% 
-  summarise(TotalSales = sum(Sales)) %>%
+  summarise(TotalSales = sum(InvoiceValue)) %>%
   ggplot(aes(x = reorder(Country, TotalSales), y = TotalSales, fill = Country)) + 
@@
 
-plot(p8)
+plot(p5)
 
 figure_file = here("results","figures","Countrywise_Sales.png")
-ggsave(filename = figure_file, plot=p8)
+ggsave(filename = figure_file, plot=p5)
+```
+As the store sells gift goods, an inspection is required to check the seasonality of sales.
+First, monthly total sales is calculated aggregating Invoice Value by months of the Invoice Date.The aggregated sales is then plotted to visualize seasonal trend which depicted a spike in the sales during the last few months of the year.
+
+```{r}
+#Aggregating Sales by months
+mydata_aggregated <- mydata %>%
+  group_by(InvoiceDate = as.yearmon(InvoiceDate)) %>%
+  summarize(TotalValue = sum(InvoiceValue))
+
+# Formatting Date for plotting purposes
+mydata_aggregated$InvoiceDate <- as.Date(as.yearmon(mydata_aggregated$InvoiceDate))
+
+p6<-ggplot(mydata_aggregated, aes(x = InvoiceDate, y = TotalValue)) +
+  geom_line() + 
+  scale_x_date(date_labels = "%b", date_breaks = "1 month") + # Adjust formatting as needed
+  labs(title = "Monthly Sales", x = "Month", y = "Sales Value") +
+  theme_minimal()
+p6
+
+figure_file = here("results","figures","Monthly_Sales.png")
+ggsave(filename = figure_file, plot=p6)
+
 ```

### 2024-03-17 R/processing-code/processing_superstore-v1.qmd
Project Part 3

@@
 
+The wholesalers are expected to purchase in large quantity.
+Additional inspection is carried out to identify wholesalers by quantity purchased and individual Purchase Amount by plotting a scatter plot.
+
+```{r}
+# Calculate purchase frequency and total invoice value per customer
+customer_stats5 <- clean_data_step4 %>%
+  group_by(InvoiceNo) %>%
+  summarise(InvoiceQty = sum(Quantity, na.rm = TRUE),
+            InvoiceValue = sum(Sales, na.rm = TRUE)) %>%
+  ungroup()  # Ensure the data is ungrouped for further analysis
+
+# Scatter plot of Total Invoice Value by Purchase Frequency
+p3 <- ggplot(customer_stats5, aes(x = InvoiceValue, y = InvoiceQty)) +
+  geom_point(alpha = 0.5) +  # Adjust alpha for point transparency, if needed
+  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add a linear regression line without standard error
+  labs(title = "Total Invoice Value by Purchase Frequency",
+       x = "Invoice Amount",
+       y = "Quantity Purchased") +
+  scale_y_continuous(labels = label_comma()) +  # This line changes the y-axis labels
+
+    theme_minimal()
+plot(p3)
+
+```
+The scatter plot reveals that most of the customers made purchases less than 1000 items and most of the invoice values were less than 2,500.
+Assuming that the sparsely distributed points represent purchases by wholesalers, observations with Quantity above 1000 and individual Invoice value above 2,500 are removed from the data.
+
+This involves first sub-setting customer_stats5 such that it includes only those observations where InvoiceQty <= 1000 and InvoiceValue <=2500 and finally, merging this sub-set with the original dataset (clean_data_step4).
+
+```{r}
+# Filtering the customer_stats5 to meet the conditions
+filtered_customer_stats5 <- customer_stats5 %>%
+  filter(InvoiceQty <= 1000, InvoiceValue <= 2500)
+
+# Excluding purchase quantities above 1000 and individual invoice values above 2500 from the data
+clean_data_step5 <- clean_data_step4 %>%
+  semi_join(filtered_customer_stats5, by='InvoiceNo')
+
+skimr::skim(clean_data_step5)
+```
+This step excluded 4485 observations from the data for a final total observation of 28117. A similar scatter plot is plotted for the new data.
+
+```{r}
+# Calculate purchase quantity and amount per invoice
+customer_stats6 <- clean_data_step5 %>%
+  group_by(InvoiceNo) %>%
+  summarise(InvoiceQty = sum(Quantity, na.rm = TRUE),
+            InvoiceValue = sum(Sales, na.rm = TRUE)) %>%
+  ungroup()  # Ensure the data is ungrouped for further analysis
+
+# Scatter plot of Purchase Quantity by Invoice Value
+p4 <- ggplot(customer_stats6, aes(x = InvoiceValue, y = InvoiceQty)) +
+  geom_point(alpha = 0.5) +  # Adjust alpha for point transparency, if needed
+  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add a linear regression line without standard error
+  labs(title = "Purchase Quantity by Invoice Value",
+       x = "Invoice Amount",
+       y = "Quantity Purchased") +
+  scale_y_continuous(labels = label_comma()) +  # This line changes the y-axis labels
+
+    theme_minimal()
+plot(p4)
+
+```
+
+```{r}
+
+```
+
 Assigning a final name to the subset to make it more relevant to the actual data.
@@
 ```{r}
-processed_superstore <- clean_data_step4
+processed_superstore <- clean_data_step5
 
 str(processed_superstore)
+```
+It is important to note that this data accounts sales of a particular item at a time as an observation leading different items sold in the same Invoice to appear as separate observations.
+As such most of the Invoice numbers appear several times in different observations of the data.
+A meaningful analysis of customer spending pattern can be done by total Invoice Sales.
+So another data is created where each Invoice is a unique observation and can be used for groupoing customers based on Recency, Frequency and Monetary value.
+
+```{r}
+processed_superstore_RFM <- processed_superstore %>%
+  group_by(InvoiceNo) %>%
+  summarise(InvoiceDate = first(InvoiceDate),
+            CustomerID = first(CustomerID),
+            InvoiceValue = sum(Sales, na.rm = TRUE),
+            Country = first(Country)) %>%
+select(InvoiceDate, InvoiceNo, CustomerID, InvoiceValue, Country)
+
+str(processed_superstore_RFM)
 ```
@@
 ```
+
+```{r}
+save_data_location <- here::here("data","processed-data","processed_superstore_RFM.rds")
+saveRDS(processed_superstore_RFM, file = save_data_location)
+
+```

### 2024-03-17 products/manuscript/Manuscript.qmd
Project Part 3

--- 2024-02-09 products/manuscript/Manuscript.qmd
@@
 ---
-title: "Example Manuscript Template for a Data Analysis Project"
+title: "Customer Segmentation"
 subtitle: ""
-author: Andreas Handel
+author: Malika Dhakhwa
 date: "`r Sys.Date()`"
@@
 ## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
+_Provide enough background on your topic that others can understand the why and how of your analysis
 
-Finite mixture models are used to analyze data presumably arising from several separate, yet unobserved, groups.
-Among these models, Gaussian mixture models and mixture of t-distributions are particularly prevalent.
-However, such models which are based on symmetric distribution often do not accurately represent data with asymmetric distribution.
-Skewed data can be more effectively managed using asymmetric distributions, thereby extending their applicability across a broader spectrum of model-based clustering scenarios [1].
-If the observed data is positive, distribution is right-skewed and assumed to come from a mixture distribution, employing mixture of gamma density is a rational choice [2].
+This project aims to explore customer segmentation through the supervised and un-supervised learning perspectives.
+In the initial phase, the analysis applies and investigates the effectiveness of logistic regression to segregate customers based on the general patterns of customer behavior.
+As the study advances, it integrates more advanced tools, such as Expectation Maximization algorithms.
 
-This study intends to identify distinct customer groups within sales data of luxury goods, with each group comprising customers who exhibit similar characteristics in terms of purchase frequency and amounts.
-Modeling customer spending as a mixture of gamma distributions helps to identify different spending behaviors.
-The objective is to develop an Expectation Maximization (EM) algorithm which segregates groups based on their spending patterns, considering the individual purchases transactions made by each customer.
+It is crucial that organizations understand their customer’s purchasing patterns to design and execute effective and customized  engagement strategies.
+This helps businesses enhance customer service, consequently increase sales.
+Based on characteristics like spending and purchase frequency, businesses can segment customers and tailor their marketing and sales efforts to meet the unique needs of each group.
+This personalized approach can enhance customer satisfaction and also improve multiple end results like customer loyalty, operational efficiency, and profitability.
+
+The original data does not have the customers categorized into a group. To use logistic regression, a group variable is created based on the purchase amount and frequency.
 
@@
 
-This study employs simulated data to assess the effectiveness of the EM algorithm. Following successful testing, the refined algorithm will be applied to actual sales data pertaining to luxury goods.
+This study utilizes the 'superstoredata'dataset, focusing on international sales from an online retail store.
+
+Initially this data was downloaded from Kaggle for a preliminary assessment of it's usability for this project.
+However, when revisiting the site for further details, it was no longer available on Kaggle.com.
+A subsequent Google search led to the discovery of the dataset in the [UC Irvine Machine Learning Repository](https://archive.ics.uci.edu/ml/datasets/Online+Retail).
+Per the repository, this is a transnational data set spanning from December 1, 2010, to December 9, 2011 from a UK-based, non-store online retail operation.
+The company specializes in unique all-occasion gifts to a clientele which also includes wholesalers.
+
+This data has nine variables as detailed below:
+
+(1)	InvoiceNo: It is a categorical variable with a 6-digit integral number uniquely assigned to each transaction. If this code starts with the letter ‘c’, it indicates a cancellation. 
+(2)	StockCode: It is a categorical variable with a 5-digit integral number uniquely assigned to each distinct product.
+(3)	Description: This represents Product name and is a categorical variable.
+(4)	Quantity: It represents quantities of each product (item) per transaction and is an integer.
+(5)	InvoiceDate: It represents the day and time each transaction was generated.
+(6)	UnitPrice: It represents product price per unit in sterling.
+(7)	CustomerID: It is a categorical variable with a 5-digit integral number uniquely assigned to each customer.
+(8)	Country: It represents the name of the country where each customer resides.
+(9) Sales: It represents product price multiplied by UnitPrice.
 
@@
 
-This study aims to answer the question, "How effective is the Expectation Maximization (EM) algorithm in identifying distinct customer groups within luxury goods sales data when accounting for individual spending by each customer, as demonstrated by initial validation using simulated data?"
-
-[1] Lee, S. X., & McLachlan, G. J. (2013). Model-based clustering and classification with non-normal mixture distributions. Statistical Methods & Applications, 22(4), 427-454.
-
-[2] Young, D.
-S., Chen, X., Hewage, D.
-C., & Nilo-Poyanco, R.
-(2019).
-Finite mixture-of-gamma distributions: estimation, inference, and model-based clustering.
-Advances in Data Analysis and Classification, 13, 1053-1082.
-
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
-
-{{< pagebreak >}}
+The objective is to understand customer segmentation from logistic regression and advanced analytical tools such as Mixture Models.
 
@@
 ## Schematic of workflow
-
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
-
-```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-#| echo: FALSE
-#knitr::include_graphics(here("products","assets","antigen-recognition.png"))
-```
-
-## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
+_Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
+is an example of some - completely random/unrelated - schematic that was generated with Biorender.
+We store those figures in the assets folder.
 
@@
 All materials, including files that contain code, should be commented well so everyone can follow along._
+
+The primary dataset for this project is the 'superstoredata' dataset, detailing transactions of an online retailer based in United Kingdom.
+A key feature of this data is that it counts sales of a particular item from an Invoice as an observation leading to different items sold in the same Invoice to appear as separate observations.
+Most of the Invoice numbers appear several times in different data observations.
+
+### Initial Data Cleaning
+
+As the project focuses on international customers, initial data cleaning involved eliminating UK transactions downsizing the data size to 46431 observations across nine variables.
+
+The preliminary inspection identified several issues needing attention such as missing values in 'CustomerID', negative 'signs in 'Quantity' and 'Sales' values, and incorrectly formatted dates.
+Variables, 'InvoiceNo', 'StockCode', 'Description' and 'Country', originally coded as characters, were converted to factor types to facilitate better data manipulation.
+
+1480 observations lacked 'CustomerID' and were removed, as customer behavior cannot be tracked without the unique identifier and 'CustomerID' was converted to factor type from numeric type.
+'Country', 'Description' and 'StockCode' were converted to factors from character variables and transaction date to the date format.
+
+There were 1372 observations with negative values in quantity and sales due to order cancellation which were excluded from the data.
+
+An examination of 'UnitPrice' distribution revealed that all the high value goods with a Unitprice above 25 were distributed between 75 and 100 percentile.
+A closer inspection at this inter-quartile range disclosed that there were 19 product description categories with the Unitprice above 25.
+Some of those categories had descriptions such as CARRIAGE, POSTAGE, Manual, which appeared to be related to logistics.
+1216 observations with such descriptions were dropped from the data since those were less likely to be directly related to customer purchase.
+
+### Refining the Dataset
+The store also sells merchandises in wholesale quantities.
+A wholesaler potentially had larger purchases and in higher frequencies.
+In absence of clear information on wholesalers, a scatter plot of Cumulative Invoice Value of customers by Purchase Frequency was plotted to identify potential wholesalers.
+The scatter plot revealed that most of the customers made purchases less than a frequency of 50 and that most cumulative purchase values were less than 50,000.
+Assuming that the outliers represented purchases by wholesalers, observations with purchase frequencies above 50 and cumulative Invoice values above 50,000 were dropped from the data.
+This step excluded 9745 observations from the data.
+The cleaned data has 32602 observations with 9 variables.
+Next, the quantities purchased in each invoice were plotted against the individual invoice value.
+The scatter plot revealed that the observations were concentrated within the purchase quantity below 1000 items and invoice amounts of 2,500.
+Assuming that the sparsely distributed points represent purchases by wholesalers, observations with Quantity above 1000 and individual Invoice value above 2,500 were further removed from the data.This step excluded 4485 observations for a final total observation of 28117.
+
+The study of customer spending pattern requires the knowledge of each individual transaction amount.
+The subsequent cleaning process reorganized the dataset by 'InvoiceNo', consolidating sales data under each invoice.
+This final data is saved as processed_superstore_RFM in the data > processed-data.
+
+### RFM Classification
+In absence of a labeled categorical variable for classification, the study segmented customers into two distinct groups using a recency, frequency and monetary (RFM) model.
+RFM model is a commonly used behavior-based model in analysing customer behavior [@yeh2009knowledge].
+Recency is measured in days or months and represent the interval between the most recent transaction time and the time of analysis, and a lower number interval is preferred.
+Frequency is the number of purchases made in a certain period, and monetary is the total amount the customer spent during that time period.[@wei2010review]
+
+In the RFM classification process, the Recency, Frequency and Monetary values were calculated for each customer, ensuring that each customer was represented only once in the modified dataset.Frequency in this data is the total no.
+of Invoices issued to the customer, and Monetary is the total purchases made by the customer during the stipulated period.
+Recency was calculated from the next day of the transaction period.
+Each customer was assigned a value ranging from 1 to 4 for the Recency, Frequency and Monetary attributes, reflecting their quartile position within each metric.
+Subsequently, these metrics were combined to formulate a three-digit score for every custoemr, positioning the Monetary value in the hundreds place, Frequency in the tens place, and Recency in the units place.
+All the customers with a score above 344 were assigned as High-Value Customers.
+
+Following RFM segmentation, logistic regression was applied to predict customer group membership.
+This method utilized the segments derived from RFM as the dependent variable, with the Recency, Frequency and Monetary attritubtes serving as independent variables.
+The objective is to develop a model for predictive analysis based on those segments.
 
@@
 
-_Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project._
+@tbl-resulttable1 depicts the descriptive statistics of various aspects of customers purchase behavior.
+There were 1402 transactions with 399 customers outside UK during the period from December 1, 2010 to	December 9, 2011.
+The store recorded a mean sales of GBP 436 per transaction, an average cumulative sales of GBP 1532 to individual customers and sales frequency of above 3 times per customer.
 
-@tbl-summarytable shows a summary of the data.
+```{r out.width="100%", fig.show='hold', warning=FALSE}
+#| label: tbl-resulttable1
+#| tbl-cap: "Descriptive Statistics"
+#| echo: FALSE
+knitr::include_graphics(here("results","tables", "gt_table.png"))
 
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
+```
+
+@fig-result1 depicts the distribution of customer segmentation and suggests that approximately 25 percentage of the customers are grouped as Valued Customers.
 
 ```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
+#| label: fig-result1
+#| fig-cap: "Distribution of Sales and Purchase frequency"
 #| echo: FALSE
-resulttable=readRDS("../../results/summarytable.rds")
-knitr::kable(resulttable)
+knitr::include_graphics(here("results", "figures", "Customer_Segment.png"))
+```
+
+@fig-result2 depicts two density plots for the Cumulative Sales and Invoice Values across customers and a histogram for Customer purchase frequencies.
+Both density plots are heavily right-skewed.T0he averages of total sales to a customer and total invoice values appear to be less than 2500 and 500 sterling pounds, respectively, supporting the mean values of 1532 and 436 in the descriptive table.
+The average frequency appears to be less than 5 times, which is also observed in the descriptive statistics table.
+
+```{r}
+#| label: fig-result2
+#| fig-cap: "Distribution of Sales and Purchase frequency"
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "Sales_distribution.png"))
+```
+@fig-result3 depicts a density plot of number of days since the last transaction made by a customer.
+The distribution is right-skewed.
+This plot indicates that most customers made a purchase transaction within the last 30 days of the reference date.It also reveals that few customers had not make a transaction for a long time.
+
+```{r}
+#| label: fig-result3
+#| fig-cap: "Distribution of Recency of Purchase"
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "Recency.png"))
+```
+
+@fig-result4 depicts a trend of monthly international sales.
+A spike is noticed in November, presumably due to Christmas in the following month, which is also supported by distribution of the Recency plot.
+```{r}
+#| label: fig-result4
+#| fig-cap: "Monthly Sales trend"
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "Monthly_Sales.png"))
+```
+@fig-result5 depicts country-wise trend of international sales by sales amount. Top customers were in Germany, France, Spain and Belgium.  
+```{r}
+#| label: fig-result5
+#| fig-cap: "Countrywise Sales trend"
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "Countrywise_Sales.png"))
 ```
@@
 
-@fig-result shows a scatterplot figure produced by one of the R scripts.
+Penalized logistic regression of customer segment was performed on Recency, Frequency and Monetary variables using 'tidymodels' package and 4-fold corss-validation techniques.
+
+@fig-result6 depicts the area under ROC curve by the range of penalty values.
+It provides visualization of validation set metrics.
+The curve has extreme high values, all above 0.9995.
+A sharp drop is noticed before the penalty value of 0.0010, suggesting the sensitivity of the performance of the model at this penalty value.
+This is followed by a sharp rise after the penalty value of 0.01.
+These features of model performance are indicative of model overfitting to the training data and the data being easily separable.
 
 ```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
+#| label: fig-result6
+#| fig-cap: "Visualization of validation set metrics"
 #| echo: FALSE
-knitr::include_graphics(here("results","height-weight-stratified.png"))
+knitr::include_graphics(here("results", "figures", "ROC(AUC)_Penalty.png"))
+```
+@fig-result7 depicts validation set ROC curve plotting true positive rate (sensitivity) against the false positive rate (1-specificity) at various threshold settings.
+The curve moves along 1-specificity and rises only from the end.This suggests that the model has no  ability to distinguish between the two customer categories.
+This suggests that the logistic regression model is not appropriate to the dataset.
+It pointed out the need to revisit the choice of model.
+
+```{r}
+#| label: fig-result7
+#| fig-cap: "Visualization of Validation set ROC curve"
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "ROC(AUC)_Validation.png"))
 ```
@@
 #| echo: FALSE
-resulttable2 = readRDS(here("results","tables-files","resulttable2.rds"))
-knitr::kable(resulttable2)
+#resulttable2 = readRDS(here("results","tables","summarytable.rds"))
+#knitr::kable(resulttable2)
 ```

### 2024-03-17 R/analysis-code/statistical-analysis_superstoreRFM.R

@@ -1,0 +1,282 @@
+###############################
+# analysis script
+#
+#this script loads the processed, cleaned data, does a simple analysis
+#and saves the results to the results folder
+
+#loading packages needed. 
+library(here) #for data loading/saving
+library(tidymodels) # for the parsnip package, along with the rest of tidymodels
+library(ggplot2) #for plotting
+library(dplyr) #for data wrangling
+library(broom) #for cleaning up output from lm()
+library(lubridate)
+library(glmnet) #for fitting GLM via penalized maximum likelihood
+library(vip) # for visualizing the variable importance scores for the top 20 features
+#library(ranger) #for random forest. Unable to sync this in renv
+
+#path to data using the here() package
+data_location <- here::here("data","processed-data","processed_superstore_RFM.rds")
+
+#load data. 
+mydata <- readRDS(data_location)
+
+#Checking data
+skimr::skim(mydata)
+head(mydata)
+
+#This study aims to use logistic regression to segment customers based on the their purchasing behavior reflected by 
+#the monetary purchase amount and loyalty of the customer   
+#In absence of a readily available classification variable in the data, one will be created based on three factors:
+# Recency, Frequency and Monetary. 
+#Recency is the number of days since the last purchase by a customer. 
+#Frequency is the no. of purchases by a customer over a given period.
+#Monetary is the total amount of purchase a customer has made over a period. 
+
+# For determining recency of a transaction, a reference date past the study period is required. The data covers transactions till 2011-12-09.
+# The next day i.e. 2011-12-10 is considered as the reference date to define recency.
+
+ref_date <- as.Date("2011-12-10")
+
+# Next, a new data frame is created which includes the segments, Recency, Frequency and Monetary
+rfm_data <- mydata %>%
+  group_by(CustomerID) %>%
+  summarise(
+    Recent_Purchase_Date=last(InvoiceDate),
+    Recency = as.numeric(ref_date - as.Date(max(InvoiceDate))),
+    Frequency = n_distinct(InvoiceNo),
+    Monetary = sum(InvoiceValue)
+    
+  ) %>%
+  ungroup()
+
+skimr::skim(rfm_data)
+
+#Density Plot of Recency
+
+p1 <- ggplot(rfm_data, aes(x = Recency)) +
[…]

### 2024-03-17 R/analysis-code/superstore.R

@@ -1,181 +1,0 @@
-library(readxl) #for loading Excel files
-library(dplyr) #for data processing/cleaning
-library(tidyr) #for data processing/cleaning
-library(skimr) #for nice visualization of data
-library(here)  #to set paths
-library(ggplot2)
-
-#path to data
-data_location <- here::here("data","raw-data","superstoredata.csv")
-raw_data <- read.csv(data_location)
-#Check data
-dplyr::glimpse(raw_data)
-head(raw_data)
-tail(raw_data)
-summary(raw_data)
-str(raw_data)
-skimr::skim(raw_data) #gives missing nos., completion rate, min, max, empty, n_unique, whitespace,mean, standard deviaiont and percentile 
-
-#By inspecting the data, we find some problems that need to be addressed:
-
-#First, there are 135080 missing values in CustomerID. The analysis looks into customer behaviors for which the unique customer ID is an essential part of the study.
-#Hence, the observations without CustomerID do not provide meaningful analysis. Such observations are dropped for analyses. Further 
-#CustomerID is loaded as a numerical variable in the data and it needs to be converted to categorical/factor variable
-#Further Country is also converted to a factor.
-
-
-# Subset the data by excluding rows where CustomerID is NA
-  clean_data_step1 <- raw_data %>%
-  filter(!is.na(CustomerID)) %>%
-  mutate(CustomerID = as.factor(CustomerID),
-         Country = as.factor(Country),
-         Description = as.factor(Description),
-         StockCode = as.factor(StockCode))
-
-# Check the dimensions of the cleaned dataset
-dim(clean_data_step1)
-
-#There are negative values in quantity, unit price and sales due to order cancellation. Let's inspect how many of such data are there
-# Filter observations where Quantity, UnitPrice, and Sales are negative
-negative_sales <- clean_data_step1 %>%
-  filter (Quantity < 0 & Sales < 0)
-
-# View the filtered observations
-count(negative_sales)
-#There are 8905 observations where either Quantity and Sales are negative.
-
-#I check to see if there is a corresponding sales (+ve entry) for these returned items. 
-
-positive_sales <- clean_data_step1 %>%
-  filter(Quantity > 0 & Sales > 0)
-count(positive_sales)
-
-#Attempt to match each negative Sale/Quantity with a corresponding positive sale by CustomerID and StockCode
-matched_sales <- negative_sales %>%
-  rowwise() %>%
[…]

### 2024-03-27 R/eda-code/eda_superstore.qmd
Project Part 3 (Revision 1)
Cleaned the repository, some corrections in all files, README files updated

@@
 #Path to data using here() package
-#data_location <- here::here("data","processed-data","processed_superstore.rds")
 data_location <- here::here("data","processed-data","processed_superstore_RFM.rds")
@@
     summarise(
+    Particulars = "Invoice Value",
     Count = n(),
-    Particulars = "Invoice Value",
         Mean = round(mean(InvoiceValue, na.rm = TRUE), 2),
@@
   summarise(
+    Particulars = "Cumulative Sales to Individual Customers",
     Count = n(),
-    Particulars = "Cumulative Sales to Individual Customers",
     Mean = round(mean(TotalSales, na.rm = TRUE), 2),
@@
 # Combine the statistics into a single data frame
-combined_stats <- bind_rows(Sales_by_Invoice, TotalSales_customer, sales_frequency_per_customer) %>%
-  select(Particulars, everything())  # Ensure Particulars column is at the front
+combined_stats <- bind_rows(Sales_by_Invoice, TotalSales_customer, sales_frequency_per_customer)
 
-# Create and format the gt table
-gt_table <- gt(combined_stats) %>%
-  tab_header(
-    title = "Descriptive Statistics of Sales") %>%
-  cols_label(
-    Particulars = "Particulars",
-    Count = "Count",
-    Mean = "Mean",
-    SD = "St.Dev.",
-    Min = "Min",
-    Max = "Max"
-  ) %>%
-  cols_align(
-    align = "center",
-    columns = c("Count", "Mean", "SD", "Min", "Max")
-  )
+combined_stats
 
-# Display the gt table
-print(gt_table)
+#Saving this statistics in the results folder.
 
-gt_table %>%
-  gtsave(here("results","tables", "gt_table.png"))
+summary_combined = here("results","tables", "combined_stats.rds")
+saveRDS(combined_stats, file = summary_combined)
 
 ```
+
 # Data exploration through figures
@@
 ```
-
-# Notes
-
-For your own explorations, tables and figures can be "quick and dirty".
-As long as you can see what's going on, there is no need to polish them.
-That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.).
-Those should look as nice, polished and easy to understand as possible.

### 2024-03-27 R/processing-code/processing_superstore-v1.qmd
Project Part 3 (Revision 1)
Cleaned the repository, some corrections in all files, README files updated

@@
 
-The purpose of this project is to study purchase pattern of customers residing outside of UK.The data set data_without_uk excludes all sales in the UK.
+The purpose of this project is to group customers based on their purchase pattern.
+It uses a transnational data set 'superstoredata' spanning from December 1, 2010, to December 9, 2011 from a UK-based, non-store online retail operation.
+The company specializes in unique all-occasion gifts to a clientele which also includes wholesalers.
+
+The project focuses on customers residing outside of UK.In the first step of cleaning, all sales in the UK was excluded.
+The data set data_without_uk excludes all sales in the UK.
 It has 46431 observations with 9 variables.
@@
 
-See here for some suggestions on how to store your processed data:
-http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata
-
-```{r}
-save_data_location <- here::here("data","processed-data","processed_superstore.rds")
-saveRDS(processed_superstore, file = save_data_location)
-
-```
-
 ```{r}

### 2024-03-27 products/manuscript/Manuscript.qmd
Project Part 3 (Revision 1)
Cleaned the repository, some corrections in all files, README files updated

@@
 
-This project aims to explore customer segmentation through the supervised and un-supervised learning perspectives.
-In the initial phase, the analysis applies and investigates the effectiveness of logistic regression to segregate customers based on the general patterns of customer behavior.
-As the study advances, it integrates more advanced tools, such as Expectation Maximization algorithms.
+This project aims to leverage statistical modeling techniques to predict customer segmentation based on their purchasing behavior pattern, using both supervised and unsupervised learning approaches.
+In the initial phase, the analysis employs logistic regression to segregate customers based on the general patterns of purchasing behavior.
+As the study advances, it integrates Machine Learning Tools into the analysis.
 
-It is crucial that organizations understand their customer’s purchasing patterns to design and execute effective and customized  engagement strategies.
-This helps businesses enhance customer service, consequently increase sales.
-Based on characteristics like spending and purchase frequency, businesses can segment customers and tailor their marketing and sales efforts to meet the unique needs of each group.
-This personalized approach can enhance customer satisfaction and also improve multiple end results like customer loyalty, operational efficiency, and profitability.
+It is crucial that organizations understand their customer’s purchasing patterns to devise and implement effective, personalized engagement strategies.
+By analyzing behaviors such as spending habits and purchase frequency, businesses can segment their customers and tailor marketing and sales efforts to cater to the unique needs of each segment.
+This personalized approach can not only boost customer satisfaction but also improve multiple end results like customer loyalty, operational efficiency, and profitability.
 
-The original data does not have the customers categorized into a group. To use logistic regression, a group variable is created based on the purchase amount and frequency.
+Given the absence of pre-categorized customer groups in the original data, a grouping variable is engineered based on purchase amount and frequency to facilitate the statistical analysis.
 
@@
 
-This study utilizes the 'superstoredata'dataset, focusing on international sales from an online retail store.
-
-Initially this data was downloaded from Kaggle for a preliminary assessment of it's usability for this project.
+This study utilizes the 'superstoredata' dataset which is a transactional dataset of a UK-based, non-store online retail operation spanning from December 1, 2010, to December 9, 2011.
+The store specializes in unique all-occasion gifts to a clientele which includes UK and non-UK based retail and wholesale customers.
+This study focuses on international sales of the store.
+The dataset features nine variables, detailing transaction numbers, product codes, product description, quantities sold, transaction dates, unit prices, customer IDs, country of residence of the customer, and sales amounts.
+Initially this data was downloaded from Kaggle for a preliminary assessment of its usability for this project.
 However, when revisiting the site for further details, it was no longer available on Kaggle.com.
 A subsequent Google search led to the discovery of the dataset in the [UC Irvine Machine Learning Repository](https://archive.ics.uci.edu/ml/datasets/Online+Retail).
-Per the repository, this is a transnational data set spanning from December 1, 2010, to December 9, 2011 from a UK-based, non-store online retail operation.
-The company specializes in unique all-occasion gifts to a clientele which also includes wholesalers.
-
-This data has nine variables as detailed below:
-
-(1)	InvoiceNo: It is a categorical variable with a 6-digit integral number uniquely assigned to each transaction. If this code starts with the letter ‘c’, it indicates a cancellation. 
-(2)	StockCode: It is a categorical variable with a 5-digit integral number uniquely assigned to each distinct product.
-(3)	Description: This represents Product name and is a categorical variable.
-(4)	Quantity: It represents quantities of each product (item) per transaction and is an integer.
-(5)	InvoiceDate: It represents the day and time each transaction was generated.
-(6)	UnitPrice: It represents product price per unit in sterling.
-(7)	CustomerID: It is a categorical variable with a 5-digit integral number uniquely assigned to each customer.
-(8)	Country: It represents the name of the country where each customer resides.
-(9) Sales: It represents product price multiplied by UnitPrice.
 
@@
 
-The objective is to understand customer segmentation from logistic regression and advanced analytical tools such as Mixture Models.
+What statistical modeling techniques are most effective in replicating and predicting customer segmentation,based on their purchase behavior pattern, as indicated by purchase amount, purchase frequency and purchase recency?
 
@@
 _Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
-
-## Schematic of workflow
-_Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the assets folder.
 
@@
 
-1480 observations lacked 'CustomerID' and were removed, as customer behavior cannot be tracked without the unique identifier and 'CustomerID' was converted to factor type from numeric type.
+1480 observations lacked 'CustomerID' and were removed, as customer purchase pattern cannot be tracked without the unique identifier and 'CustomerID' was converted to factor type from numeric type.
 'Country', 'Description' and 'StockCode' were converted to factors from character variables and transaction date to the date format.
@@
 ### Refining the Dataset
-The store also sells merchandises in wholesale quantities.
+The store also sold merchandises in wholesale quantities.
 A wholesaler potentially had larger purchases and in higher frequencies.
@@
 This step excluded 9745 observations from the data.
-The cleaned data has 32602 observations with 9 variables.
-Next, the quantities purchased in each invoice were plotted against the individual invoice value.
+The cleaned data at this point had 32602 observations with 9 variables.
+Next, the quantities purchased in each invoice were plotted against the individual invoice values.
 The scatter plot revealed that the observations were concentrated within the purchase quantity below 1000 items and invoice amounts of 2,500.
@@
 ### RFM Classification
-In absence of a labeled categorical variable for classification, the study segmented customers into two distinct groups using a recency, frequency and monetary (RFM) model.
+In absence of a pre-existing customer groups, the study segmented customers into two distinct groups using a recency, frequency and monetary (RFM) model.
 RFM model is a commonly used behavior-based model in analysing customer behavior [@yeh2009knowledge].
@@
 
-In the RFM classification process, the Recency, Frequency and Monetary values were calculated for each customer, ensuring that each customer was represented only once in the modified dataset.Frequency in this data is the total no.
+In the RFM classification process, the Recency, Frequency and Monetary values are calculated for each customer, ensuring that each customer was represented only once in the modified dataset.Frequency in this data is the total no.
 of Invoices issued to the customer, and Monetary is the total purchases made by the customer during the stipulated period.
@@
 
-Following RFM segmentation, logistic regression was applied to predict customer group membership.
-This method utilized the segments derived from RFM as the dependent variable, with the Recency, Frequency and Monetary attritubtes serving as independent variables.
+Following RFM segmentation, logistic regression is applied to predict customer group membership.
+This method utilized the segments derived from RFM as the dependent variable, with the Recency, Frequency and Monetary attributes serving as independent variables.
 The objective is to develop a model for predictive analysis based on those segments.
@@
 #| echo: FALSE
-knitr::include_graphics(here("results","tables", "gt_table.png"))
+resulttable1 = readRDS(here("results", "tables", "combined_stats.rds"))
+knitr::kable(resulttable1)
 
@@
 @fig-result2 depicts two density plots for the Cumulative Sales and Invoice Values across customers and a histogram for Customer purchase frequencies.
-Both density plots are heavily right-skewed.T0he averages of total sales to a customer and total invoice values appear to be less than 2500 and 500 sterling pounds, respectively, supporting the mean values of 1532 and 436 in the descriptive table.
+Both density plots are heavily right-skewed.The averages of total sales to a customer and total invoice values appear to be less than 2500 and 500 sterling pounds, respectively, supporting the mean values of 1532 and 436 in the descriptive table.
 The average frequency appears to be less than 5 times, which is also observed in the descriptive statistics table.
@@
 This is followed by a sharp rise after the penalty value of 0.01.
-These features of model performance are indicative of model overfitting to the training data and the data being easily separable.
+These features of model performance are indicative of the training data and the data being easily separable.
 

### 2024-03-27 R/eda-code/edacode_superstore.R

@@ -1,162 +1,0 @@
-## ---- packages --------
-#load needed packages. make sure they are installed.
-library(here) #for data loading/saving
-library(dplyr)
-library(skimr)
-library(ggplot2)
-
-## ---- loaddata --------
-#Path to data. Note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","processed_superstore.rds")
-#load data
-mydata <- readRDS(data_location)
-
-## ---- table1 --------
-summary_df = skimr::skim(mydata)
-print(summary_df)
-# save to file
-summarytable_file = here("results", "summarytable.rds")
-saveRDS(summary_df, file = summarytable_file)
-
-## ---- Invoice Sales --------
-
-# Histogram of Invoice values
-## First calculate total invoice value per invoice
-invoice_values <- mydata %>%
-  group_by(InvoiceNo) %>%
-  summarise(TotalInvoiceValue = sum(Sales, na.rm = TRUE))
-
-##Create histogram
-p1 <- ggplot(invoice_values, aes(x = TotalInvoiceValue)) +
-  geom_histogram(binwidth = 50, fill = "lightblue", color = "darkblue") + # Adjust binwidth as needed
-  labs(title = "Histogram of Total Invoice Values",
-       x = "Total Invoice Value", 
-       y = "Frequency") +
-  theme_minimal()
-plot(p1)
-
-figure_file = here("results","figures","Sales_distribution.png")
-ggsave(filename = figure_file, plot=p1) 
-
-# Histogram of Invoice Quantity
-## First calculate total invoice quantity per invoice
-invoice_quantity <- mydata %>%
-  group_by(InvoiceNo) %>%
-  summarise(TotalInvoiceQuantity = sum(Quantity, na.rm = TRUE))
-
-p2 <- ggplot(invoice_quantity, aes(x = TotalInvoiceQuantity)) +
-  geom_histogram(binwidth = 50, fill = "lightblue", color = "darkblue") + # Adjust binwidth as needed
-  labs(title = "Histogram of Total Invoice Quantity",
-       x = "Total Invoice Quantity", 
-       y = "Frequency") +
-  theme_minimal()
-plot(p2)
-
-figure_file = here("results","figures","Quantity_distribution.png")
-ggsave(filename = figure_file, plot=p2)
-
-## ---- UnitPrice-Price range of products --------
-# Create histogram of Unit Price
-
[…]

### 2024-04-06 R/analysis-code/statistical-analysis_superstoreRFM.qmd
Project Part 4

---
title: "Analysis Script"
---
## Full Analysis

This script loads the processed, cleaned data, does a simple analysis and saves the results to the results folder

First, required packages are installed and loaded.

```{r}
#| message: false
#| warning: false

library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(ggplot2) # for plots
library(tidymodels)
library(glmnet) ##for fitting GLM via penalized maximum likelihood
library(broom) #for cleaning up output from lm()
library(lubridate)
library(forcats) # for clubbing the factor variables
library(GGally)#for correlation plots
library(ranger) #for random forest.
library(dials) #for tuning parameters
library(patchwork)# For illustrating plots side by side
library(vip)#variable importance plots
```

Setting a random seed during analysis step

```{r}
rngseed = 1234
```

Path to data using here() function

```{r}
# path to data using here function
data_location <- here::here("data","processed-data","processed_superstore_RFM.rds")

#load data. 
mydata <- readRDS(data_location)

```

Checking the data

```{r}

skimr::skim(mydata)
```

This study aims to use logistic regression to segment customers based on their purchasing behavior reflected by the monetary purchase amount and loyalty of the customer. In absence of a readily available classification variable in the data, one will be created based on three factors: Recency, Frequency and Monetary. 
Recency is the number of days since the last purchase by a customer. 
Frequency is the no. of purchases by a customer over a given period.
Monetary is the total amount of purchase a customer has made over a period.

For determining recency of a transaction, a reference date past the study period is required. The data covers transactions till 2011-12-09. The next day i.e. 2011-12-10 is considered as the reference date to measure recency.
```{r}
ref_date <- as.Date("2011-12-10")
```

Next, a new data frame is created which includes the segments, Recency, Frequency and Monetary

```{r}
rfm_data <- mydata %>%
  group_by(CustomerID) %>%
  summarise(
    Recent_Purchase_Date=last(InvoiceDate),
    Recency = as.numeric(ref_date - as.Date(max(InvoiceDate))),
    Frequency = n_distinct(InvoiceNo),
    Monetary = sum(InvoiceValue)

  ) %>%
  ungroup()

skimr::skim(rfm_data)
```

### Pairwise correlations

Next, I created a pairwise correlation plot for the continuous variables to check if any two variables are highly correlated.

```{r}
#| message: false
#| warning: false

#Using ggpairs to create the correlation plot
cor_pairs <- rfm_data[, c("Monetary", "Frequency", "Recency")]

p_correlations<-ggpairs(cor_pairs,
        upper = list(continu = wrap("cor", size = 5)),#Display correlation coefficients
        lower = list(continuous = "smooth") #Display scatter plots with smooth lines
        )

p_correlations

figure_file_correlations = here("results","figures","correlations.png")
ggsave(filename = figure_file_correlations, plot=p_correlations)

```
The correlation between Frequency and Monetary is observed to be 0.787, between Recency and Monetary is -0.334 and between Recency and Frequency is -0.376. The correlations are not excessiveindicating that collinearity is unlikely to be and issue for the analysis.

### Feature Engineering

Each segment is additionally divided into 4 categories based on which quartile of Recency, Frequency and Monetary value the customer belongs to. For example a customer belonging to 3rd quartile of Recency, 1st quartile of Frequency and 4th quartile of Monetary value, will get a value of 3 for Recency, 1 for Frequency and 4 for Monetary value columns.

```{r}
# Creating quartiles
rfm_data <- rfm_data %>%
  mutate(
    R_Quartile = ntile(Recency, 4),
    F_Quartile = ntile(Frequency, 4),
    M_Quartile = ntile(Monetary, 4)
  )
```

The value of 4 for F_Quartile and M_Quartile represents the preferred customer outcomes which are higher frequency and values of purchases respectively. Contrarily, the value of 4 for R-Quartile indicates that the customer had not made a purchase in long time which is not a favorable outcome. To create a uniform direction of preferability for all three metric values, the original value of R_Quartile is subtracted by 5. With this adjustment, now the value 4 for R_Quartile indicates that the customer has recently purchased from the store.

Adjusting the quartile ranking for Recency metric

```{r}
rfm_data$R_Quartile_adj <- 5 - rfm_data$R_Quartile
#Checking the data
skimr::skim(rfm_data)
```

The customers are ranked in the order of their purchase amount, frequency of purchase and recent purchases, a combined score (RFM_Score) is derived based on which quartile the customer falls in for each of these metrics. This score will be used to label valued customers.

```{r}
rfm_data <- rfm_data %>%
  mutate(RFM_Score = M_Quartile*100 + F_Quartile * 10 + R_Quartile_adj)%>%
  arrange(desc(RFM_Score))

```

Checking unique RFM_score in the data
```{r}
unique_score <- unique(rfm_data$RFM_Score)
print(unique_score)
length(unique_score)
```
Counting the no. of observations in unique RFM_scores

```{r}
score_counts <- table(rfm_data$RFM_Score)
score_counts

summary(rfm_data$RFM_Score)
```

There were 55 unique RFM scores identified.The top 25% percentage of customers have a RFM_score above 344. Setting 344 as a threshold for classifying customers as valued customers.

```{r}
rfm_data$ValuedCust <- as.factor(ifelse(rfm_data$RFM_Score > 344, 1, 0))
glimpse(rfm_data)
```

Creating a box plot to visualize the segments.
```{r}
p2 <- ggplot(rfm_data, aes(x = ValuedCust)) +
  geom_bar(fill='light blue') +
  labs(title = "Customer Segment", x="Valued Customers",y = "Count")+
  theme_minimal()
plot(p2)

figure_file = here("results","figures","Customer_Segment.png")
ggsave(filename = figure_file, plot=p2)
```

```{r}
#Creating data set for analysis
final_data <- rfm_data %>% select(Monetary, Frequency, Recency, ValuedCust)
```

## Model Building
The next step is the analysis phase, which involves fitting various models with all predictors.
1. Logistic Regression
2. Logistic Regression with cross-validation
3. Random Forest (RF)

###Logistic Regression
First I fitted logistic regression without applying cross-validation. I splitted 75% of the data to training set and 25% to test set. The following codes are an attempt to build a model using the training set.  
```{r}
#Splitting 75% of the data to training set and 25% to test set
data_split <- initial_split(final_data, prop = 0.75)
train_data <- training(data_split)
test_data <- testing(data_split)

# Defining the logistic regression model specification
logistic_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

# Fitting the model to the training data 
logistic_fit_all <- logistic_spec %>%
  fit(ValuedCust ~ Monetary + Frequency + Recency , data = train_data)
```
R returned a warning that the algorithm did not converge and that the fitted probabilities numerically 0 or 1 occurred.
I computed the accuracy and AUC

```{r}
# Computing the accuracy and AUC of the model
model_acc <- logistic_fit_all %>% 
  predict(train_data) %>% 
  bind_cols(train_data) %>% 
  metrics(truth = ValuedCust, estimate = .pred_class) %>% 
  filter(.metric == "accuracy") 
model_auc <-  logistic_fit_all %>%
  predict(train_data, type = "prob") %>%
  bind_cols(train_data) %>%
  roc_auc(truth = ValuedCust, .pred_1)
# Combine the statistics into a single data frame
logistic_combined_stats <- bind_rows(model_acc, model_auc)
logistic_combined_stats

#Saving this statistics in the results folder.

logistic_combinedstat = here("results","tables", "logit_combinedstats.rds")
saveRDS(combined_stats, file = logistic_combinedstat)

```
The model has an accuracy of 1 however, roc_auc of 0.

##Logistic Regression with cross-validation

In the following steps I used penalized logistic regression with the help of tidymodels package (https://www.tidymodels.org/start/case-study/)

```{r}
#Checking proportion of Valued Customers
final_data %>% 
  count(ValuedCust) %>%
  mutate(prop = n/sum(n))
```

```{r}
set.seed(rngseed) #For reproducibility

#using stratified random sampling as 'Valued Customer' is imbalanced
splits <- initial_split(final_data, strata = ValuedCust)

final_data_train <- training(splits)
final_data_test <- testing(splits)

#training set proportions by 'ValuedCust'

final_data_train %>%
  count(ValuedCust)%>%
  mutate(prop = n/sum(n))

#test set proportions by 'ValuedCust'
final_data_test %>%
  count(ValuedCust) %>%
  mutate(prop=n/sum(n))
```

```{r}
#Defining 10-fold cross-validation on the training data, using stratification

set.seed(rngseed)
cv_folds<-vfold_cv(final_data_train, v=10, strata = ValuedCust)

cv_folds

```

Using glmnet engine to specify a penalized logistic regression model. tune()is model hyperparameter that will be used to tune to find the best value for making predictions with the data. Setting mixture=1 potentially remove irrelevant predictors and choose a simple model

```{r}
lr_mod <- 
  logistic_reg(penalty = tune(), mixture = 1)%>%
  set_engine("glmnet")
```

Creating recipe. It helps preprocess the data before training the model.It is built as a series of pre-processing steps.

```{r}
lr_recipe <- 
  recipe(ValuedCust ~ ., data = final_data_train)%>%
  step_normalize(all_date_predictors()) #centers and scales numeric variable
```

Creating the workflow: bundling the model and recipe into a single workflow(). This makes management of the R objects easier.

```{r}
lr_workflow <- 
  workflow() %>%
  add_model(lr_mod) %>%
  add_recipe(lr_recipe)

# Define a penalty parameter object
penalty_param <- penalty(range = c(-5, 2))

# Create a grid of potential penalty values (log scale)
lr_reg_grid <- grid_regular(penalty_param, levels = 50)

#lambda_values <- 10^seq(-5, 2, length.out = 50)
#lr_reg_grid <- grid_regular(penalty = lambda_values, levels = 50)

#Creating the grid for tuning using a one-column tibble with 30 candidate values:
#lr_reg_grid <- penalty(range = c(-5, 2)) %>% 
 # grid_regular(levels = 50)

lr_reg_grid %>% top_n(-5) #lowest penalty values
lr_reg_grid %>% top_n(5) # highest penalty values
```

Training and tuning the model. Control_grid() saves the validation set predictions so that diagnostic information is available after the model fit.
Area under ROC Curve measures the model performance across a continuum of event thresholds

```{r}
lr_res <-
    tune_grid(lr_workflow,
             resamples=cv_folds, #Using CV folds
             grid = lr_reg_grid,
            control = control_grid(save_pred = TRUE),
            metrics = metric_set(roc_auc))

#Visualizing the validation set metrics by plotting the area under the ROC curve
#against the range of penalty values
lr_plot <-
    lr_res%>%
    collect_metrics()%>%
    ggplot(aes(x=penalty, y=mean))+
    geom_point()+
    geom_line() +
    ylab("Area under the ROC Curve")+
    scale_x_log10(labels= scales::label_number())

lr_plot

figure_file = here("results","figures","ROC(AUC)_Penalty.png")
ggsave(filename = figure_file, plot=lr_plot)
```

The ROC curve looks unusually perfect. I am still trying to find out why it is so. Is it because the data is very simple and too easy to classify?

```{r}
# Selecting the best model automatically
lr_best1 <- select_best(lr_res, metric="roc_auc")

lr_best<- lr_res %>% 
  filter_parameters(parameters = lr_best1)%>% 
  collect_metrics()
lr_best

lr_auc <-
  lr_res %>% 
  collect_predictions(parameters = lr_best) %>%
  #names()
  roc_curve(ValuedCust, .pred_1)%>%
  mutate(model="Logistic Regression")

p4<-autoplot(lr_auc)

p4

figure_file = here("results","figures","ROC(AUC)_Validation.png")
ggsave(filename = figure_file, plot=p4)
```
The curve stays below the diagonal line for the entire range of 1-specificity indicating that the model choice is notappropriate.

In the below codes, I am double checking the results by creating a confusion matrix. First I created an object best_predictions based on the best hyperparameters. Then I merged it with the training data, calculated performance metrics, confusion matrix and roc_auc.

```{r}
#checking the best hyperparameters 
best_predictions <- lr_res %>% 
  collect_predictions(parameters = lr_best)

# Creating a predicted class column based on a threshold of 0.5
best_predictions <- best_predictions %>%
  mutate(.pred_class = if_else(.pred_1 > 0.5, 1, 0), #classifying prediction above 0.5 as 1 and other as 0
         .pred_class=factor(.pred_class, levels=c(0,1), labels=c("0", "1")),#converting the numeric variables to factor for confusion matrix
         ValuedCust_best=ValuedCust)%>% #
         select(-ValuedCust)

#Creating row id which will be used to merge with the object best_predictions
final_data_train_observed <- final_data_train %>%
  mutate(row_id = row_number())

# Joining the predictions with the observed values based on the row identifier
comparison <- final_data_train_observed %>%
  left_join(best_predictions, by = c("row_id" = ".row"))

# Calculating performance metrics
metrics <- comparison %>%
  yardstick::metrics(truth = ValuedCust, estimate = .pred_class)

# Calculating confusion matrix
conf_matrix <- comparison %>%
  yardstick::conf_mat(truth = ValuedCust, estimate = .pred_class)

#Calculating roc_auc to compare with the confusion matrix 
lr_auc_table1 <- comparison %>%
  yardstick::roc_auc(truth = ValuedCust, .pred_1)

# Viewing the results
print(metrics)
print(conf_matrix)
print(lr_auc_table1)

```
The confusion matrix suggests that the model perform very well on the training data as it almost perfectly identifies both classes. The roc_auc value does not align with the confusion matrix results.I re-calculated the ROC AUC using .pred_0

```{r}
#ROC AUC calculation based on .pred_0
lr_auc_table2 <- comparison %>%
  yardstick::roc_auc(truth = ValuedCust, .pred_0)

lr_auc_table2

#Visualizing the validation set ROC curve based on .pred_0
lr_auc_rev <-
  lr_res %>% 
  collect_predictions(parameters = lr_best) %>%
  #names()
  roc_curve(ValuedCust, .pred_0)%>%
  mutate(model="Logistic Regression")

p5<-autoplot(lr_auc_rev)

p5

figure_file = here("results","figures","ROC(AUC)_Validation_rev.png")
ggsave(filename = figure_file, plot=p5)
```
The roc-auc values aligns with the confusion matrix using the predicted probabilities for the negative class i.e. '.pred_0'.
The model seems to perform exceptionally well. The model shows high sensitivity for class 1 and high specificity for class 0, with correctly identifying almost all classes. This excellent performance may also suggest overfitting, which occurs when a model learns the training data too closely. In the next step, I applied the model to the test data to see its performance on unseen data.

###Applying the Model on Test Data

```{r}
# Step 1: Finalizing the Model
final_lr_mod <- logistic_reg(penalty = lr_best$penalty, mixture = 1) %>%
  set_engine("glmnet")

# Step 2: Fitting the Final Model on the entire training data
final_lr_recipe <- recipe(ValuedCust ~ ., data = final_data_train) %>%
  step_normalize(all_predictors(), -all_outcomes())

final_lr_workflow <- workflow() %>%
  add_model(final_lr_mod) %>%
  add_recipe(final_lr_recipe) %>%
  fit(data = final_data_train)

# Step 3: Predicting on Test Data
predictions <- predict(final_lr_workflow, final_data_test, type = "prob")

# Adding a column for the predicted class based on the probability threshold of 0.5
predictions <- predictions %>%
  mutate(predicted_class = if_else(.pred_1 > 0.5, "ValuedCust", "NotValuedCust"))
#predictions

# Step 4: Evaluate Model Performance
test_results <- final_data_test %>%
  bind_cols(predictions) %>%
  roc_curve(truth = ValuedCust, .pred_0) %>%
  autoplot()
test_results

p6<-test_results

p5+p6

figure_file = here("results","figures","ROC(AUC)_traintest.png")
ggsave(filename = figure_file, plot=p5+p6)

```
The roc_auc curve suggests that the model is also performing exceptionally well on the test data in classifying the two classes, with high accuracy, sensitivity and specificity. The data is relatively straightforward. This may have made it easier for the model to learn and predict accurately on the test data.

###Random Forest (RF)

In the following steps I used Random Forest to model the customer classification.

First I created the workflow following setting up the model for tuning and preparing recipe for the Random Forest model. I set up the no. of trees at 1000.
```{r}
#Setting up the model for tuning
rf_mod <- 
  rand_forest(mtry = tune(), min_n = tune(), trees = 1000) %>% 
  set_engine("ranger", seed=rngseed) %>% 
  set_mode("classification")

#Recipe for RF model
rf_recipe <- 
  recipe(ValuedCust ~ ., data = final_data_train)

#workflow for tunable RF model
rf_workflow <- 
  workflow() %>% 
  add_model(rf_mod) %>% 
  add_recipe(rf_recipe)

#rf_mod
#extract_parameter_set_dials(rf_mod)
```

I fitted the RF model using a space-filling design to tune, with 25 candidate models and plotted the results of the tuning process.

```{r}
#Running the model
rf_res <- 
  rf_workflow %>% 
  tune_grid(cv_folds,
            grid = 25,
            control = control_grid(save_pred = TRUE),
            metrics = metric_set(roc_auc))

rf_res %>% 
  show_best(metric = "roc_auc")

#plotting the results
autoplot(rf_res)
```
The results indicates that for optimal performance, two or three predictors are required at each node whereas the minimal number of data points required to keep splitting varied.

Next, I selected the best model according to the ROC AUC metric.
```{r}
#selecting best model
rf_best <- 
  rf_res %>% 
  select_best(metric = "roc_auc")
rf_best

```
Below codes collected the predictions required to plot the ROC curve. 
```{r}
rf_res %>% 
  collect_predictions()
```
Next, I filtered the predictions for the best random forest model.

```{r}
#collecting predictions from the best RF model
rf_auc <- 
  rf_res %>% 
  collect_predictions(parameters = rf_best) %>% 
  roc_curve(ValuedCust, .pred_0) %>% 
  mutate(model = "Random Forest")
```

I compared the validation set ROC curves for the top penalized logistic regression model and random forest model.

```{r}
#comparing roc curves from logistic regression and random forest models
p7<- bind_rows(rf_auc, lr_auc_rev) %>% 
  ggplot(aes(x = 1 - specificity, y = sensitivity, col = model)) + 
  geom_path(lwd = 1.5, alpha = 0.8) +
  geom_abline(lty = 3) + 
  coord_equal() + 
  scale_color_viridis_d(option = "plasma", end = .6)
p7

figure_file = here("results","figures","ROC(AUC)_lrvsrf.png")
ggsave(filename = figure_file, plot=p7)

```
Both models seem to perform equally good across event probability threshold.

In the subsequnt phase, the random forest model is evaluated on the testing data. 
First the model is finalized using the best hyperparameters identified during the tuning process. First I updated the model specification with the best parameters and then trained the final model on the entire training dataset.

```{r}
#Step1: Finalizing the model with best hyperparameters
final_rf_workflow <- finalize_workflow(rf_workflow, parameters = rf_best)

#Step2: Fit the model on the entire training data
final_rf_fit <- fit(final_rf_workflow, data = final_data_train)

#Step3: Predict on the test data
rf_test_predicts <- predict(final_rf_fit, new_data = final_data_test, type = "prob")

#Adding a colum for the predicted class based on the probability threshold of 0.5
rf_test_predicts <- rf_test_predicts%>%
  mutate(predicted_class = if_else(.pred_1>0.5, "ValuedCust", "NotValuedCust"))

#Step4: Evaluate the model performance on test data
rf_roc_auc_test <- final_data_test%>%
  bind_cols(rf_test_predicts)%>%
  roc_curve(truth = ValuedCust, .pred_0)%>%
  autoplot()
p8<-rf_roc_auc_test
p8

figure_file = here("results","figures","ROC(AUC)_rftest.png")
ggsave(filename = figure_file, plot=p8)

```

### 2024-04-06 products/manuscript/Manuscript.qmd
Project Part 4

@@
 
-This project aims to leverage statistical modeling techniques to predict customer segmentation based on their purchasing behavior pattern, using both supervised and unsupervised learning approaches.
+This project aims to leverage statistical modeling techniques to predict customer segmentation based on their purchasing behavior pattern, using the Supervised and Machine Learning approaches.
 In the initial phase, the analysis employs logistic regression to segregate customers based on the general patterns of purchasing behavior.
@@
 Each customer was assigned a value ranging from 1 to 4 for the Recency, Frequency and Monetary attributes, reflecting their quartile position within each metric.
-Subsequently, these metrics were combined to formulate a three-digit score for every custoemr, positioning the Monetary value in the hundreds place, Frequency in the tens place, and Recency in the units place.
+Subsequently, these metrics were combined to formulate a three-digit score for every customer, positioning the Monetary value in the hundreds place, Frequency in the tens place, and Recency in the units place.
 All the customers with a score above 344 were assigned as High-Value Customers.
@@
 
-{{< pagebreak >}}
-
 # Results
@@
 ## Exploratory/Descriptive analysis
-
 @tbl-resulttable1 depicts the descriptive statistics of various aspects of customers purchase behavior.
@@
 #| label: fig-result1
-#| fig-cap: "Distribution of Sales and Purchase frequency"
+#| fig-cap: "Distribution of Valued Customers"
 #| echo: FALSE
@@
 
-@fig-result2 depicts two density plots for the Cumulative Sales and Invoice Values across customers and a histogram for Customer purchase frequencies.
-Both density plots are heavily right-skewed.The averages of total sales to a customer and total invoice values appear to be less than 2500 and 500 sterling pounds, respectively, supporting the mean values of 1532 and 436 in the descriptive table.
-The average frequency appears to be less than 5 times, which is also observed in the descriptive statistics table.
+@fig-result2 depicts a correlation plot for Monetary, Frequency and Recency embedded with density plots.
+All of the density plots are heavily right-skewed.
+Monetary amount has a positive correlation of 0.787 with Frequency and a negative correlation of 0.334 with Recency.
+Likewise, the correlation between Recency and Frequency is negativ with a magnitude of 0.376.
+The correlation values indicated that collinearity is unlikely to be an issue for the analysis.
 
@@
 #| label: fig-result2
-#| fig-cap: "Distribution of Sales and Purchase frequency"
+#| fig-cap: "Correlation Plot"
 #| echo: FALSE
-knitr::include_graphics(here("results", "figures", "Sales_distribution.png"))
+#knitr::include_graphics(here("results", "figures", "Sales_distribution.png"))
+knitr::include_graphics(here("results", "figures", "correlations.png"))
 ```
-@fig-result3 depicts a density plot of number of days since the last transaction made by a customer.
-The distribution is right-skewed.
-This plot indicates that most customers made a purchase transaction within the last 30 days of the reference date.It also reveals that few customers had not make a transaction for a long time.
+{{< pagebreak >}}
 
+@fig-result3 depicts a trend of monthly international sales.
+A spike is noticed in November, presumably due to Christmas in the following month, which is also supported by distribution of the Recency plot.
 ```{r}
 #| label: fig-result3
-#| fig-cap: "Distribution of Recency of Purchase"
-#| echo: FALSE
-knitr::include_graphics(here("results", "figures", "Recency.png"))
-```
-
-@fig-result4 depicts a trend of monthly international sales.
-A spike is noticed in November, presumably due to Christmas in the following month, which is also supported by distribution of the Recency plot.
-```{r}
-#| label: fig-result4
 #| fig-cap: "Monthly Sales trend"
@@
 ```
-@fig-result5 depicts country-wise trend of international sales by sales amount. Top customers were in Germany, France, Spain and Belgium.  
+{{< pagebreak >}}
+
+@fig-result4 depicts country-wise trend of international sales by sales amount. Top customers were in Germany, France, Spain and Belgium.  
 ```{r}
-#| label: fig-result5
+#| label: fig-result4
 #| fig-cap: "Countrywise Sales trend"
@@
 
-Penalized logistic regression of customer segment was performed on Recency, Frequency and Monetary variables using 'tidymodels' package and 4-fold corss-validation techniques.
+First, I fitted the traditional logistic regression model to segregate the high value clients without applying the cross-validation.
+The results  returned a warning message that the algorithm did not converge and that the fitted probabilities numerically 0 or 1 occurred.
+I computed the accuracy and AUC despite this message.
 
-@fig-result6 depicts the area under ROC curve by the range of penalty values.
-It provides visualization of validation set metrics.
-The curve has extreme high values, all above 0.9995.
-A sharp drop is noticed before the penalty value of 0.0010, suggesting the sensitivity of the performance of the model at this penalty value.
-This is followed by a sharp rise after the penalty value of 0.01.
-These features of model performance are indicative of the training data and the data being easily separable.
+{{< pagebreak >}}
 
-```{r}
-#| label: fig-result6
-#| fig-cap: "Visualization of validation set metrics"
+@tbl-resulttable2 depicts the accuracy of and ROC AUC metrics of the model.
+The model appears to have a perfect accuracy in predicting the customer segment based on the accuracy metric however it does not seem to have not even random guessing ability based on the ROC AUC metric.
+
+```{r out.width="100%", fig.show='hold', warning=FALSE}
+#| label: tbl-resulttable2
+#| tbl-cap: "Model Metrics"
 #| echo: FALSE
-knitr::include_graphics(here("results", "figures", "ROC(AUC)_Penalty.png"))
-```
-@fig-result7 depicts validation set ROC curve plotting true positive rate (sensitivity) against the false positive rate (1-specificity) at various threshold settings.
-The curve moves along 1-specificity and rises only from the end.This suggests that the model has no  ability to distinguish between the two customer categories.
-This suggests that the logistic regression model is not appropriate to the dataset.
-It pointed out the need to revisit the choice of model.
+resulttable2 = readRDS(here("results", "tables", "logit_combinedstats.rds"))
+knitr::kable(resulttable2)
 
-```{r}
-#| label: fig-result7
-#| fig-cap: "Visualization of Validation set ROC curve"
-#| echo: FALSE
-knitr::include_graphics(here("results", "figures", "ROC(AUC)_Validation.png"))
 ```
@@
 
-#Example @tbl-resulttable2 shows a summary of a linear model fit.
+### Penalized Logistic Regression with cross-validation
+For further analysis, I fitted a penalized logistic regression to the data using 'tidymodels' package and 10-fold cross-validation technique.
+The goal was to obtain the best value for predictions by tuning the model hyperparametes and using a mixture that chooses a simple model.
+
+The data has 25% of the customers grouped as Valued Customer.
+Due to this imbalanced proportions of the segment, the data was stratified by the outcome variable for splitting into training and testing sets to ensure that those subsets include similar proportion of Valued Customer.
+
+{{< pagebreak >}}
+
+@fig-result5 depicts the area under ROC curve by the range of penalty values.
+It provides visualization of validation set metrics.
+The curve has extreme high values, all at 1 before the penalty value of 1 followed by a sharp drop, suggesting the sensitivity of the performance of the model at the corresponding penalty value.
+These features of model performance are indicative of the training data being easily separable.
 
 ```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
+#| label: fig-result5
+#| fig-cap: "Area under ROC curve by Penalty"
 #| echo: FALSE
-#resulttable2 = readRDS(here("results","tables","summarytable.rds"))
-#knitr::kable(resulttable2)
+knitr::include_graphics(here("results", "figures", "ROC(AUC)_Penalty.png"))
 ```
+{{< pagebreak >}}
+
+@fig-result6 depicts validation set ROC curve plotting true positive rate (sensitivity) against the false positive rate (1-specificity) at various threshold settings.
+The curve moves along 1-specificity and rises only from the end.This suggests that the model has no  ability to distinguish between the two customer categories.
+Contrarily, as in the case of traditional logistic model, the accuracy metrics of this validation set has a high value, 0.9966555.This suggested the need of digging down to the reason of these opposing model performances.
+The confusion matrix revealed that it almost perfectly identified both classes except for one date point.
+This indicated a potential inaccuracy or misinterpretation in the ROC curve plotting.
+
+```{r}
+#| label: fig-result6
+#| fig-cap: "Visualization of Validation set ROC curve"
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "ROC(AUC)_Validation.png"))
+```
+{{< pagebreak >}}
+
+I plotted the ROC curve for the predicted probabilities for the negative class i.e.
+customer category '0' representing 'Not a Valued Customer'.
+Using this approach, the ROC-AUC values aligned with the confusion matrix and resulted into a almost perfect curve.
+The model showed high sensitivity for class 0 and high specificity for class 1, with correctly identifying almost all classes.
+### (I need to figure out if this result is valid or interpretable or what caused this. The subsequent parts are assuming this result is valid)
+
+```{r}
+#| label: fig-result7
+#| fig-cap: "Validation set using Penalized Logistic Regression"
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "ROC(AUC)_Validation_rev.png"))
+```
+In the next step, I applied the model to the test data to see its performance on unseen data.
+The resulting ROC_AUC curve also suggested that the model performed exceptionally well on the test data in classifying the two classes, with high accuracy, sensitivity and specificity.
+This excellent performance might potentially due to overfitting, which occurs when a model learns the training data too closely.
+The data is relatively straightforward.
+This may have made it easier for the model to learn and predict accurately on the test data.
+
+### Random Forest with cross-validation
+In the next phase, I fitted a Random Forrest model with 10-fold cross validation using a space-filling design to tune, with 25 candidate models and plotted the results of the tuning process.
+The results indicated that for optimal performance, two or three predictors are required at each node with a varied minimal number of data points for splitting.
+The model is found to perform exceptionally well with high accuracy.
+
+{{< pagebreak >}}
+
+I compared the validation set ROC curves for the top penalized logistic regression model and random forest model. Both models are found to perform well.
+
+```{r}
+#| label: fig-result8
+#| fig-cap: "Validation set: Penalized Logistic Regression vs Fandom Forest"
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "ROC(AUC)_lrvsrf.png"))
+```
+
+In the subsequent phase, the random forest model is evaluated on the testing data and is found to perform equally well on the testing data.
 

### 2024-04-06 R/analysis-code/statistical-analysis_superstoreRFM.R

@@ -1,282 +1,0 @@
-###############################
-# analysis script
-#
-#this script loads the processed, cleaned data, does a simple analysis
-#and saves the results to the results folder
-
-#loading packages needed. 
-library(here) #for data loading/saving
-library(tidymodels) # for the parsnip package, along with the rest of tidymodels
-library(ggplot2) #for plotting
-library(dplyr) #for data wrangling
-library(broom) #for cleaning up output from lm()
-library(lubridate)
-library(glmnet) #for fitting GLM via penalized maximum likelihood
-library(vip) # for visualizing the variable importance scores for the top 20 features
-#library(ranger) #for random forest. Unable to sync this in renv
-
-#path to data using the here() package
-data_location <- here::here("data","processed-data","processed_superstore_RFM.rds")
-
-#load data. 
-mydata <- readRDS(data_location)
-
-#Checking data
-skimr::skim(mydata)
-head(mydata)
-
-#This study aims to use logistic regression to segment customers based on the their purchasing behavior reflected by 
-#the monetary purchase amount and loyalty of the customer   
-#In absence of a readily available classification variable in the data, one will be created based on three factors:
-# Recency, Frequency and Monetary. 
-#Recency is the number of days since the last purchase by a customer. 
-#Frequency is the no. of purchases by a customer over a given period.
-#Monetary is the total amount of purchase a customer has made over a period. 
-
-# For determining recency of a transaction, a reference date past the study period is required. The data covers transactions till 2011-12-09.
-# The next day i.e. 2011-12-10 is considered as the reference date to define recency.
-
-ref_date <- as.Date("2011-12-10")
-
-# Next, a new data frame is created which includes the segments, Recency, Frequency and Monetary
-rfm_data <- mydata %>%
-  group_by(CustomerID) %>%
-  summarise(
-    Recent_Purchase_Date=last(InvoiceDate),
-    Recency = as.numeric(ref_date - as.Date(max(InvoiceDate))),
-    Frequency = n_distinct(InvoiceNo),
-    Monetary = sum(InvoiceValue)
-    
-  ) %>%
-  ungroup()
-
-skimr::skim(rfm_data)
-
-#Density Plot of Recency
-
-p1 <- ggplot(rfm_data, aes(x = Recency)) +
[…]

### 2024-04-22 R/analysis-code/statistical_analysis-all.qmd
Project Part 5

---
title: "The Value of Loyalty: Customer impact on Online Store Revenue"
format: html
editor: visual
---

## Background

The project aims to deepen the understanding of the customer base by identifying the core customers who demonstrate loyalty.This is achieved by focusing on customers who have purchased from the store during both the typically steady first five months of the year and the peak sales season.The sales data indicated that while the initial months show a modest performance, the latter part of the year, influenced by the holiday season, witnessed a surge in sales activity. Customers purchasing in the absence of major festive events are presumed to be driven by the intrinsic value of the product range, rather than seasonal discounts. Conversely, during the peak season, when the competition and promotional offers peak, those who return to make purchases can be considered as the store's regular/loyal clientele. These customers are invaluable throughout the year, contributing to consistent sales and enhancing revenue during peak times.

A rigorous analysis is set to measure the impact of this customer segment on the overall business performance. The outputs of the analysis is anticipated to help refining marketing strategies and improving customer service practices to strengthen customer retention.

First required packages are loaded.

```{r}
#| echo: false
#| warning: false
#| message: false
#loading packages needed. 
library(here) #for data loading/saving
library(tidymodels) # for the parsnip package, along with the rest of tidymodels
library(ggplot2) #for plotting
library(dplyr) #for data wrangling
library(gridExtra) # for combined display of plots
library(broom) #for cleaning up output from lm()
library(lubridate)#for date formating
library(glmnet) #for fitting GLM via penalized maximum likelihood
library(mgcv) #for Generalized Additive model

```

The processed data file is imported and visualied the summaries.

```{r}
#| echo: false
#| warning: false
#| message: false
#path to data using the here() package
data_location <- here::here("data","processed-data","processed_superstore.rds")
#load required data. 
mydata <- readRDS(data_location)
#Summarizing and checking data
skimr::skim(mydata)
head(mydata)
```

The dataset encompasses 14951 transactions involving 4117 customers across 33 countries, with 3785 distinct product types being sold. It is comprehensive, with no missing values and includes essential information such as Invoice No., Invoice Date, Stock Code, Description, CustomerID, Country, Quantity, Unit Price, and Sales.

### Feature Engineering

To adapt the data for customer-level analysis, we will aggregate sales by CustomerID, thus summing up the total purchases for each customer. We will also count the unique Invoices per CustomerID to determine the purchase frequency.

Preliminary data exploration has shown that sales peaked from September to December. We will classify any customer who has made at least one purchase before June 2011 and at least one between September and December as 'loyal.'

To discern the purchasing behaviors across different regions, we have introduced a binary variable that distinguishes between UK-based customers and those from other countries.

The sum of purchases by customers will serve as the dependent variable in our study. The independent variables will include customer loyalty, purchase frequency, and geographic location (UK or not). Given the limited number of predictors, strategic analysis execution is essential.

In order to understand the relationship between the sum of purchases and customer loyalty, we will evaluate three distinct predictive models, each offering unique approaches to regression analysis:

### Linear Regression Model:

This model is chosen for its simplicity and interpretability. It assumes a linear relationship between the predictors and the dependent variable, which is a good starting point for many modeling tasks. It will serve as a benchmark for the performance of more complex models.

### LASSO Regression:

LASSO (Least Absolute Shrinkage and Selection Operator) is a type of linear regression that includes a penalty term to regularize the coefficients. This model is particularly useful when some of the predictors are irrelevant for the output. LASSO can help in feature selection by shrinking the less important feature coefficients to zero, thus simplifying the model and potentially improving prediction performance.

### Generalized Additive Model (GAM):

GAMs extend linear models by allowing for non-linear relationships between each predictor and the response. This is achieved by fitting a smooth function to each predictor. This flexibility makes GAMs powerful tools for capturing complex patterns in the data, which linear models might miss. They are also still interpretable, as we can visualize the effect of each predictor on the response.

By comparing these models, we can understand how the addition of complexity and flexibility impacts the predictive accuracy, as measured by RMSE. The selected model will balance the trade-off between bias and variance, aiming for the lowest possible RMSE, which indicates the smallest average prediction error over our dataset."

```{r}
#| echo: false
#| warning: false
#| message: false

#Classifying loyal customers - Customers who purchased at least once before June 2011 and then came back at least once during the period from September through December

mydata <- mydata %>%
  group_by(CustomerID) %>%
  # Creating binary variables for purchases before June 2011 and after October 2011
  mutate(
    PurchaseBeforeJun2011 = any(InvoiceDate < as.Date("2011-06-01")),
    PurchaseAfterOct2011 = any(InvoiceDate > as.Date("2011-10-01"))
  ) %>%
  # A customer is coded as loyal if they made at least one purchase in each period
  mutate(loyal_cust = as.integer(PurchaseBeforeJun2011 & PurchaseAfterOct2011)) %>%
  # Calculate frequency of purchase and sum of total purchase by each customer
  mutate(
    freq_purchase = n_distinct(InvoiceNo),
    sum_purchase = sum(Sales)
  ) %>%
  ungroup() %>%
  # Create a binary variable for UK sales and convert loyal_cust to a factor
  mutate(
    UK_sales = ifelse(Country == "United Kingdom", 1, 0),
    loyal_cust = as.factor(loyal_cust)
  ) %>%
  # Remove intermediate variables for clarity
  select(-PurchaseBeforeJun2011, -PurchaseAfterOct2011)
skimr::skim(mydata)
```

We created a final dataset for analysis wherein each customer is uniquely observed.As per the exploratory data analysis, both the sum of purchases and frequency of purchases are heavily right-skewed. Thus, these predictors are log transformed for the linear model.

```{r}
#Creating a new file by aggregating sales and purchase counts by CustomerID with required columns
final_data <- mydata%>%
  group_by(CustomerID)%>%
  summarise(
    sum_purchase=first(sum_purchase),
    freq_purchase=first(freq_purchase),
    loyal_cust=first(loyal_cust),
    UK_sales=first(UK_sales)
    )%>%
  mutate(loyal_cust=as.factor(loyal_cust),
         #log transforming sum of total purchase and frequency of purchase 
         log_sumpurchase = log(sum_purchase),
         log_purfreq = log (freq_purchase))

#Summarizing the data
skimr::skim(final_data)
#checking the ratio of loyal customers
final_data%>%count(loyal_cust)%>%mutate(proportion = n /sum(n))
```

We visualize the distribution of loyal customers using a box plot. We also visualize the distribution of sum of purchase by purchase frequency stratified by loyal customers.

```{r}
# Boxplot for sum_purchase by loyal_cust
p1<-ggplot(final_data, aes(x = as.factor(loyal_cust), y = sum_purchase)) + 
    geom_boxplot() +
    labs(title = "Total Purchases by Customer Status",
       x = "loyal customer",
       y = "sum of purchase")+
    theme_minimal()
p1

# Scatterplot of sum_purchase vs freq_purchase stratified by loyal customers 
p2<-ggplot(final_data, aes(x = freq_purchase, y = sum_purchase)) + 
    geom_point(aes(color = as.factor(loyal_cust))) +
    labs(title = "Frequency of Purchase vs Total Purchases",
       x = "Purchase frequency",
       y = "Sum of purchase") +
    theme_minimal()
p2

figure_file = here("results","figures","Purchase by loyaly.png")
ggsave(filename = figure_file, plot=p1)

figure_file = here("results","figures","Purchase by frequency.png")
ggsave(filename = figure_file, plot=p2)

```

```{r}
# Visualize the distribution of log transformed sum_purchase
p3<-ggplot(final_data, aes(x = log_sumpurchase)) + 
    geom_histogram(binwidth = 1, fill = "lightgrey", color = "black") +
    labs(title = "Distribution of log transformed sum of Purchases",
       x = "log of Purchase")+
    theme_minimal()

# Visualize the distribution of log transformed freq_purchase
p4<-ggplot(final_data, aes(x = log_purfreq)) + 
    geom_histogram(binwidth = 1, fill = "lightgrey", color = "black") +
    labs(title = "Distribution of log transformed Purchase frequencies",
         x="log of Purchase frequency")+
    theme_minimal()

plot1<-grid.arrange(p3, p4, ncol = 2, nrow=1)

plot1

figure_file = here("results","figures","distribution of transformed vars.png")
ggsave(filename = figure_file, plot=plot1)

```

### Data Set Up

Now we partition the data into train and test sets. We use the train set to teach the model to recognize patterns and ultimately use the selected model on the test set to evaluate the selected model's performance, mimicking real world application.The split of 75:25, balances the model's exposure to enough data for learning with an adequate reserve for unbiased evaluation.

```{r}
#Set random number by setting the seed for reproducibility
set.seed(123)

# Split the data into training and testing sets in the ratio of 75:25
data_split <- initial_split(final_data, prop = 0.75)
#Create training and testing sets
train_data <- training(data_split)
test_data <- testing(data_split)
```

First, we examine a simple linear model, with 'loyal customer' serving as the only predictor of the log-transformed sum of purchases, utilizing the tidy model workflow. Following this, we will enhance our analysis by fitting a multivariate linear model, incorporating additional predictors: the log-transformed frequency and the geographical region of sales. Subsequently, we will predict outcomes using both models on the training set and compare their performance.

### Simple Linear Regression Model

```{r}
# Create a recipe for fitting the model with only main predictor 
slr_recipe <- recipe(log_sumpurchase ~ loyal_cust, data = train_data)

# Define the simple linear regression model specification
slr_model <- linear_reg() %>%
    set_engine("lm") %>%
    set_mode("regression")

# Bundle the recipe and model spec into a workflow
slr_workflow <- workflow() %>%
    add_recipe(slr_recipe) %>%
    add_model(slr_model)

# Fit the model
slr_fit <- slr_workflow %>%
    fit(data = train_data)

# Summarize the model fit
slr_fit_df<-tidy(slr_fit)

file_path = here("results","tables","slr_fit_df.rds")
saveRDS(slr_fit_df, file_path, )

# Predict on the training set
slr_log_predicts <- predict(slr_fit, new_data = train_data) %>%
    bind_cols(train_data)

#Converting the predicts to original scale

slr_predicts <- slr_log_predicts%>%
  mutate(slr_predicts = exp(.pred)) %>%
  select(CustomerID, sum_purchase, slr_predicts, log_sumpurchase, .pred)

# Evaluate the model
metrics <- metric_set(rmse, rsq)
slr_metrics <- metrics(data = slr_predicts, truth = sum_purchase, estimate = slr_predicts)
print(slr_metrics)

```

### Multivariate Linear Regression

```{r}
# Create a recipe for fitting the model with other predictors 
mlr_recipe <- recipe(log_sumpurchase ~ loyal_cust + log_purfreq + UK_sales, data = train_data)

# Define the multivariate linear regression model specification
mlr_model <- linear_reg() %>%
    set_engine("lm") %>%
    set_mode("regression")

# Bundle the recipe and model spec into a workflow
mlr_workflow <- workflow() %>%
    add_recipe(mlr_recipe) %>%
    add_model(mlr_model)

# Fit the model
mlr_fit <- mlr_workflow %>%
    fit(data = train_data)

# Summarize the model fit
tidy(mlr_fit)

#Model Evaluation: Predict on the training set
mlr_log_predicts <- predict(mlr_fit, new_data = train_data) %>%
    bind_cols(train_data)

#Converting the predicts to original scale
mlr_predicts <- mlr_log_predicts%>%
  mutate(mlr_predicts = exp(.pred)) %>%
  select(CustomerID, sum_purchase, mlr_predicts, log_sumpurchase, .pred)

# Evaluate the model
metrics <- metric_set(rmse, rsq)
mlr_metrics <- metrics(data = mlr_predicts, truth = sum_purchase, estimate = mlr_predicts)
print(mlr_metrics)

```

We found that the multivariate model significantly improved the prediction accuracy, evidenced by a reduced RMSE of 689, compared to the simple linear model's RMSE of 1308. This substantial reduction in RMSE indicates that the addition of log-transformed frequency and geographical region as predictors enhanced the model's ability to predict the log-transformed sum of purchases.

Next, we used cross-validation techniques of 10-fold on the multivariate linear regression model to further validate its predictive accuracy. This method helps in mitigating overfitting and provides a comprehensive view of how well the multivariate model generalizes to unseen data.

```{r}
# Create a recipe for preprocessing
mlr_cv_recipe <- recipe(log_sumpurchase ~ loyal_cust + log_purfreq + UK_sales, data = train_data)

# Define the model specification
mlr_cv_spec <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

# Set up cross-validation at 10-fold
mlr_cv_folds <- vfold_cv(train_data, v = 10, repeats = 1)

# Bundle the recipe and model spec into a workflow
mlr_cv_workflow <- workflow() %>%
  add_recipe(mlr_cv_recipe) %>%
  add_model(mlr_cv_spec)

# Perform cross-validation
mlr_cv_fit <- fit_resamples(
  mlr_cv_workflow,
  mlr_cv_folds,
  control = control_resamples(save_pred = TRUE)
)

#Model Evaluation: Predict on the training set
mlr_cv_log_predicts <- mlr_cv_fit %>% 
  collect_predictions()

# Transform predictions and actual values back to the original scale
mlr_cv_log_predicts<- mlr_cv_log_predicts %>%
  mutate(pred_original_scale = exp(mlr_cv_log_predicts$.pred),
         sum_purchase = exp(log_sumpurchase))

# Evaluate the model
metrics <- metric_set(rmse, rsq)
# Collect metrics
mlr_cv_metrics <- metrics(data=mlr_cv_log_predicts, truth=sum_purchase, estimate= pred_original_scale)
print(mlr_cv_metrics)

```

The RMSE from Cross-validation is 690.49 which is very close to the RMSE of 689.29 of the model without cross-validation. this suggests that the multivariate linear model is relatively stable.

To further enhance model performance and explore the potential for a more parsimonious model, we next turn to LASSO regression. LASSO is an extension of linear regression which introduces a regularization term that shrinks some coefficients of the variables to zero, effectively eliminating those variables from the the model to simplify it. Though our model has only three predictors, this property of LASSO helps selecting only those predictors which are relevant to the outcome variable. By penalizing the magnitude of the coefficients, LASSO aims to reduce overfitting, improve model interpretability, and potentially uncover a more predictive subset of features.

First I fit lasso model without tuning of penalty by setting up a minimal penalty of 0.1. Factor variables are encoded using dummy variables. As normalization is used in fitting lasso, numerical variables are kept at original scale.

```{r}
#Preparing recipe
#Recipe for lasso model 
lasso_recipe<-
  recipe(sum_purchase ~ loyal_cust + freq_purchase + UK_sales, data = train_data) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_normalize(all_predictors())

# Defining the lasso model setting the penalty at 0.1
lasso_model<- linear_reg(penalty = 0.1, mixture = 1)%>%
  set_engine("glmnet")%>%
  set_mode("regression")

#workflow for lasso model without tuning of penalty
lasso_workflow <- workflow()%>%
  add_model(lasso_model)%>%
  add_recipe(lasso_recipe)
#Fitting the lasso model
lasso_fit <- lasso_workflow %>%
  fit(data = train_data)

```

Next step is model prediction and model performance evaluation on the train data.

```{r}
#Computing model predictions
lasso_predicts <- predict(lasso_fit, new_data = train_data)

#augment to evaluate performance metric
lasso_aug <- augment(lasso_fit, train_data)
lasso_aug %>% select(sum_purchase,.pred)

#Model evaluation: Calculatin of RMSE
lasso_metrics <- lasso_aug %>% metrics(truth=sum_purchase, .pred)
lasso_metrics

```

We can see the improvement in model performance with a reduced RMSE of 675.38 compared to the previously observed RMSE of 690.49 from the linear model with cross-validation.

Subsequently, we focused on refining the LASSO model by incorporating tuning and leveraging cross-validation. The goal is to further optimize model parameters and ensure robustness by systematically evaluating the model's performance across various data subsets.

We tuned the lasso model employing 10-fold cross-validation strategy with 5 repeats.

```{r}
#This recipe is same as it was in the previous simple lasso model
lasso_recipe<-
  recipe(sum_purchase ~ loyal_cust + freq_purchase + UK_sales, data = train_data) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_normalize(all_predictors())

#Setting up the model so that the tuning function (tune())can work
lasso_tunable <- linear_reg(penalty = tune(), mixture = 1)%>%
  set_engine("glmnet")%>%
  set_mode("regression")

#workflow for tunable LASSO model
lasso_wf_tunable <- workflow()%>%
  add_model(lasso_tunable)%>%
  add_recipe(lasso_recipe) 
```

We executed the tuning using a grid search over penalty values ranging from 10\^-1 to 10\^3.

```{r ehco = TRUE}

# Define a penalty parameter object with a range from 10^-1 to 10^3
penalty_param <- penalty(range = c(-1, 3))
# Create a regular grid of penalty values with 50 levels
penalty_grid <- grid_regular(penalty_param, levels = 50)

```

we tuned the model using a 10-fold cross validation.

```{r}
#For reproducibility
set.seed(1234)

#Resampling the data using cross validation.   
resamples_cv<-vfold_cv(train_data, v=10, repeats = 5)

#Tuning the lasso model
lasso_tune_results_cv <- lasso_wf_tunable %>%
  tune_grid(resamples = resamples_cv, grid = penalty_grid)

```

Visualizing the diagnostics from the lasso tuning results to evaluate model performance.

```{r}
#plotting the diagnostics
p3<-lasso_tune_results_cv%>%
  autoplot()
p3

figure_file = here("results","figures","lasso_diagnostics.png")
ggsave(filename = figure_file, plot=p3)

```

The model diagnostic plot reveals that lasso performs well at lower penalty values as evidenced by the value of corresponding RMSE metrics. As the penalty value goes up, especially after a certain threshold, the model becomes more regularized and some coefficients may shrink up-to zero increasing bias and the RMSE value. Lasso behaves similar to a linear model for very low penalty values resulting to similar values of RMSE. It is because at lower RMSE values the model is less regularized.

Now, we evaluate the model performance and also examine the variables that are retained in the final model. We first identify the penalty parameter (lambda) that resulted in the lowest RMSE during the tuning process.

```{r}
#Retrieving the best penalty parameter from the tuning process 
best_lambda <- lasso_tune_results_cv %>%
  select_best(metric = "rmse") 
best_lambda
```

The penalty value of the best model is similar to what we have used on the initial Lasso model. So, we expect the model metrics of the best model to be similar to the initial model. We set this explicit parameter and retrained the model on the entire training set.Then we compute the RMSE of the best model. We also extracted coefficients of the best model which allowed us to examine the predictor variables retained in the model.

```{r}
# Specify the LASSO model with the best penalty parameter
final_lasso_model <- linear_reg(penalty = best_lambda$penalty, mixture = 1) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

# Create the final recipe (same as the previous)
final_lasso_recipe <- recipe(sum_purchase ~ loyal_cust + freq_purchase + UK_sales, data = train_data) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_normalize(all_predictors())

# Create the workflow with the final model and recipe
final_lasso_workflow <- workflow() %>%
  add_model(final_lasso_model) %>%
  add_recipe(final_lasso_recipe)

# Fit the final model on the entire training dataset
final_lasso_fit <- final_lasso_workflow %>%
  fit(data = train_data)

# Retrieving prediction to compute RMSE using the best penalty parameter 
final_lasso_predicts <- predict(final_lasso_fit, train_data) %>%
  bind_cols(train_data)

# Evaluate the model
metrics <- metric_set(rmse, rsq)
final_lasso_metrics <- metrics(data = final_lasso_predicts, truth = sum_purchase, estimate = .pred)

# Extract coefficients from the final model
final_lasso_coef <- final_lasso_fit %>%
  extract_fit_parsnip() %>%
  tidy()

final_lasso_coef
final_lasso_metrics
```

Apparently, the cross-validated lasso fitted model retained all three predictor variables and resulted RMSE similar to that from the untuned LASSO model. This outcome suggested the relevance of all predictor variables in the model.

In the next step, we explore the possibility of nonlinear relationships between predictors and the outcome variables.We will apply Generalized Additive Model(GAM). It is chosen for its ability to effectively identify and illustrate complex relationships in the data.

GAM is fitted using the REML method. The use of REML invokes the use of restricted Maximum Likelihood for estimating the smoothness parameters of the model. REML helps in objectively choosing a level of smoothness that balances fit and complexity, based on the data itself.

```{r}
# Create a recipe for fitting the model with other predictors 
gam_recipe <- recipe(sum_purchase ~ loyal_cust + freq_purchase + UK_sales, data = train_data)%>%
  step_normalize(all_numeric(), -all_outcomes())#normalizes the numeric predictor 'freq_purchase' ensuring exclusion of the outcom variable in this process

# Prep and bake the recipe to apply transformations
prepped_data <- prep(gam_recipe, training = train_data)
#Apply the transformation to the training data
baked_data <- bake(prepped_data, new_data = NULL)  # Apply transformations to the entire dataset

# Fit the GAM model with the preprocessed data usint the method "REML"
gam_fit <- gam(sum_purchase ~ s(freq_purchase) + loyal_cust + UK_sales, data = baked_data, method = "REML")

# Summarize the model fit
summary(gam_fit)

#Model Evalutation: Predict on the baked training set
gam_predicts <- predict(gam_fit, new_data = baked_data, type = "response")

#Combining the predicts with the baked_data for the purpose of metrics calculations
gam_predicts_df<-data_frame(gam_predicts)%>%
  bind_cols(baked_data)

#Converting to a tibble for compatibility with yardstick functions
gam_predicts_df <- as_tibble(gam_predicts_df)

#Evaluate the model
gam_metrics <- gam_predicts_df %>%
  metrics(truth = sum_purchase, estimate = gam_predicts)

print(gam_metrics)

```

We can see the improvement in model performance using GAM with a reduced RMSE of 661.05 compared to the RMSEs of 675.38 from the Lasso model and 690.49 from the cross-validated linear model.

Following this, we focused on enhancing the performance of GAM by implementing cross-validation techniques. The objective is to further optimize model parameters aiming for an optimal balance between fit and complexity.

```{r}

#Set random number by setting the seed for reproducibility
set.seed(123)

#Creating 10-fold cross-validation from training data.   
gam_cv<-vfold_cv(train_data, v=10)

#As per ChatGPT, 'parsnip' doesn't directly support 'gam()' from 'mgcv'. So we loop through each of the CV folds, fit the GAM model using 'mgcv' directly and collect the prediction.
gam_cv_results<- gam_cv$splits%>%

  map_df(function(split){
    #Prepare the training and testing sets for this fold
    train_split <- training(split)
    test_split <- testing(split)

    #Apply the recipe to the training data
    prepped_data <- prep(gam_recipe, training = train_split)
    baked_train <- bake(prepped_data, new_data = NULL)
    baked_test <- bake(prepped_data, new_data = test_split)

    #Fit the GAM model to the baked training data
    gam_cv_fit <-gam(sum_purchase ~ loyal_cust + s(freq_purchase)+ UK_sales,
                  data = baked_train, method = "REML")

    #Predict on the baked testing set
    gam_cv_predicts<- predict(gam_cv_fit, newdata=baked_test, type="response")

    #Return the actual values and predictions in a tibble
    tibble(
      truth = baked_test$sum_purchase,
      estimate=gam_cv_predicts
    )
  })
#Collecting performance metrics
gam_cv_metrics <- gam_cv_results %>%
  metrics(truth = truth, estimate = estimate)
gam_cv_metrics

```

We see that the RMSE goes upto 720.50 by using cross-validated GAM model compared to that of 661.05 with GAM model without cross-validation. This increase in RMSE suggests that the model's predictions are, on average, less accurate across different subsets of the data than suggested by the initial evaluation on the training data alone.

### Model Selection

We list the RMSEs of all the models that we have fitted so far and select the one with lowest RMSE.

```{r}
# Extract RMSE values
slr_rmse <- slr_metrics %>% 
  filter(.metric == "rmse") %>% 
  pull(.estimate)

mlr_rmse <- mlr_metrics %>% 
  filter(.metric == "rmse") %>% 
  pull(.estimate)

mlr_cv_rmse <- mlr_cv_metrics %>% 
  filter(.metric == "rmse") %>% 
  pull(.estimate)

lasso_rmse <- lasso_metrics %>%
  filter(.metric == "rmse") %>%
  pull(.estimate)

lasso_cv_rmse <- final_lasso_metrics %>%
  filter(.metric == "rmse") %>%
  pull(.estimate)

gam_rmse<-gam_metrics %>%
  filter(.metric == "rmse") %>%
  pull(.estimate)

gam_cv_rmse<-gam_cv_metrics %>%
  filter(.metric == "rmse") %>%
  pull(.estimate)

# Create a combined table
model_comparison <- data.frame(
  Model = c("Simple Linear", "Multivariate Linear", "Multivariate Linear Cross-validated", "Lasso without Tuning", "Lasso - Tuned and Cross-validated", "GAM model", "GAM Cross-validated"),
  RMSE = c(slr_rmse, mlr_rmse, mlr_cv_rmse, lasso_rmse, lasso_cv_rmse, gam_rmse, gam_cv_rmse)
)

file_path = here("results","tables", "RMSE_comparison.rds")
saveRDS(model_comparison, file_path)
print(model_comparison)

```

We can see that the simple GAM model has the lowest RMSE, suggesting it performs well in prediction accuracy. However, when subjected to cross-validation, the GAM model's RMSE reveal a decrease in prediction accuracy across various subsets of the data compared to its initial evaluation based solely on the training data. Both the tuned/cross-validated and untuned Lasso models demonstrate the next lowest RMSE values. For simplicity, we choose the untuned Lasso model as our final selection.

Finally, we evaluate the performance of the selected model i.e. simple Lasso model by fitting it on the testing set.

```{r}
#Computing model predictions
lasso_test_predicts <- predict(lasso_fit, new_data = test_data)

#augment to evaluate performance metric
lasso_test_aug <- augment(lasso_fit, test_data)
lasso_test_aug %>% select(sum_purchase,.pred)

#Model evaluation: Calculatin of RMSE
lasso_test_metrics <- lasso_test_aug %>% metrics(truth=sum_purchase, .pred)

file_path = here("results","tables", "RMSE_Test.rds")
saveRDS(lasso_test_metrics, file_path)

lasso_test_metrics

```

We see that the RMSE of the model on test data is slightly higher at 683.87 compared to that of 675.38 on the train set.

Now lets plot Residuals vs. Predicted values.

```{r}
lasso_test_aug<- lasso_test_aug %>%
  mutate(residuals = .pred - sum_purchase)

plot_residual<- ggplot(lasso_test_aug, aes(x=.pred, y=residuals))+
  geom_point(size=1)+
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +  
  labs(x = "Predicted Values", y = "Residuals") +  
  theme_bw()

plot_residual

figure_file = here("results","figures","Residual.png")
ggsave(filename = figure_file, plot=plot_residual)

```

The residual plot exhibits a fan shaped pattern. Residuals are not evenly spread around the horizontal axis indicating the presence of heteroscedasticity. Higher values seem to have more variability in prediction errors than lower values. This can be indication of other factors that influence the outcome variable which are missing in the model.

Next we plot the predicted values against the observed values for both the train and the test set.

```{r}
#creating an object for outcome variable sum_purchase from the training data
Observed_train <- train_data$sum_purchase

#creating an object for outcome variable sum_purchase from the test data
Observed_test <- test_data$sum_purchase

#Creating a new data frame containing columns for observed and predicted values of the train set 
lasso_train <- data.frame(Observed= Observed_train, Predicted = lasso_aug$.pred, Model="Train")

#Creating a new data frame containing columns for observed and predicted values of the Test set
lasso_test<- data.frame(Observed = Observed_test, Predicted = lasso_test_aug$.pred, Model="Test")

#Combining all the Observed and predicted values of the train and test sets by rows to create a long format data
lasso_combined <- rbind(lasso_train, lasso_test)

#Plotting of combined predicted vs observed data for the train data and test data

p4<-ggplot(lasso_combined, aes(x=Observed, y=Predicted, color=Model))+
  geom_point()+ 
  scale_color_manual(values = c("Train"="blue", "Test"="red")) + 
  geom_abline(intercept = 0, slope = 1, linetype = "solid", color="black") + #45 degree line
   labs(x= "Observed Values", y ="Predicted Values", title = "Predicted vs. Observed Values")+
  theme_minimal() # Use a minimal theme
p4
figure_file = here("results","figures","Predicted_Observed.png")
ggsave(filename = figure_file, plot=p4)
```

For the lower values, both the train and test data seems cluster around the diagonal line indicating that the model predict the lower values well for both the data.However, both the data spreads away from the line of perfect fit as the value increases suggesting that the model is less accurate for higher values. The test data exhibits a wider spread for higher values than the train data suggesting that the model does not generalize well on the unseen data. Overall, while the model exhibits a reasonable predictive ability for observations with lower values, the increased spread among higher values and the differences between the training and test sets indicate that there may be room for improvement. There may be other factors that are instrumental in predicting the outcome variable.

### 2024-04-22 R/eda-code/eda_superstore.qmd
Project Part 5

@@
 ---
+The purpose of this task is to perform exploratory analysis of the superstore data.
 
@@
 library(gridExtra)
-library(webshot2) #Capture as an Image the gttable before converting it to pdf
 library(zoo) #To plot seasonality of sales
+library(scales)
 ```
@@
 #Path to data using here() package
-data_location <- here::here("data","processed-data","processed_superstore_RFM.rds")
+
+data_location <- here::here("data","processed-data","processed_superstore.rds")
 #load data
-mydata <- readRDS(data_location)
+superstore <- readRDS(data_location)
 ```
@@
 ```{r}
-summary_df = skimr::skim(mydata)
+summary_df = skimr::skim(superstore)
 print(summary_df)
@@
 ```
+The data summary reveals that there were 14951 transactions to 4117 different customers during 2010-12-01 and	2011-12-09.
 
 # Descriptive Statistics tables
+The processed data accounts sales of a particular item as an observation.
+This means different items sold in the same Invoice appears as separate observations and it is most likely that each Invoice number appeared in separate observations.
+For descriptive analysis, we need to combine the sales under one invoice as a single observation.
 
-A meaningful analysis of customer spending behavior require the Invoice Sales Amount, total purchase by a Customer during a period, frequency of purchase by a customer during a period or a combination of those.
-A table is created which provides the descriptive statistics of sales by Invoice value, Customer purchase and frequency of customer purchase.
-This table reveals that there were 1402 transactions to 399 different customers during 2010-12-01 and	2011-12-09.
+```{r}
+#Creating data frame where Sales under a unique Invoice No. is bundled as a single observation
+store_InvNo <- superstore %>%
+  #grouping the data by the unique InvoiceNo
+  group_by(InvoiceNo) %>%
+  summarise(InvoiceValue = sum(Sales, na.rm = TRUE), #sums all the sales under the unique InvoiceNo
+            InvoiceDate = first(InvoiceDate),  #Selects the first Invoice Date for the data grouped by InvoiceNo 
+            CustomerID = first(CustomerID), #Selects the first CustomerID for the data grouped by InvoiceNo
+            Country = first(Country)) %>%   ##Selects the first Country for the data grouped by InvoiceNo
+select(InvoiceDate, InvoiceNo, CustomerID, InvoiceValue, Country)
+str(store_InvNo)
+skimr::skim(store_InvNo)
+```
+
+For further exploratory analysis, we require customer spending behavior indicators, Recency, Frequency and Monetary.
+Recency is the number of days since the last purchase by a customer.
+Frequency is the no.
+of purchases by a customer over a given period.
+Monetary is the total amount of purchase a customer has made over a period.
+
+For determining recency of a transaction, a reference date past the study period is required.
+The data covers transactions till 2011-12-09.
+The next day i.e.
+2011-12-10 is considered as the reference date to measure recency.
+
+```{r}
+#fixing the reference date for recency
+ref_date <- as.Date("2011-12-10")
+
+#Creating a data frame that includes Recency, frequency and monetary of customer purchase.
+store_CustomerID <- superstore %>%
+  group_by(CustomerID) %>%
+  summarise(Frequency = n_distinct(InvoiceNo),
+            Recency = as.numeric(ref_date - as.Date(max(InvoiceDate))),
+            Monetary=sum(Sales))%>%
+  ungroup()
+
+```
+A table is created which provides the descriptive statistics of sales by Invoice value, Customer purchase, frequency and recency of customer purchase.
 
@@
 # Aggregate total sales by InvoiceNo for calculating mean, standard deviation, minimum and maximum
-Sales_by_Invoice <- mydata %>%
+Sales_by_Invoice <- store_InvNo %>%
     summarise(
+    Count = n(),
     Particulars = "Invoice Value",
-    Count = n(),
     Mean = round(mean(InvoiceValue, na.rm = TRUE), 2),
@@
 # Aggregate total sales by CustomerID to calculate mean, standard deviation, minimum and maximum
-TotalSales_customer <- mydata %>%
-  group_by(CustomerID) %>%
-  summarise(TotalSales = sum(InvoiceValue, na.rm = TRUE)) %>%
-  ungroup() %>%
+TotalSales_customer <-store_CustomerID%>%
     summarise(
-    Particulars = "Cumulative Sales to Individual Customers",
     Count = n(),
-    Mean = round(mean(TotalSales, na.rm = TRUE), 2),
-    SD = round(sd(TotalSales, na.rm = TRUE), 2),
-    Min = round(min(TotalSales, na.rm = TRUE), 2),
-    Max = round(max(TotalSales, na.rm = TRUE), 2)
+    Particulars = "Cumulative Purchase",
+    Mean = round(mean(Monetary, na.rm = TRUE), 2),
+    SD = round(sd(Monetary, na.rm = TRUE), 2),
+    Min = round(min(Monetary, na.rm = TRUE), 2),
+    Max = round(max(Monetary, na.rm = TRUE), 2)
   )
@@
 # Frequency of sales per CustomerID (i.e., the count of unique InvoiceNo per CustomerID)for calculating mean, standard deviation, minimum and maximum
-
-sales_frequency_per_customer <- mydata %>%
-  group_by(CustomerID) %>%
-  summarise(FrequencyOfSales = n_distinct(InvoiceNo)) %>%
-  ungroup() %>%
+sales_frequency_per_customer <-store_CustomerID%>%
     summarise(
     Count = n(),
-    Mean = round(mean(FrequencyOfSales, na.rm = TRUE), 2),
-    SD = round(sd(FrequencyOfSales, na.rm = TRUE), 2),
-    Min = round(min(FrequencyOfSales, na.rm = TRUE), 2),
-    Max = round(max(FrequencyOfSales, na.rm = TRUE), 2)
+    Mean = round(mean(Frequency, na.rm = TRUE), 2),
+    SD = round(sd(Frequency, na.rm = TRUE), 2),
+    Min = round(min(Frequency, na.rm = TRUE), 2),
+    Max = round(max(Frequency, na.rm = TRUE), 2)
   ) %>%
-  mutate(Particulars = "Sales frequency to Individual Customers")
+  mutate(Particulars = "Purchase frequency")
 
+#How Recent a customer bought
+Sales_Recency<- store_CustomerID%>%
+  summarise(
+    Count = n(),
+    Particulars = "Recency of Purchase",
+    Mean = round(mean(Recency, na.rm = TRUE), 2),
+    SD = round(sd(Recency, na.rm = TRUE), 2),
+    Min = round(min(Recency, na.rm = TRUE), 2),
+    Max = round(max(Recency, na.rm = TRUE), 2)
+  )
 # Combine the statistics into a single data frame
-combined_stats <- bind_rows(Sales_by_Invoice, TotalSales_customer, sales_frequency_per_customer)
+combined_stats <- bind_rows(Sales_by_Invoice, TotalSales_customer, sales_frequency_per_customer, Sales_Recency)
 
 combined_stats
-
-#Saving this statistics in the results folder.
 
@@
 
-Next, density plots were created for Sales across customers and Invoice Values.
-Both figures are heavily right-skewed.The averages of total sales to a customer and total invoice values appear to be less than 2500 and 1250 sterling pounds respectively supporting the mean values of 2025 and 563 in the descriptive table.
-
-A histogram of purchase frequency of each customer is created by grouping each customer by invoice no.The average frequency appears to be less than 5 times for most of the customers which is also observed in the descriptive statistics table.
+Next, histograms were created for Sales across customers and Invoice Values. Histograms were also created for purchase frequency and recency of customers.
 
 ```{r}
-# Aggregate sales by CustomerID
-TotalSales_customer <- mydata %>%
-  group_by(CustomerID) %>%
-  summarise(TotalSales = sum(InvoiceValue, na.rm = TRUE)) %>%
-  arrange(desc(TotalSales))
-
-#Density Plot of Sales per Customer
-# Assuming sales_per_customer is already calculated
-p1 <- ggplot(TotalSales_customer, aes(x = TotalSales)) +
-  geom_density(fill = "skyblue", alpha = 0.5) +
-  labs(title = "Density Plot of Sales across Customers", x = "Total Sales", y = "Density")
+#Histogram of Total Sales per Customer
+p1 <- ggplot(store_CustomerID, aes(x = Monetary)) +
+  geom_histogram(binwidth = 100,fill = "green", color = "skyblue") +
+  labs(title = "Histogram of Customer's Total Purchases", x = "Total Sales", y = "Count")
 
 # Histogram of Sales by Invoice values
-
-p2 <- ggplot(mydata, aes(x = InvoiceValue)) +
-  geom_density(fill = "lightgreen", alpha = 0.5) +
-  labs(title = "Density Plot of  Invoice Sales",
-       x = "Invoice Value", 
-       y = "Density") +
+p2 <- ggplot(store_InvNo, aes(x = InvoiceValue)) +
+  geom_histogram(binwidth = 20, fill = "lightgreen", color = "darkblue") +
+  labs(title = "Histogram of  Invoice Value", x = "Invoice Value", y = "Count") +
   theme_minimal()
 
-# Calculate purchase frequency per customer
-purchase_frequency <- mydata %>%
-  distinct(CustomerID, InvoiceNo) %>% # Ensure unique InvoiceNo per CustomerID
-  count(CustomerID, name = "PurchaseFrequency") # Count the number of invoices per customer
-library(ggplot2)
-
 # Histogram of purchase frequencies
-p3 <- ggplot(purchase_frequency, aes(x = PurchaseFrequency)) +
+p3 <- ggplot(store_CustomerID, aes(x = Frequency)) +
   geom_histogram(binwidth = 1, fill = "salmon", color = "darkblue") + # Adjust binwidth as needed
@@
        x = "No. of Purchases", 
+       y = "Count of Customers") +
+  theme_minimal()
+# Histogram of purchase recencies
+p4 <- ggplot(store_CustomerID, aes(x = Recency)) +
+  geom_histogram(binwidth = 10, fill = "coral", color = "darkblue") + # Adjust binwidth as needed
+  labs(title = "Histogram of Customer Purchase Recency",
+       x = "Days since last Purchase", 
        y = "Count of Customers") +
@@
 p3 <- p3 + theme(plot.title = element_text(size = 11)) # Adjusting font size of the title of p3
+# Modify p4 to adjust title font size
+p4 <- p4 + theme(plot.title = element_text(size = 11)) # Adjusting font size of the title of p3
 
-plot1<-grid.arrange(p1, p2, p3, ncol = 2, nrow=2)
+plot1<-grid.arrange(p1, p2, p3,p4, ncol = 2, nrow=2)
 
@@
 ```
-
-Next, total purchase by a customer was plotted against purchase frequency.A few outliers were noted.
+All histograms exhibited heavily right-skewness.
+The histogram of Customer's Total Purchases indicated majority of the customers are associated with lower sales amount, typical of individual consumer transactions.
+Histogram of Invoice value aligns with this finding by showing that most invoices had values consistent with retail purchases.
+Histogram of Customer Purchase Frequencies indicated that most customers engage in a number of transactions typical for retail shopping behavior.
+Histogram of Customer Purchase Recency exhibited a higher frequency of recent purchases, which suggest regular shopping habits expected in retail context.
+ed.
 
@@
 #| warning: false
-# Calculate purchase frequency and total invoice value per customer
-customer_stats <- mydata %>%
-  group_by(CustomerID) %>%
-  summarise(PurchaseFrequency = n_distinct(InvoiceNo),
-            TotalInvoiceValue = sum(InvoiceValue, na.rm = TRUE)) %>%
-  ungroup()  # Ensure the data is ungrouped for further analysis
 
 # Scatter plot of Total Invoice Value by Purchase Frequency
-p4 <- ggplot(customer_stats, aes(x = PurchaseFrequency, y = TotalInvoiceValue)) +
-  geom_point(alpha = 0.5) +  # Adjust alpha for point transparency, if needed
+p5 <- ggplot(store_CustomerID, aes(x = Frequency, y = Monetary)) +
+  geom_point(alpha = 0.5) +  
   geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add a linear regression line without standard error
@@
        x = "Purchase Frequency",
-       y = "Total Invoice Value") +
+       y = "Cummulative Invoice Value") +
   theme_minimal()
-plot(p4)
+p5
 
-figure_file = here("results","figures","Sales_trend.png")
-ggsave(filename = figure_file, plot=p4)
+figure_file = here("results","figures","MonetarybyFreq_distribution.png")
+ggsave(filename = figure_file, plot=p5)
 
 ```
+The scatter plot visualizing Total Invoice Value by Purchase Frequency indicates a positive correlation between the two components.
+This suggests that as the frequency of purchases by a customer increases, so does their total amount spent.
+Such a finding is typical in retail scenarios where regular customers often contribute more to overall sales than infrequent ones.
+This trend underscores the importance of strategies aimed at increasing customer engagement and purchase frequency as a means to boost revenue.
 
-Next, Bar chart is created for Sales Distribution by Country. The top five sales are in Germany, France, Spain, Switzerland and Japan.
-
-```{r}
-p5 <- mydata %>% 
-  group_by(Country) %>% 
-  summarise(TotalSales = sum(InvoiceValue)) %>%
-  ggplot(aes(x = reorder(Country, TotalSales), y = TotalSales, fill = Country)) + 
-  geom_bar(stat = "identity") + 
-  coord_flip() + 
-  theme(legend.position = "none")
-
-plot(p5)
-
-figure_file = here("results","figures","Countrywise_Sales.png")
-ggsave(filename = figure_file, plot=p5)
-```
-As the store sells gift goods, an inspection is required to check the seasonality of sales.
-First, monthly total sales is calculated aggregating Invoice Value by months of the Invoice Date.The aggregated sales is then plotted to visualize seasonal trend which depicted a spike in the sales during the last few months of the year.
+As the store is in gift selling business, an inspection is required to check the seasonality of sales.
+First, monthly total sales is calculated aggregating Invoice Value by months of the Invoice Date.The aggregated sales is then plotted to visualize seasonal trend.
 
@@
 #Aggregating Sales by months
-mydata_aggregated <- mydata %>%
+superstore_aggregated <- store_InvNo %>%
   group_by(InvoiceDate = as.yearmon(InvoiceDate)) %>%
@@
 # Formatting Date for plotting purposes
-mydata_aggregated$InvoiceDate <- as.Date(as.yearmon(mydata_aggregated$InvoiceDate))
+superstore_aggregated$InvoiceDate <- as.Date(as.yearmon(superstore_aggregated$InvoiceDate))
 
-p6<-ggplot(mydata_aggregated, aes(x = InvoiceDate, y = TotalValue)) +
+p6<-ggplot(superstore_aggregated, aes(x = InvoiceDate, y = TotalValue)) +
   geom_line() + 
-  scale_x_date(date_labels = "%b", date_breaks = "1 month") + # Adjust formatting as needed
+  scale_x_date(date_labels = "%b", date_breaks = "1 month") + 
+  scale_y_continuous(labels = label_number()) + # formats the y-axis labels as full numbers
   labs(title = "Monthly Sales", x = "Month", y = "Sales Value") +
   theme_minimal()
+
 p6
@@
 ```
+The data spans from 2010-12-01 and	2011-12-09, revealing a good growth in the December of 2011 compared to the same month previous year.
+The monthly sales exhibits noteworthy seasonal variation.
+A sharp peak in May suggests a strong spring season and the plateau from June to August indicates stability in consumer purchases.
+Starting in October, the sales trajectory showed an upward trend, reaching maximum in November/December, which could be due to early holiday shopping influenced by Black Friday and the approaching Christmas season.
+It is reasonable to anticipate that past customers will engage in purchasing activities during these peak sales periods.
+
+To gain further insights, an examination of the recency by frequency plot is carried out.
+This will help to understand the relationship between the recency of purchases and the frequency of customer transactions.
+
+```{r}
+# Scatter plot of Recency by Frequency
+p7 <- ggplot(store_CustomerID, aes(x = Frequency, y = Recency)) +
+  geom_point(alpha = 0.5) +  
+  labs(title = "Recency by Frequency Plot",
+       x = "Purchase Frequency",
+       y = "No. of days since Last Purchase") +
+  theme_minimal()
+p7
+
+figure_file = here("results","figures","Recency_by_Frequency.png")
+ggsave(filename = figure_file, plot=p7)
+```
+While the Recency by Frequency plot revealed an expected pattern where customers with more frequent purchases tended to have more recent transactions, an unexpected trend was also noted.
+There are some evidences that a segment of customers with a history of multiple purchases did not return during the peak sales period.
+This divergence from the usual retail pattern suggests the need of further investigation.
+
+To complement this analysis, another chart is created to examine whether there is a correlation between the recency of purchases and the monetary value of those purchases.
+
+```{r}
+# Scatter plot of Recency by Frequency
+p8 <- ggplot(store_CustomerID, aes(x = Monetary, y = Recency)) +
+  geom_point(alpha = 0.5) +  
+  labs(title = "Recency by Monetary plot",
+       x = "Monetary",
+       y = "No. of days since Last Purchase") +
+  theme_minimal()
+p8
+
+figure_file = here("results","figures","Recency_by_Monetary.png")
+ggsave(filename = figure_file, plot=p8)
+```
+The Recency by Monetary plot mirrored the trend observed in the 'Recency by Frequency' plot.
+The plot revealed that a subset of customers who have purchased over 10,000 in monetary value, did not make subsequent purchases within the following 100 days.
+
+The store has customer base inside and outside of the UK. To visualize the sales distribution by countries, a bar chart is plotted.
+
+```{r}
+p9 <- store_InvNo %>% 
+  group_by(Country) %>% 
+  summarise(TotalSales = sum(InvoiceValue)) %>%
+  ggplot(aes(x = reorder(Country, TotalSales), y = TotalSales, fill = Country)) + 
+  geom_bar(stat = "identity") +
+  scale_y_continuous(labels = label_number())+
+  coord_flip() + 
+  theme(legend.position = "none")
+
+plot(p9)
+
+figure_file = here("results","figures","Countrywise_Sales.png")
+ggsave(filename = figure_file, plot=p9)
+```
+As exhibited, majority of the sales is in UK followed by Germany and France.
+
+To better understand the customer base and identify the most loyal shoppers, the project will focus on customers who have shopped with the store in two distinct periods: the first half of the year and the peak sales season.
+As revealed by the monthly sales trend, sales during first six months of the store were comparatively slower than the later period of the year.
+The first half of the year also typically doesn't have big festive drives to attract the buyers.
+Customers who made purchases during this time might be doing so based on the everyday value and appeal of the products, free from the influence of holiday discount or promotions.
+During the peak sales season, the shopping activity naturally increases due to holidays surging the store's sales.
+On the other hand, competition among retailers intensifies, with promotions proliferating across the market and challenging the customer retention.
+Amidst this competition, customers who chose to return to the same store for their holiday purchases potentially are the ones who shop with the store regularly, regardless of festive season or deal offers.
+By looking at customers who have made purchases in both of these times, a group that is valuable round the year can be identified.
+Such group potentially keep steady sales in the slower months and boost the numbers during the busy times.
+Understanding this group can help focus the marketing efforts, improve the customer service, and ultimately, keep the business strong.
+
+The analysis part of the project is designed to quantify the contributions made by this segment of customers to the store's overall business.
+Such data-driven understanding is anticipated to help marketing initiatives and customer service enhancements in order to retain loyal customer base.The analysis can be found at R>analysis-code-statistical_analysis.qmd

### 2024-04-22 R/processing-code/processing_superstore.qmd
Project Part 5

--- 2024-03-07 R/processing-code/processing_superstore-v1.qmd
@@
 ---
+
+The purpose of this project is to analyze the impact of customer loyalty on business of a non-store online retail by examining the contribution in revenue by customers who returned to purchase during the holiday season.
+It uses a transnational data set 'superstoredata' spanning from December 1, 2010, to December 9, 2011 from a UK-based, non-store online retail operation.
+The company specializes in unique all-occasion gifts to a clientele which also includes large wholesalers.This study exclude larger sales which are indicative of sales to larger wholesales, from the analysis to avoid potential bias.
+
+The raw data accounts sales of a particular item as an observation.
+This means different items sold in the same Invoice appears as separate observations and it is most likely that each Invoice number appeared in separate observations.
+
 # Setup
@@
 ```{r}
+#Checking data summaries
 dplyr::glimpse(raw_data)
@@
 # Cleaning
-
-The purpose of this project is to study purchase pattern of customers residing outside of UK.The data set data_without_uk excludes all sales in the UK.
-It has 46431 observations with 9 variables.
-It is important to note that this data set accounts sales of a particular item as an observation.
-This means different items sold in the same Invoice appears as separate observations and it is most likely that an Invoice number could appear in separate observations.
-
-```{r}
-# Exclude "United Kingdom" from the data set
-data_without_uk <- subset(raw_data, Country != "United Kingdom")
-skimr::skim(data_without_uk)
-```
+The data set has 541909 observations with 9 variables.Based on the unique Invoice No., and Description of sales, the store made 25900 transactions of 4224 types of merchandise to its clients in 38 countries during the period.
 
@@
 
-First, there are 1480 missing values in CustomerID.
+First, there are 135080 missing values in CustomerID.
 Customer behavior is a crucial part of this analysis which cannot be tracked without the unique customerID.
@@
 Further, CustomerID is coded as a numerical variable in the original data and it needs to be converted to factor variable.
-Country, Description and StockCode are also converted to factors from character variables.
+InvoiceNo, Country, Description and StockCode are also converted to factors from character variables and InvoiceDate converted to Date format.
 
@@
 # Subset the data by excluding rows where CustomerID is NA
-  clean_data_step1 <- data_without_uk %>%
+  clean_data_step1 <- raw_data %>%
   filter(!is.na(CustomerID)) %>%   #excludes observation without CustomerID
@@
          Description = as.factor(Description), #converts Description to a factor
-         StockCode = as.factor(StockCode)) #converts StockCode to a factor
+         StockCode = as.factor(StockCode),  #converts StockCode to a factor
+        InvoiceDate = as.POSIXct(InvoiceDate, format = "%m/%d/%Y") ) #converts to date format from character
 
@@
 There are negative values in quantity and sales due to order cancellation.
-An inspection is carried out to check the number of such observations are in the data set.
-It is found that there are 1372 observations with negative Quantity and Sales.
+An inspection is carried out to check the number of such observations in the data set.
+It is found that there are 8905 observations with negative Quantity and Sales.
 
@@
 ```
-
-Next step is to find matching original sales entries corresponding to the negative entries based on CustomerID, StockCode, Quantity and Sales figures.
-1372 matches were found, which is equal to the no.
-of observations with negative Quantity and Sales figures.
+Observations with negative Quantity and Sales are dropped. This brought down the total number of observations to 397884 for total transactions of 18532 of 3877 products with 4338 different customers.
 
 ```{r}
-positive_sales <- clean_data_step1 %>%
-  filter(Quantity > 0 & Sales > 0)
-count(positive_sales)
-
-#Checking for matches
-matched_sales <- negative_sales %>%
-  rowwise() %>%
-  mutate(match_found = any(StockCode %in% positive_sales$StockCode &
-                             CustomerID %in% positive_sales$CustomerID))
-
-count(matched_sales)
-```
-
-#The next attempt is to match returns to their original sales and get rid of both matched transactions.
-However, this needs further work as the below codes are resulting 582 observations in original sales instead of 1372.
-
-```{r}
-# Add an identifier to each row in both data sets to track them (Marked as comment because this chunk of code needs further work and also takes a longer time to run)
-
-"negative_sales <- negative_sales %>% mutate(neg_id = row_number())
-positive_sales <- positive_sales %>% mutate(pos_id = row_number())
-
-# Initialize a vector to keep track of matched negative_sales' IDs
-matched_neg_ids <- integer(0)
-
-# Initialize an empty data frame to store matches
-matches <- data.frame(neg_id = integer(), pos_id = integer())
-
-# Iterate over positive_sales to find matches in  (Marked as comment because it takes a longer time to run this code)
-for (pos_row in 1:nrow(positive_sales)) {
-  pos_sale <- positive_sales[pos_row, ]
-
-  # Attempt to match with negative_sales, excluding already matched rows
-  potential_matches <- negative_sales %>%
-    filter(
-      !neg_id %in% matched_neg_ids, # Exclude already matched ones
-      CustomerID == pos_sale$CustomerID,
-      StockCode == pos_sale$StockCode,
-      Quantity == -pos_sale$Quantity, # Matching negative of Quantity
-      Sales == -pos_sale$Sales # Matching negative of Sales
-    )
-
-  if (nrow(potential_matches) > 0) {
-    # If a match is found, take the first one
-    first_match <- potential_matches[1, ]
-    matched_neg_ids <- c(matched_neg_ids, first_match$neg_id)
-
-    # Record the match
-    matches <- rbind(matches, data.frame(neg_id = first_match$neg_id, pos_id = pos_sale$pos_id))
-  }
-}
-
-# Extract matched rows based on IDs 
-matched_negative_sales <- negative_sales %>% filter(neg_id %in% matches$neg_id)
-matched_positive_sales <- positive_sales %>% filter(pos_id %in% matches$pos_id)
-print(matched_negative_sales)
-print(matched_positive_sales)"
-```
-
-Observations with negative Quantity and Sales are dropped though unsuccessful in finding exact no. of matching original transactions. This brought down the total number of observations to 43563.
-
-```{r}
+#Creating a new dataset that includes only Quantity and Sales with positive values
 clean_data_step2 <- clean_data_step1 %>%
@@
 ```
-
-Inspection is required in the distribution of Quantity, unit price and Sales, the higher values of these variable are concentrated at the higher percentiles.
-I focused on distribution of UnitPrice at small intervals from 75% to 100%.
+The summary above indicated that higher values of Quantity, Unit price and sales are concentrated in the 75-100 percentile range. We need to look deeper into these variables at the higher percentiles.
 
 ```{r}
-# Calculate higher percentiles within the 3rd quartile to max range
+# Decomposing further the Unit Price at 75-100 percentiles 
 percentiles <- quantile(clean_data_step2$UnitPrice, probs = c(0.75, 0.90, 0.95, 0.99, 0.995, 0.996, 0.997, 0.998, 0.999, 1))
@@
 Observations wherein UnitPrice is above 25 are filtered for further examination.
-It resulted into 433 observations.
+It resulted into 908 observations.
 
@@
 
-The components of description was checked using the unique function which resulted into 2903 levels for 433 observations.
+The components of description was checked using the unique function which resulted into 3896 levels for 908 observations.
 With the help of ChatGPT, it was found that in R, when a subset of a data frame is created, the factor levels in the subset are not automatically dropped even if they are not present in the subset.
@@
 ```{r}
-#Attempt without dropping the unused factor levels
+#Attempt to find unique categories of stocks with unit price above 25 without dropping the unused factor levels
 unique_categories_initial <- unique(UnitPrice_above25$Description)
@@
 
-# Finding the unique categories again
+# Finding again the unique categories of stocks with unit price above 25 without dropping the unused factor levels
 unique_categories_final <- unique(UnitPrice_above25$Description)
@@
 
-The above code resulted into 19 categories in product description which have the Unitprice above 25.
-Some of the high UnitPrice are related to the descriptions such as CARRIAGE, POSTAGE, Manual, appearing to be related to logistics.
-1216 observations with such descriptions were dropped from the data since those are less likely to be directly related to customer purchase.
+The above code resulted into 36 categories in product description which have the Unitprice above 25.
+Some of the high UnitPrice are related to the descriptions such as CARRIAGE, POSTAGE, Manual, DOTCOM POSTAGE, Next Day Carriage, each appearing to be related to logistics.
+1673 observations with such descriptions were dropped from the data since those are less likely to be directly related to customer purchase.
 The no.
-of observations came down to 42347.
+of observations came down to 396211.
 
@@
                                 !grepl("POSTAGE", clean_data_step2$Description) &
-                                !grepl("CARRIAGE", clean_data_step2$Description), ]
+                                !grepl("CARRIAGE", clean_data_step2$Description) &
+                                !grepl("DOTCOM POSTAGE", clean_data_step2$Description) &
+                                !grepl("Next Day Carriage", clean_data_step2$Description)  , ]
 
@@
 ```
-Next the InovoiceDate is in character format, and needs to be converted to date format.
-
-```{r}
-clean_data_step3$InvoiceDate <- as.POSIXct(clean_data_step3$InvoiceDate, format = "%m/%d/%Y %H:%M")
-
-skimr::skim(clean_data_step3)
-
-```
-
-The store also sells merchandises in wholesale quantities.
-As this study focuses on end user purchases, observations of sales to wholesales need to be disregarded.
+The store also sells merchandises in large quantities to wholesalers.
+The lack of explicit indicators distinguishing sales to wholesalers prompted an analysis based on purchase frequency and total invoice value.
 A wholesaler potentially purchases in big volumes and in higher frequencies.
@@
        x = "Purchase Frequency",
-       y = "Total Invoice Value") +
+       y = "Total Purchase") +
   scale_y_continuous(labels = label_comma()) +  # This line changes the y-axis labels from exp. number to actual value
@@
 
-The scatter plot reveals that most of the customers made purchases less than 50 times during that period and most of the cumulative purchase values were less than 50,000.
-Assuming that the outliers represent purchases by wholesalers, observations with purchase frequencies above 50 and cumulative Invoice value above 50,000 are dropped from the data.
+The scatter plot reveals that most of the customers made purchases less than 50 times during that period and most of the cumulative purchase values were less than 25,000.
+Assuming that the outliers represent purchases by wholesalers, observations with purchase frequencies above 50 and cumulative Invoice value above 25,000 are dropped from the data.
 
-This involves first sub-setting customer_stats such that it includes only those observations where PurchaseFrequency <= 50 and TotalInvoiceValue <=50000 and finally, merging this sub-set with the original dataset (clean_data_step3).
+This involves first sub-setting customer_stats such that it includes only those observations where PurchaseFrequency <= 50 and TotalInvoiceValue <=25000 and finally, merging this sub-set with the original dataset (clean_data_step3).
 
@@
 filtered_customer_stats <- customer_stats %>%
-  filter(PurchaseFrequency <= 50, TotalInvoiceValue <= 50000)
+  filter(PurchaseFrequency <= 50, TotalInvoiceValue <= 25000)
 
-# Excluding purchase frequencies above 50 and cumulative invoice values above 50000 from the data
+# Excluding purchase frequencies above 50 and cumulative invoice values above 25000 from the data
 clean_data_step4 <- clean_data_step3 %>%
@@
 
-This step excluded 9745 observations from the data for a final total observation of 32602. A similar scatter plot is plotted for the new data.
+This step excluded 45216 observations from the data for a final total observation of 350995. A similar scatter plot is plotted for the new data.
 
@@
 
+The wholesalers are expected to purchase in large quantity. Additional inspection is carried out to identify wholesalers by quantity purchased and individual Invoice Amount by plotting a scatter plot.
+
+```{r}
+# Calculate purchase quantity and value per transaction represented by InvoiceNo 
+customer_stats5 <- clean_data_step4 %>%
+  group_by(InvoiceNo) %>%
+  summarise(InvoiceQty = sum(Quantity, na.rm = TRUE),
+            InvoiceValue = sum(Sales, na.rm = TRUE)) %>%
+  ungroup()  # Ensure the data is ungrouped for further analysis
+
+# Scatter plot of Total Quantity purchased by Invoice Amount
+p3 <- ggplot(customer_stats5, aes(x = InvoiceValue, y = InvoiceQty)) +
+  geom_point(alpha = 0.5) +  # Adjust alpha for point transparency, if needed
+  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add a linear regression line without standard error
+  labs(title = "Total Invoice Value by Quantity Purchased",
+       x = "Invoice Amount",
+       y = "Quantity Purchased") +
+  scale_y_continuous(labels = label_comma()) +  # This line changes the y-axis labels
+
+    theme_minimal()
+plot(p3)
+skimr::skim(customer_stats5)
+
+```
+The scatter plot reveals that most of the customers made purchases less than 1000 items and most of the invoice values were less than 2,500.
+Assuming that the sparsely distributed points represent purchases by wholesalers, observations with Quantity above 1000 and individual Invoice value above 2,500 are removed from the data.
+
+This involves first sub-setting customer_stats5 such that it includes only those observations where InvoiceQty <= 1000 and InvoiceValue <=2500 and finally, merging this sub-set with the original dataset (clean_data_step4).
+
+```{r}
+# Filtering the customer_stats5 to meet the conditions
+filtered_customer_stats5 <- customer_stats5 %>%
+  filter(InvoiceQty <= 1000, InvoiceValue <= 2500)
+
+# Excluding purchase quantities above 1000 and individual invoice values above 2500 from the data
+clean_data_step5 <- clean_data_step4 %>%
+  semi_join(filtered_customer_stats5, by='InvoiceNo')
+
+skimr::skim(clean_data_step5)
+```
+This step excluded 12897 observations from the data for a final total observation of 338175. A similar scatter plot is plotted for the new data.
+
+```{r}
+# Calculate purchase quantity and amount per invoice
+customer_stats6 <- clean_data_step5 %>%
+  group_by(InvoiceNo) %>%
+  summarise(InvoiceQty = sum(Quantity, na.rm = TRUE),
+            InvoiceValue = sum(Sales, na.rm = TRUE)) %>%
+  ungroup()  # Ensure the data is ungrouped for further analysis
+
+# Scatter plot of Purchase Quantity by Invoice Value
+p4 <- ggplot(customer_stats6, aes(x = InvoiceValue, y = InvoiceQty)) +
+  geom_point(alpha = 0.5) +  # Adjust alpha for point transparency, if needed
+  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add a linear regression line without standard error
+  labs(title = "Purchase Quantity by Invoice Value",
+       x = "Invoice Amount",
+       y = "Quantity Purchased") +
+  scale_y_continuous(labels = label_comma()) +  # This line changes the y-axis labels
+
+    theme_minimal()
+plot(p4)
+
+```
+
+```{r}
+
+```
+
+The scatter plot of the new data further pointed out that the distribution is densed below quantity of 500 and amount of 1000.I cleaned the data further in an additional attempt to remove purchases that are likely to be from wholesalers.
+
+```{r}
+# Filtering the customer_stats5 to meet the conditions
+filtered_customer_stats5 <- customer_stats5 %>%
+  filter(InvoiceQty <= 500, InvoiceValue <= 1000)
+
+# Excluding purchase quantities above 500 and individual invoice values above 1000 from the data
+clean_data_step5 <- clean_data_step4 %>%
+  semi_join(filtered_customer_stats5, by='InvoiceNo')
+
+skimr::skim(clean_data_step5)
+```
+This step excluded  observations from the data for a final total observation of 289132. A similar scatter plot is plotted for the new data.
+
+```{r}
+# Calculate purchase quantity and amount per invoice
+customer_stats6 <- clean_data_step5 %>%
+  group_by(InvoiceNo) %>%
+  summarise(InvoiceQty = sum(Quantity, na.rm = TRUE),
+            InvoiceValue = sum(Sales, na.rm = TRUE)) %>%
+  ungroup()  # Ensure the data is ungrouped for further analysis
+
+# Scatter plot of Purchase Quantity by Invoice Value
+p4 <- ggplot(customer_stats6, aes(x = InvoiceValue, y = InvoiceQty)) +
+  geom_point(alpha = 0.5) +  # Adjust alpha for point transparency, if needed
+  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add a linear regression line without standard error
+  labs(title = "Purchase Quantity by Invoice Value",
+       x = "Invoice Amount",
+       y = "Quantity Purchased") +
+  scale_y_continuous(labels = label_comma()) +  # This line changes the y-axis labels
+
+    theme_minimal()
+plot(p4)
+
+```
+This distribution looks more smooth now. If there are many wholesalers buying in relatively smaller quantities, they potentially still be in the data.
+
 Assigning a final name to the subset to make it more relevant to the actual data.
@@
 ```{r}
-processed_superstore <- clean_data_step4
-
+processed_superstore <- clean_data_step5
 str(processed_superstore)
+skimr::skim(processed_superstore)
 ```
+Finally, the cleaned data has 289065 observations of products purchased by 4117 customers residing in 33 different countries in 14951 transactions.
+There were a total of 3785 types of products purchased by the customers in this final data.
 
@@
 
-See here for some suggestions on how to store your processed data:
-http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata
-
 ```{r}

### 2024-04-22 products/manuscript/Manuscript.qmd
Project Part 5

--- 2024-03-27 products/manuscript/Manuscript.qmd
@@
 ---
-title: "Customer Segmentation"
+title: "The Value of Loyalty: Customer impact on Online Store Revenue"
 subtitle: ""
@@
 ```{r, echo=FALSE, message=FALSE}
+#| warning: false
+#| message: false
+
 # load a few R packages
@@
 # Summary/Abstract
-_Write a summary of your project._
+This project investigates the revenue implications of customer return behaviors during the holiday season for a UK-based online retailer specializing in unique gifts.
+In the current landscape, where e-commerce has reshaped retail, understanding customer loyalty becomes crucial for revenue growth.
+Utilizing a dataset of transactions spanning from December 2010 to December 2011, this study evaluates the revenue contributions of returning customers versus those who do not return.
 
-{{< pagebreak >}}
+The analysis employs Linear, LASSO, and Generalized Additive Models (GAM) to explore the relationship between the sales revenue of the retailer and customer loyalty defined by returns of the customer during holiday season.
+The models are assessed using the root mean square error (RMSE) metric, with cross-validation reinforcing the analysis.
+LASSO emerged as the most effective model.
+The study also addresses the challenges of forecasting customer purchasing patterns using a dataset with limited information.
+
+For more comprehensive understanding of customer behavior, future studies should consider including product-specific information in the analysis and explore the feasibility of accessing demographic information about the customers.
 
 # Introduction 
+This project examines the revenue dynamics associated with customer returns during the crucial holiday season for a UK-based non-store online retailer specializing in unique all-occasion gifts.
+Given the transformation of the retail sector driven by online shopping[@khouloud2020we], which saw global e-commerce sales surpass 5.78 trillion U.S.
+dollars in 2023 and anticipated to hit 6.3 trillion in 2024[@statista_online_shopping], understanding customer behaviors becomes imperative for retailers[@chow1997toward;@heskett1994putting].
+The intersection of advancing IT, increased internet accessibility, and evolving consumer expectations necessitates strategies to foster customer loyalty[@reichheld2000loyalty] as a means to mitigate turnover costs and bolster revenue.
 
-## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis
+However, the presumed correlation between loyalty and profitability is not always straightforward[@Reinartz2002The; @thomas2004getting].While loyal customers are generally seen as valuable due to their repeat business, not every instance of loyalty may translate to noticeable revenue when considering factors like customer service costs and the nature of purchases.Identifying which segment of customers are truly beneficial is crucial for efficient distribution of resources [@Reinartz2002The].
+This project scrutinizes the revenue contributions from customers who returned for holiday shopping compared to those who did not, with the anticipation of reinforcing the notion that loyal customers are inherently good source of revenue generation.
+The findings aim to offer actionable insights for the retailer to enhance revenue growth strategically.
 
-This project aims to leverage statistical modeling techniques to predict customer segmentation based on their purchasing behavior pattern, using both supervised and unsupervised learning approaches.
-In the initial phase, the analysis employs logistic regression to segregate customers based on the general patterns of purchasing behavior.
-As the study advances, it integrates Machine Learning Tools into the analysis.
+## Background
+Building upon the impact of online shopping on consumer behavior and the importance of customer loyalty, we now turn to the subject of study.
+The unit is a non-store online retail business based on the United Kingdom.
+It specializes in unique all-occasion gifts and deals with more than three thousand types of merchandises.
+Its clientele base expands from retail customers to wholesales customers across United Kingdom and other geographical locations.
 
-It is crucial that organizations understand their customer’s purchasing patterns to devise and implement effective, personalized engagement strategies.
-By analyzing behaviors such as spending habits and purchase frequency, businesses can segment their customers and tailor marketing and sales efforts to cater to the unique needs of each segment.
-This personalized approach can not only boost customer satisfaction but also improve multiple end results like customer loyalty, operational efficiency, and profitability.
+We have one year of sales data of the unit spanning from December of 2010 to December of 2011.
+The data reveals a pattern of sales performance with initial months showing steady yet modest results and gaining noticeable momentum as the holiday season approaches.
+This uptick suggests that the latter part of the year, with its festive attractions, plays a pivotal role in the overall sales activity.
 
-Given the absence of pre-categorized customer groups in the original data, a grouping variable is engineered based on purchase amount and frequency to facilitate the statistical analysis.
+Purchases made outside of major festive events suggests a customer base attracted by the inherent appeal and variety of the product offerings, as opposed to seasonal promotions.
+In contrast, the peak season not only intensifies competition bringing plethora of special offers but also serves as a test for customer loyalty.
+It is during this time that returning customers, who are seemingly less swayed by market-wide offers, affirm their status as the cornerstone of the store's revenue, demonstrating loyalty that extends beyond the allure of holiday discounts.
+These insights into customer buying patterns are instrumental as we explore the operational subtleties of sales trends and their implications for sustained revenue growth.
 
-## Description of data and data source
-_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
+## Hypothesis
 
-This study utilizes the 'superstoredata' dataset which is a transactional dataset of a UK-based, non-store online retail operation spanning from December 1, 2010, to December 9, 2011.
-The store specializes in unique all-occasion gifts to a clientele which includes UK and non-UK based retail and wholesale customers.
-This study focuses on international sales of the store.
-The dataset features nine variables, detailing transaction numbers, product codes, product description, quantities sold, transaction dates, unit prices, customer IDs, country of residence of the customer, and sales amounts.
-Initially this data was downloaded from Kaggle for a preliminary assessment of its usability for this project.
-However, when revisiting the site for further details, it was no longer available on Kaggle.com.
-A subsequent Google search led to the discovery of the dataset in the [UC Irvine Machine Learning Repository](https://archive.ics.uci.edu/ml/datasets/Online+Retail).
-
-## Questions/Hypotheses to be addressed
-_State the research questions you plan to answer with this analysis._
-
-What statistical modeling techniques are most effective in replicating and predicting customer segmentation,based on their purchase behavior pattern, as indicated by purchase amount, purchase frequency and purchase recency?
+This project is centered around understanding the implication of customer loyalty on revenue generation at a UK-based online retailer specializing in unique gift items.
+Focusing on a one-year observational period, this study aims to investigate the revenue associated with customer return during the critical holiday season.
+Our guiding hypothesis is that customers who returned to purchase during this peak shopping period had a disproportionate and positive contribution on the retailer's revenue streams in comparison to customers who did not make return visits within the same season.
 
@@
 
-_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
+## Data and Source
+The data comprises nine variables capturing essential aspects of transactions such as transaction numbers, product codes, product descriptions, quantities sold, transaction dates, unit prices, customer IDs, the country of residence of the customer, and sales amounts.
+It encompasses sales data from December 1, 2010, to December 9, 2011.
+The data does not segregate its sales to wholesalers and retailers.
 
-## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
+The dataset was initially acquired from Kaggle for a preliminary assessment of its suitability for this project.
+However, upon a subsequent visit to the site for additional information, it was discovered that the dataset had been removed from Kaggle.com.
+A follow-up search on Google revealed the dataset's availability at the [UC Irvine Machine Learning Repository](https://archive.ics.uci.edu/ml/datasets/Online+Retail).
+Notably, the version found in the UC Irvine repository lacks the sales amount column, which is calculable by multiplying the quantities sold by the unit prices.
+The analysis for this project was conducted using the dataset originally downloaded from Kaggle.
 
-The primary dataset for this project is the 'superstoredata' dataset, detailing transactions of an online retailer based in United Kingdom.
-A key feature of this data is that it counts sales of a particular item from an Invoice as an observation leading to different items sold in the same Invoice to appear as separate observations.
-Most of the Invoice numbers appear several times in different data observations.
+## Preliminary Data Cleaning
 
-### Initial Data Cleaning
+A key feature of this data is its treatment of each sale of an item on an Invoice as a separate observation, leading to multiple items sold on the same invoice to appear as distinct observations.
+Initially comprising 541,909 observations, the dataset was subjected to comprehensive cleaning.
+This process involved removal of records missing crucial customer ID information, conversion of key variables (CustomerID, Country, Description, StockCode) from character to factor types for better manipulation, and the exclusion of observations with negative quantities and sales, indicative of order cancellations.Further scrutiny led to the identification and removal of transactions with unusually high unit prices, likely representing logistical and transportation costs rather than direct transactional values.
 
-As the project focuses on international customers, initial data cleaning involved eliminating UK transactions downsizing the data size to 46431 observations across nine variables.
+Moreover, the dataset's inability to distinguish sales to wholesalers presented a potential analysis bias due to the inclusion of bulk transactions.
+To mitigate this, scatter plots were employed to identify and eliminate observations corresponding to abnormally large sales amounts or quantities attributed to a single customer.
 
-The preliminary inspection identified several issues needing attention such as missing values in 'CustomerID', negative 'signs in 'Quantity' and 'Sales' values, and incorrectly formatted dates.
-Variables, 'InvoiceNo', 'StockCode', 'Description' and 'Country', originally coded as characters, were converted to factor types to facilitate better data manipulation.
+Following these cleaning efforts, the dataset was refined to 289,065 observations.For a comprehensive overview of the cleaning process, please refer to the detailed explanation in processing section (R>processing-code.processing_superstore.qmd).
 
-1480 observations lacked 'CustomerID' and were removed, as customer purchase pattern cannot be tracked without the unique identifier and 'CustomerID' was converted to factor type from numeric type.
-'Country', 'Description' and 'StockCode' were converted to factors from character variables and transaction date to the date format.
+## Feature Engineering
 
-There were 1372 observations with negative values in quantity and sales due to order cancellation which were excluded from the data.
+This analysis is comprised of four variables.
+The variable of interest in this analysis is the cumulative spending of each customer over the transaction period.
+To assess customer loyalty, which we have identified as our main predictor, we examined whether customers made repeat purchases during the critical holiday season.
+Additional predictors include the total number of transactions per customer throughout the period and whether the customer resides in the United Kingdom.
 
-An examination of 'UnitPrice' distribution revealed that all the high value goods with a Unitprice above 25 were distributed between 75 and 100 percentile.
-A closer inspection at this inter-quartile range disclosed that there were 19 product description categories with the Unitprice above 25.
-Some of those categories had descriptions such as CARRIAGE, POSTAGE, Manual, which appeared to be related to logistics.
-1216 observations with such descriptions were dropped from the data since those were less likely to be directly related to customer purchase.
+We define 'customer loyalty' as a dummy variable based on transactional activity.
+Customers were considered loyal or returning if they had made at least one purchase prior to June 2011 and have returned for at least one additional purchase between October and December 2011.
+This classification was made based on the invoice date.
 
-### Refining the Dataset
-The store also sold merchandises in wholesale quantities.
-A wholesaler potentially had larger purchases and in higher frequencies.
-In absence of clear information on wholesalers, a scatter plot of Cumulative Invoice Value of customers by Purchase Frequency was plotted to identify potential wholesalers.
-The scatter plot revealed that most of the customers made purchases less than a frequency of 50 and that most cumulative purchase values were less than 50,000.
-Assuming that the outliers represented purchases by wholesalers, observations with purchase frequencies above 50 and cumulative Invoice values above 50,000 were dropped from the data.
-This step excluded 9745 observations from the data.
-The cleaned data at this point had 32602 observations with 9 variables.
-Next, the quantities purchased in each invoice were plotted against the individual invoice values.
-The scatter plot revealed that the observations were concentrated within the purchase quantity below 1000 items and invoice amounts of 2,500.
-Assuming that the sparsely distributed points represent purchases by wholesalers, observations with Quantity above 1000 and individual Invoice value above 2,500 were further removed from the data.This step excluded 4485 observations for a final total observation of 28117.
+To calculate the number of purchases, we aggregated the data by invoice number and subsequently tallied the invoices for each Customer ID.
+Similarly, the total sum of purchases is derived by aggregating sales figures under each Customer ID.
 
-The study of customer spending pattern requires the knowledge of each individual transaction amount.
-The subsequent cleaning process reorganized the dataset by 'InvoiceNo', consolidating sales data under each invoice.
-This final data is saved as processed_superstore_RFM in the data > processed-data.
+The final groundwork is consolidated on a per Customer ID basis, ensuring each customer was represented as a unique observation.
+This step enables a robust statistical analysis of spending patterns, customer loyalty, and geographic influence on purchasing behavior setting the stage for insightful findings on the dynamics of customer engagement and revenue generation.
 
-### RFM Classification
-In absence of a pre-existing customer groups, the study segmented customers into two distinct groups using a recency, frequency and monetary (RFM) model.
-RFM model is a commonly used behavior-based model in analysing customer behavior [@yeh2009knowledge].
-Recency is measured in days or months and represent the interval between the most recent transaction time and the time of analysis, and a lower number interval is preferred.
-Frequency is the number of purchases made in a certain period, and monetary is the total amount the customer spent during that time period.[@wei2010review]
+## Model Development
 
-In the RFM classification process, the Recency, Frequency and Monetary values are calculated for each customer, ensuring that each customer was represented only once in the modified dataset.Frequency in this data is the total no.
-of Invoices issued to the customer, and Monetary is the total purchases made by the customer during the stipulated period.
-Recency was calculated from the next day of the transaction period.
-Each customer was assigned a value ranging from 1 to 4 for the Recency, Frequency and Monetary attributes, reflecting their quartile position within each metric.
-Subsequently, these metrics were combined to formulate a three-digit score for every custoemr, positioning the Monetary value in the hundreds place, Frequency in the tens place, and Recency in the units place.
-All the customers with a score above 344 were assigned as High-Value Customers.
+For the predictive analysis, three distinct models were fitted and evaluated: linear models, regularization-based models, and a non-linear model.
+The dataset was partitioned, allocating 75% for model training purposes and the remaining 25% designated for testing the ultimately selected model.
+To ensure model robustness, cross-validation was employed across all models, adhering to a ten-fold resampling methodology.
+The analysis was conducted within the tidymodels framework, with the exception of the cross-validation process for the Generalized Additive Model (GAM), which required handling separately due to its unique characteristics.
 
-Following RFM segmentation, logistic regression is applied to predict customer group membership.
-This method utilized the segments derived from RFM as the dependent variable, with the Recency, Frequency and Monetary attributes serving as independent variables.
-The objective is to develop a model for predictive analysis based on those segments.
+## Model Overview
 
-## Statistical analysis
-_Explain anything related to your statistical analyses._
+### Linear Model
+A linear model was deemed suitable given the limited number of predictor variables and the anticipated linear relationship between these predictors and the outcome variables.
+This model serves as a benchmark, providing a baseline from which the performance of more complex models can be evaluated.
 
-{{< pagebreak >}}
+### Least Absolute Shrinkage and Selection Operator (LASSO) Regression
+In search of a model that balances complexity with interpretability, LASSO regression was selected.
+LASSO has useful feature of selectiing and effectively reducing the coefficients of less significant predictors to zero.
+This approach simplifies the model and also aimed at enhancing predictive accuracy.
+
+### Generalized Additive Model (GAM)
+To explore and accommodate potential non-linear relationships between the predictors and outcome variables, the Generalized Additive Model (GAM) was incorporated into the analysis.
+Chosen for its flexibility, GAM allows for the modeling of complex, non-linear interactions in a way that is both interpretable and visually demonstrable, offering a deeper insight into the underlying data structure.
+
+## Model Evaluation Metric
+The efficacy of these models was quantitatively assessed using the root mean square error (RMSE).
+RMSE is the square root of the average squared differences between the predicted and observed values.
+It offers a comprehensive measure of prediction accuracy, serving as the key metric for model comparison and selection.
 
@@
 
-## Exploratory/Descriptive analysis
-
+## Descriptive analysis
+The proportion of customers who purchased in both the periods is 30.72 percent.
 @tbl-resulttable1 depicts the descriptive statistics of various aspects of customers purchase behavior.
-There were 1402 transactions with 399 customers outside UK during the period from December 1, 2010 to	December 9, 2011.
-The store recorded a mean sales of GBP 436 per transaction, an average cumulative sales of GBP 1532 to individual customers and sales frequency of above 3 times per customer.
+The store made 14951 transactions with 4117 customers during the period from December 1, 2010 to	December 9, 2011, by recording a mean sales of GBP 290.19 (SD=GBP 187.36) per transaction, an average cumulative sales of GBP 1053.84 (SD = GBP 1429.07) to individual customers and an average sales frequency of above 3 times (SD = 4.28) per customer.
+The average no.
+of days before a customer returns to shop is recorded at 97.90 (SD=102.36) days.
 
@@
 ```
-
-@fig-result1 depicts the distribution of customer segmentation and suggests that approximately 25 percentage of the customers are grouped as Valued Customers.
-
+@fig-result1 presents distribution of sum of customer purchase, invoice value, frequency and recency.
+For fitting linear model, the sum of purchase and frequency were log transformed due to heavy right-skewness.
+The metrics were converted into original scale after fitting the model.
 ```{r}
 #| label: fig-result1
-#| fig-cap: "Distribution of Sales and Purchase frequency"
+#| fig-cap: "Distribution"
 #| echo: FALSE
-knitr::include_graphics(here("results", "figures", "Customer_Segment.png"))
+knitr::include_graphics(here("results", "figures", "Sales_distribution.png"))
 ```
-
-@fig-result2 depicts two density plots for the Cumulative Sales and Invoice Values across customers and a histogram for Customer purchase frequencies.
-Both density plots are heavily right-skewed.The averages of total sales to a customer and total invoice values appear to be less than 2500 and 500 sterling pounds, respectively, supporting the mean values of 1532 and 436 in the descriptive table.
-The average frequency appears to be less than 5 times, which is also observed in the descriptive statistics table.
+@fig-result2 depicts the distribution of monthly sales of the store.
+The sales are noted to be modest with occassional and nominal variations during the initial period.
+Starting from October, the sales trajectory showed an upward trend, reaching maximum in November, which could be due to early holiday shopping for the approaching Christmas season.
+It is reasonable to anticipate that past customers would engage in purchasing activities during this peak sales period.
 
@@
 #| label: fig-result2
-#| fig-cap: "Distribution of Sales and Purchase frequency"
+#| fig-cap: "Distribution of Monthly Sales"
 #| echo: FALSE
-knitr::include_graphics(here("results", "figures", "Sales_distribution.png"))
+knitr::include_graphics(here("results", "figures", "Monthly_Sales.png"))
 ```
-@fig-result3 depicts a density plot of number of days since the last transaction made by a customer.
-The distribution is right-skewed.
-This plot indicates that most customers made a purchase transaction within the last 30 days of the reference date.It also reveals that few customers had not make a transaction for a long time.
+
+@fig-result3 and @fig-result4 depict two plots, Recency by Frequency plot and Recency by Monetary plot.
+While the Recency by Frequency plot revealed an expected pattern where customers with more frequent purchases tended to have more recent transactions, an unexpected trend was also noted.
+There are some evidence that a segment of customers with a history of multiple purchases did not return during the peak sales period.
+The Recency by Monetary plot mirrored the trend observed in the 'Recency by Frequency' plot.
+The plot revealed that a subset of customers who have purchased over 2,500 in monetary value in the past, did not make subsequent purchases within the following 100 days.
 
@@
 #| label: fig-result3
-#| fig-cap: "Distribution of Recency of Purchase"
+#| fig-cap: "Recency Plots"
 #| echo: FALSE
-knitr::include_graphics(here("results", "figures", "Recency.png"))
+knitr::include_graphics(here("results", "figures", "Recency_by_Frequency.png"))
+knitr::include_graphics(here("results", "figures", "Recency_by_Monetary.png"))
 ```
+@fig-result5 provides a visual impression of the the cumulative purchase by customers who returned during the holiday season (loyal customers) versus those who did not return during the holiday season.
+It suggests that the returning customers contributed more to the store's total sales than the non-returning customers.
 
-@fig-result4 depicts a trend of monthly international sales.
-A spike is noticed in November, presumably due to Christmas in the following month, which is also supported by distribution of the Recency plot.
-```{r}
-#| label: fig-result4
-#| fig-cap: "Monthly Sales trend"
-#| echo: FALSE
-knitr::include_graphics(here("results", "figures", "Monthly_Sales.png"))
-```
-@fig-result5 depicts country-wise trend of international sales by sales amount. Top customers were in Germany, France, Spain and Belgium.  
 ```{r}
 #| label: fig-result5
+#| fig-cap: "Distribution of Cummulative Purchase by Returning status of the Customer "
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "Purchase by loyaly.png"))
+```
+@fig-result6 presents country-wise sales trend. Majority of the customers are noted to be from the United Kingdom, followed by Germany, France, Belgium and Spain.  
+```{r}
+#| label: fig-result6
 #| fig-cap: "Countrywise Sales trend"
@@
 
-## Basic statistical analysis
+## Basic Statistical Analysis
 
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
+A simple linear model was fitted to examine the association between total purchase a customer makde and the returning status of the customer.
+@tbl-resulttable2 depicts the results of the regression.
+It indicates that there is a positive association between customers returning during the peak season and the cummulative purchase of the customer.
 
-Penalized logistic regression of customer segment was performed on Recency, Frequency and Monetary variables using 'tidymodels' package and 4-fold corss-validation techniques.
+```{r out.width="100%", fig.show='hold', warning=FALSE}
+#| label: tbl-resulttable2
+#| tbl-cap: "Model Metrics"
+#| echo: FALSE
+resulttable2 = readRDS(here("results", "tables", "slr_fit_df.rds"))
+knitr::kable(resulttable2)
 
-@fig-result6 depicts the area under ROC curve by the range of penalty values.
-It provides visualization of validation set metrics.
-The curve has extreme high values, all above 0.9995.
-A sharp drop is noticed before the penalty value of 0.0010, suggesting the sensitivity of the performance of the model at this penalty value.
-This is followed by a sharp rise after the penalty value of 0.01.
-These features of model performance are indicative of the training data and the data being easily separable.
+```
 
-```{r}
-#| label: fig-result6
-#| fig-cap: "Visualization of validation set metrics"
+## Full Analysis
+
+Three models were fitted to predict the sum of customer purchases: Linear Model, LASSO and GAM.
+@tbl-resulttable3 shows the model comparison for estimating the sum of purchases.
+The GAM model has the lowest RMSE, suggesting it performs well in prediction accuracy.
+However, when subjected to cross-validation, the GAM model's RMSE revealed a decrease in prediction accuracy across various subsets of the data compared to its initial evaluation based solely on the training data.
+Both the tuned and untuned Lasso models demonstrated the next lowest RMSE values.
+We  chose the simple Lasso model as our final model for its simplicity.
+
+```{r out.width="100%", fig.show='hold', warning=FALSE}
+#| label: tbl-resulttable3
+#| tbl-cap: "Model comparison"
 #| echo: FALSE
-knitr::include_graphics(here("results", "figures", "ROC(AUC)_Penalty.png"))
+resulttable3 = readRDS(here("results", "tables", "RMSE_comparison.rds"))
+knitr::kable(resulttable3)
+
 ```
-@fig-result7 depicts validation set ROC curve plotting true positive rate (sensitivity) against the false positive rate (1-specificity) at various threshold settings.
-The curve moves along 1-specificity and rises only from the end.This suggests that the model has no  ability to distinguish between the two customer categories.
-This suggests that the logistic regression model is not appropriate to the dataset.
-It pointed out the need to revisit the choice of model.
+The performance of the selected model, the simple Lasso model was evaluated by fitting it on the test set. @tbl-resulttable4 presents the RMSE of the model on the test data.
+```{r out.width="100%", fig.show='hold', warning=FALSE}
+#| label: tbl-resulttable4
+#| tbl-cap: "Test Data: Model Metrics"
+#| echo: FALSE
+resulttable4 = readRDS(here("results", "tables", "RMSE_Test.rds"))
+knitr::kable(resulttable4)
+
+```
+@fig-result7 depicts the Predicted versus Observed plot for the training and test data.
 
@@
 #| label: fig-result7
-#| fig-cap: "Visualization of Validation set ROC curve"
+#| fig-cap: "Predicted vs. Observed Plot"
 #| echo: FALSE
-knitr::include_graphics(here("results", "figures", "ROC(AUC)_Validation.png"))
+knitr::include_graphics(here("results", "figures", "Predicted_Observed.png"))
 ```
 
-## Full analysis
-
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
-
-#Example @tbl-resulttable2 shows a summary of a linear model fit.
+@fig-result8 presents Residual plot of the test data.
 
 ```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
+#| label: fig-result8
+#| fig-cap: "Residual Plot"
 #| echo: FALSE
-#resulttable2 = readRDS(here("results","tables","summarytable.rds"))
-#knitr::kable(resulttable2)
+knitr::include_graphics(here("results", "figures", "Residual.png"))
 ```
-
-{{< pagebreak >}}
 
@@
 ## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
+Three predictive models were applied to compare the total purchase amount by customers who returned during the peak season versus those who did not return: a Linear Model, LASSO, and GAM.
+Among these, the GAM model exhibited the lowest RMSE, signaling superior prediction accuracy.
+However, its performance dipped under cross-validation, displaying a decrease in accuracy across various data subsets, which diverged from its initial evaluation based on training data alone.
+
+Both tuned and untuned LASSO models recorded the next lowest RMSE values.
+This similarity in performance resulted because the tuning process yielded a penalty value for the optimal model close to that of the untuned version, with all predictors being retained.
+Owing to its simplicity, the simple Lasso model was chosen as the final model.
+When evaluated on the test set, this model showed a slightly higher RMSE of 683.87, compared to 675.38 on the training set, indicating a consistent yet marginally less accurate prediction capability.
+
+The Predicted versus Observed plots for both training and test data revealed that, for lower-values, the model's predictions were close to the diagonal line, suggesting relatively accurate predictions.
+However, as the values increased, both sets of data deviated from the perfect fit line, suggesting diminished accuracy for higher values.
+This pattern is suggestive of the potential influence of additional factors not captured by the model.
+
+Similarly, the residual plot for the test data revealed a fan-shaped pattern, indicative of heteroscedasticity and variance in prediction errors which increased with higher value predictions.
+This further points to the existence of other influential factors not considered in the model.
+
+Despite its merits, the Lasso model did not fully capture the complexities of the dataset.
+The diagnostic plots for both training and test data underscore the need for further model refinement to predict the difference in sum of purchases by returning and not-returning customers during the peak season with more accuracy.
 
 ## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
+
+This project explored a complex relationship between the outcome and predictors by examining Linear, LASSO, and GAM models using limited available information.
+It also ensured the model's reliability through cross-validation.
+
+Customer purchasing behavior is inherently multifaceted and deeply influenced by various demographic factors, including but not limited to gender, age, race, and educational background.
+These aspects can remarkably shape buying habits, preferences, and overall engagement with retail services.
+In our analysis, the models deployed were constrained by the scope of available data, primarily focusing on purchase frequency, recency, and the total amount spent.
+This limited information restricts our ability to fully understand and accurately predict customer behavior fully.
+
+Considering the retail store's extensive inventory of over 3,000 items, along with associated quantities and unit prices, incorporating these information may potentially enhance the model's accuracy.
 
 ## Conclusions
-_What are the main take-home messages?_
-
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.
-
-{{< pagebreak >}}
+Future research could explore how the product-specific data contribute to predicting outcomes more effectively.
 

### 2024-04-25 reviews/Roth-project-review.qmd
Part 5 - Roth Review
Upload of Roth's review document and creation of new folder called "review"

---
title: Project Review Template 
author: Cassia Roth
date: date-modified
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: The Value of Loyalty: Customer Impact on Online Store Revenue

Name of project author(s): Malika Dhakhwa

Name of project reviewer: Cassia Roth

# Specific project content evaluation

## Background, Context and Motivation

How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

You clearly make the case for analyzing repeat customers to online retailers during holiday seasons given the importance of online shopping to the global consumer economy. I would clarify what you mean by repeat customers. Do you mean during the same holiday season or across multiple holiday seasons? You answer this question in the Background 2.1 section, but I would clarify it in the Introductory paragraph as well.

### Summary assessment

-   **strong contextualization and motivation**

## Question description

How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

The question, from my understanding, is: Do customers who make repeat purchases at one online retailer during a holiday season have a higher positive contribution to the retailer's annual revenue than customers who did not make repeat visits during that same season? Section 2.2 is great, but it is missing or doesn't clearly explain some of the important information you have in your EDA file: that customers who purchase during non-holiday periods are more likely to return during the holiday season and that their purchases during the holiday season only have a higher positive contribution than customers who only purchase during the holiday season. This is a key point that should be included in your question in a clear manner.

### Summary assessment

-   **question/hypotheses somewhat explained**

## Data description

How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

You do a superb job of describing your overall data. The codebook is clear as is the data source.

### Summary assessment

-   **source and overall structure of data well explained**

## Data wrangling and exploratory analysis

How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

The EDA is really well thought out and thorough. I appreciated all the visualizations and clear descriptions, both as text and as in-line comments to your r-code. My main question is why you dropped negative values. You say these are only order cancellations, but aren't these item returns (or also item returns)? This might be correlated with whether or not customers return to purchase during the holiday season (if these item returns occurred prior). In other words, I think there is a relationship here you could further explore that might be related to your question and hypothesis. Your description of feature engineering was very clear and made logical sense. Here, how you describe loyal customers is very easy to understand. You might want to bring a version of this into the introduction and hypothesis sections prior.

### Summary assessment

-   **essentially no weaknesses in wrangling and exploratory component**

## Appropriateness of Analysis

Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

Your descriptions of your models is very clear and informative. From my understanding of these models, they make sense. However, you never clearly define your exposure/outcome variables in this section, which would be helpful for your reader.

### Summary assessment

-   **strong and reasonable analysis**

## Presentation

How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

You might want to use the flextables package to make your tables more visualizing appealing in word. The label for Figures 3 and 4 did not come out correctly on my word document when I rendered the Quarto file. You might also want to think about only including half of the images you have in your final paper, and creating a supplemental file with other images.

When describing your linear model (table 2), you might want to clarify that the dependent variable is cumulative purchase price in UK pound sterling (it is not clear from the table itself because your description just says cumulative purchase.

### Summary assessment

-   **results are very well presented**

## Discussion/Conclusions

Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

You do an excellent describing why you end up going with the LASSO model, but you don't spend very much time explaining what the model actually means. What are the findings from the LASSO model? Do returning customers contribute positively to the retailer's revenue per your hypothesis? This is missing from the discussion, which focuses only on model performance. If your paper is on the prediction of these models in relation to this type of question, then you will need to change that in your introductory section.

### Summary assessment

-   **minor parts wrong, missing or unclear**

## Further comments

Malika, you've done an excellent job here. I would suggest changing some of the language in your paper, since I found it somewhat confusing. In the abstract you state the "revenue implications of customer return behaviors." After reading more of the paper and running all of your code, I realize you mean customers who return to the website to purchase more items. However, "return behavior" could also imply customers returning items they don't like. Perhaps using the term "repeate customers" would be helpful.

# Overall project content evaluation

Evaluate overall features of the project by filling in the sections below.

## Structure

Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

Everything was well structured and cleaned up. Great job.

### Summary assessment

-   **well structured**

## Documentation

How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

Everything was very well documented and easy to follow.

### Summary assessment

-   **fully and well documented**

## Reproducibility

Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

The master readme was very clear and easy to follow for reproducibility. I was able to reproduce the entire project without any problem.

### Summary assessment

-   **fully reproducible without issues**

## Thoroughness

How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

I think you could better address your question/hypothesis in the discussion and conclusion sections. You could also beef up your conclusions to think about what this means for online retailers and how they might use this information to increase their revenue.

### Summary assessment

-   **strong level of thoroughness**

## Further comments

Great job Malika! I would double check your manuscript.qmd because often you have small grammatical errors, including spelling errors and particularly missing spaces between references or sentences.
